### 2024-03-03 article.Rmd
initial commit

---
title: "sfislands: An R Package for Accommodating Islands and Disjoint Zones in Areal Spatial Modelling"
date: "2022-11-09"
abstract: >
  Fitting areal spatial models can be a cumbersome task, particularly when the geographical units are not well-behaved. The presence of islands gives rise to particular issues when creating neighbourhood structures based on contiguity. We might also want to edit the output of a condition-based neighbourhood structure. It might be appropriate to add or remove connections to take account of infrastructure links or other domain-specific issues which are not apparent from a basic set of polygons. This package streamlines these processes with tools to account for the presence of islands and to intuitively make manual alterations to neighbourhood structures with the help of convenient mapping functions. Beyond the setting up of a satisfactory neighbourhood structure, it has further helper functions specific to the `mgcv` package which reduce the human workload required to extract estimates from models with spatial and hierarchical structures.
draft: true
author:  
  - name: Kevin Horan
    affiliation: 
    - Hamilton Institute, Maynooth University
    address:
    - Maynooth
    - Co. Kildare, Ireland
    url: https://github.com/horankev
    orcid: 0009-0003-9378-0084
    email:  kevin.horan.2021@mumail.ie
  - name: Katarina Domijan
    affiliation:
    - Department of Mathematics and Statistics, Maynooth University
    address:
    - Maynooth
    - Co. Kildare, Ireland
    email: katarina.domijan@mu.ie
    orcid: 0000-0002-4268-2236
  - name: Chris Brunsdon
    affiliation:
    - National Centre for Geocomputation, Maynooth University
    address:
    - Maynooth
    - Co. Kildare, Ireland
    email: Christopher.Brunsdon@mu.ie
    orcid: 0000-0003-4254-1780
type: package
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")

# install.packages("devtools")
devtools::install_github("horankev/sfislands")

# required libraries
library(tidyverse)
library(kableExtra)
library(sf)
library(sfdep)
library(sfislands)
library(mgcv)
library(here)
library(ggsflabel)
library(ggspatial)
library(ggpubr)
library(AER) # to test for overdispersion
library(ggmagnify)

# To scrape data
library(rvest)
library(httr)
library(polite)
library(janitor)

# for geom_sf_label_repel consistency
set.seed(123)

# set consistent theme
theme_set(
  theme_bw() +
    theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", linewidth=1.5), 
          axis.title = element_blank(),
          axis.ticks = element_blank())
)

```

```{r}
# load data

# datasets are all available at https://github.com/horankev/quake_data 
# or https://github.com/horankev/london_liverpool_data
# alongside details on precisely how they were sourced etc

# indonesia
quakes_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/quakes_df.rds")))
provinces_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/provinces_df.rds"))) |> 
  rename(damaging_quakes_total = quake_mlxl_total,
         damaging_quakes_density = quake_mlxl_density)
counties_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/counties_df.rds")))
faults_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/faults_df.rds")))
nearby_countries_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/nearby_countries_df.rds")))
indonesia_outline <- provinces_df |> summarise()

# london
london <- readRDS(gzcon(url("https://github.com/horankev/london_liverpool_data/raw/main/data/london.rds")))
thames <- readRDS(gzcon(url("https://github.com/horankev/london_liverpool_data/raw/main/data/thames.rds")))
crossings_roadped <- readRDS(gzcon(url("https://github.com/horankev/london_liverpool_data/raw/main/data/crossings_roadped.rds")))

# liverpool
liverpool <- readRDS(gzcon(url("https://github.com/horankev/london_liverpool_data/raw/main/data/liverpool.rds")))
msoa_sf <- liverpool |> 
  group_by(msoa_cd) |> 
  summarise()
lsoa_sf <- liverpool |> 
  group_by(lsoa_cd) |> 
  summarise()
mersey <- readRDS(gzcon(url("https://github.com/horankev/london_liverpool_data/raw/main/data/mersey.rds")))

```

# Introduction

A key feature which differentiates spatial statistics is the
non-independence of observations and the expectation that neighbouring
units will be more similar than non-neighbouring ones [@tobler]. If this
is not accounted for, the assumptions of many types of models will be
violated. The relationships between all spatial units in a study can be
represented numerically in a spatial weights matrix. In order to build
this, we must first decide on what constitutes being a neighbour. We can
see this as a continuous relationship where degree of neighbourliness is
a function of connectivity, which could be represented as some measure
of distance. Alternatively it could be a binary situation where each
pair of units either are (1) or are not (0) neighbours. This can be
based on a condition such as contiguity of some sort, or a distance
constraint. It is the job of the modeller to formulate a hypothesis
which justifies their choice of neighbourhood structure.

For R users, the \CRANpkg{spdep} package [@spdep] has long been popular for the
creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
structures and their compatibility with the \CRANpkg{tidyverse}  [@tidyverse], the \CRANpkg{sfdep}  [@sf] package [@sfdep] has presented the same
functionality in addition to extra features in a tidy structure based on
lists in a similar way to \CRANpkg{sf}.

The most appropriate form of neighbourhood structure will depend on the specific context. @BrizRedn2021 compared different structures in the context of COVID-19 data. They note that @Earnest2007 found that distance-based matrices were more appropriate when examining birth defects in Australia, whereas @Duncan2017 found that a first-order contiguity structure produced a better fit than others in the context of lip cancer incidence in Scotland.

The most commonly used neighbourhood structure is one based on first-order queen contiguity, where units are considered neighbours if they
share at least a vortex of boundary. However, as the name suggests, this
will lead to problems when non-contiguous units such as islands or
exclaves are present. Less obviously, depending on how the geographic
units are described, areas on either sides of rivers may be
inappropriately classified as neighbours or not neighbours. Furthermore,
the presence of infrastructure such as tunnels, bridges or ferry
services might be satisfactory to meet our hypothesis of the required
degree of connectivity to be considered neighbours.

The aim of `sfislands` is to deal with the situations described above in a
convenient and open manner. It allows us to set up a structure, quickly
map it, and then examine it. The structure can then be edited and the process re-iterated until we have described a
spatial relationship structure with which we are satisfied.

The above can be considered as the *pre-functions* of the package. A
further set of feature of `sfislands`, which we refer to as *post-functions*,
are for use with the \CRANpkg{mgcv} package [@mgcv]. It is straightforward to use
`sfislands` neighbourhood structures in models with \CRANpkg{mgcv} but it can be quite awkward
to extract the estimates. These *post-functions* conveniently provide the estimates in tidy format of random effects and intrinsic conditional autoregressive (ICAR) components from such models and give the ability to quickly map them for visualisation purposes.

## Typical use-cases

In this paper, we will look at three examples to show different use-cases for
`sfislands`. The first example focuses on earthquakes in
Indonesia. It shows a scenario where all of the functions are used, from
setting up contiguities to modelling and examining the estimates of the
model.

The second example looks at London and how, despite the lack of islands,
the presence of a river means that some of the pre-functions of
`sfislands` can be useful.

The final example focuses on Liverpool. There are no islands or issues
of discontiguity in this dataset. Instead, we will show how the package
can be used to fit a multilevel model with an ICAR component at the lowest
level. We show how the results can be presented in tidy form and quickly
visualised. We also show how different types of contiguity from the
\CRANpkg{sfdep} package can used within the \CRANpkg{mgcv} modelling
framework and how the estimates derived from them can be conveniently
compared.

# Why use `sfislands`?

It greatly speeds up the workflow of fitting areal spatial models.

* **pre-functions** for setting up contiguities:

  + it addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data. `sfislands` will include a contiguity for all units,

  + it gives tools to immediately visualise this structure as a map,

  + these maps are created using \CRANpkg{ggplot2}  [@ggplot2], which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax,

  + as the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence,

  + connections made by the package which are not contiguities can be accessed to ensure openness in the process.

* For modelling:

  + these neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms}. [@brms], \pkg{r-inla}  [@r-inla] and more.

* **post-functions** for `mgcv` models:

  + it simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models),

  + these effects can be quickly visualised in \CRANpkg{ggplot2} maps.

# Pre-functions

The first group of functions, shown in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:prefunc-html)', '\\@ref(tab:prefunc-latex)'))`, deals with the creation
of a contiguity structure in the presence of discontiguities. The
resultant structure can be quickly mapped to check if it is
satisfactory. Connections can be manually added or removed by name or
index number. By an iterative process of changes and examination of a
quick map, a satisfactory structure can be decided upon.

Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:prefunc-html)', '\\@ref(tab:prefunc-latex)'))` shows these functions and their purpose.

```{r}
prefunc <- tribble(
  ~`function`, ~purpose,
  "st_bridges()", "create a neighbourhood contiguity structure, with a k-nearest neighbours condition for islands",
  "st_quickmap_nb()", "check structure visually on map",
  "st_check_islands()", "check the contiguities which have been assigned to islands",
  "st_manual_join_nb()", "make manual changes by adding connections",
  "st_manual_cut_nb()", "make manual changes by removing connections"
)

```

```{r prefunc-html, eval = knitr::is_html_output(), layout = "center"}
knitr::kable(prefunc, format = "html", caption = "Pre-functions: setting up a neighbourhood structure.") |> 
  kable_styling(full_width = TRUE)
```

```{r prefunc-latex, eval = knitr::is_latex_output()}
knitr::kable(prefunc, format = "latex", caption = "Pre-functions: setting up a neighbourhood structure.") |> 
  kableExtra::kable_styling(font_size = 9) |> 
  column_spec(2, width = "7cm")
```

They deal with the creation
of a contiguity structure in the presence of discontiguities. The
resultant structure can be quickly mapped to check if it is
satisfactory. Connections can be manually added or removed by name or
index number. By an iterative process of changes and examination of a
quick map, a satisfactory structure can be decided upon.
Going through each function in more detail:

Use the rectangles shown in Figure \@ref(fig:rects1) as a simple demonstration. Rectangles 1-3
are contiguous while 4 and 5 are islands.

```{r}
rect1 <- st_polygon(list(rbind(c(0, 0), c(0, 2), c(2, 2), c(2, 0), c(0, 0))))
rect2 <- st_polygon(list(rbind(c(2, 0), c(2, 2), c(4, 2), c(4, 0), c(2, 0))))
rect3 <- st_polygon(list(rbind(c(2, 2), c(2, 4), c(4, 4), c(4, 2), c(2, 2))))
rect4 <- st_polygon(list(rbind(c(5, 0), c(5, 1), c(6, 1), c(6, 0), c(5, 0))))
rect5 <- st_polygon(list(rbind(c(0.8, 3), c(0.8, 4), c(1.8, 4), c(1.8, 3), c(0.8, 3))))

# Create sf objects for each rectangle
rect1_sf <- st_sf(geometry = st_sfc(rect1))
rect2_sf <- st_sf(geometry = st_sfc(rect2))
rect3_sf <- st_sf(geometry = st_sfc(rect3))
rect4_sf <- st_sf(geometry = st_sfc(rect4))
rect5_sf <- st_sf(geometry = st_sfc(rect5))

# Combine the rectangles into one sf object
rectangles <- rbind(rect1_sf, rect2_sf, rect3_sf, rect4_sf, rect5_sf) |> 
  mutate(name = c("Rect1","Rect2","Rect3","Rect4","Rect5"))
```

```{r, rects1, fig.width=4, fig.height=2, fig.cap = "Simplified scenario with five rectangles. "}
ggplot(rectangles) + geom_sf(colour="black", linewidth=1) + geom_sf_label(aes(label=name),size=2.5) + 
  coord_sf(datum=NA) + 
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5))

```

## st_bridges()

This function requires as arguments an \CRANpkg{sf} dataframe and the name of a column of unique identifiers, preferably names, of each spatial unit. The k-nearest neighbours for connecting islands can then be specified. The output is a *named* neighbourhood structure in either list or matrix form as desired, which can be either a standalone object or included as an additional column in the original \CRANpkg{sf} dataframe. While it is not necessary in all modelling packages for the neighbourhood list or matrix to be *named*, it is good practice to do so and is mandatory when using, for example, \CRANpkg{mgcv}.

In the following examples, we have chosen to ignore islands with the argument `remove_islands = TRUE` and to return a list and matrix structure respectively by specifying this in the `nb_structure` argument and choosing `add_to_dataframe = FALSE`:

```{r, echo=TRUE}
# output a named list

st_bridges(rectangles, 
           "name", 
           remove_islands = TRUE, 
           nb_structure = "list", 
           add_to_dataframe = FALSE) |>
  head()

```

```{r, echo=TRUE}
# output a named matrix

st_bridges(rectangles, 
           "name", 
           remove_islands = TRUE, 
           nb_structure = "matrix", 
           add_to_dataframe = FALSE) |> 
  head()

```

In these examples, we choose to join islands to their 1 nearest
neighbour, which is the default setting, and to return the output as a column called "nb" in the original
\CRANpkg{sf} dataframe (this is the default setting):

```{r, echo=TRUE}
# output a named list as a column "nb" in original dataframe

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1, 
           nb_structure = "list") |> 
  head()

```

```{r, echo=TRUE}
# output a named matrix as a column "nb" in original dataframe

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1, 
           nb_structure = "matrix") |> 
  head()

```

These structures can serve as the input to models in \CRANpkg{brms},
\pkg{r-inla}, \pkg{rstan}  [@rstan] or \CRANpkg{mgcv}. Rather than having a separate neighbours
column, it is included as a named list or matrix in the original \CRANpkg{sf}
dataframe, in the spirit of the \CRANpkg{sfdep} package.

## st_quickmap_nb()

It is much more intuitive to examine these structures visually than in
matrix or list format. This can be done with the `st_quickmap_nb()`
function as shown in Figure \@ref(fig:rects2).

```{r, rects2, fig.width=2, fig.height=1, echo=TRUE, fig.cap = "Queen contiguity and islands connected to nearest neighbour. "}
# default is 'nodes = "point"'

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_quickmap_nb()

```

If we wish to make edits, it might be more useful to represent the nodes
numerically rather than as points (Figure \@ref(fig:rects3)).

```{r, rects3, fig.width=2, fig.height=1, echo=TRUE, fig.cap = "Queen contiguity and islands connected to nearest neighbour. Nodes are shown as numeric indices. "}
# with 'nodes = "numeric"'

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_quickmap_nb(nodes = "numeric")

```

## st_check_islands()

This function will tell us openly what connections have been made which
are not based on contiguity. It gives both the name and index number of each pair of added connections.

```{r, echo=TRUE}
st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_check_islands()

```

## st_manual_join_nb()

If we feel that 4 should also be connected to 3, this can be done
manually (Figure \@ref(fig:rects4)).

```{r, rects4, fig.width=2, fig.height=1, echo=TRUE, fig.cap = "With an additional connection between 3 and 4. "}
st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_quickmap_nb(nodes = "numeric")
```

## st_manual_cut_nb()

And perhaps there is a wide river between rectangles 1 and 2 which
justifies removing the connection. We will edit it this time using
names (Figure \@ref(fig:rects5)).

```{r, rects5, fig.width=2, fig.height=1, echo=TRUE, fig.cap = "With an additional connection between 1 and 2. "}
st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_manual_cut_nb("Rect1","Rect2") |> 
  st_quickmap_nb(nodes = "numeric")
```

The use of these structures is particularly common in CAR-type models.
These are often implemented in a Bayesian framework using
\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. The pre-functions will output the
neighbourhood structure in the desired format for use in any of these
frameworks. A convenient frequentist alternative is to use the
\CRANpkg{mgcv} package. It has the functionality to create such models
using `bs="mrf"`. It also has the ability to combine these with a
hierarchical structure using `bs="re"`. While the outputs from the
Bayesian structures mentioned above can be extracted in the same way as
any other component of the model, it can be a bit awkward to get the
estimates from \CRANpkg{mgcv} model. `sfislands` has two post-functions to conveniently extract and visualise these.

# Post-functions

Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:postfunc-html)', '\\@ref(tab:postfunc-latex)'))` shows the second set of functions in the package and their purpose.

```{r}
postfunc <- tribble(
  ~`function`, ~purpose,
  "st_augment()", "augment the original dataframe with model predictions",
  "st_quickmap_preds()", "generate quick maps of these predictions"
)

```

```{r postfunc-html, eval = knitr::is_html_output(), layout = "center"}
knitr::kable(postfunc, format = "html", caption = "Post-functions: tidy estimates from `mgcv`.") |> 
  kable_styling(full_width = TRUE)
```

```{r postfunc-latex, eval = knitr::is_latex_output()}
knitr::kable(postfunc, format = "latex", caption = "Post-functions: tidy estimates from mgcv.") |> 
  kableExtra::kable_styling(font_size = 9, full_width = FALSE)
```

## st_augment()

This function augments the original dataframe with the estimated means
and standard errors from the \CRANpkg{mgcv} model in a similar manner to
how the \CRANpkg{broom}  [@broom] package operates. The `geometry` column, as per convention, remains as the last column of the augmented dataframe, while the new random effects columns are positioned immediately before it. The estimates which `st_augment()` will extract are random effects (which are called in \CRANpkg{mgcv} with `bs='re'`) and ICAR components (`bs='mrf'`).

Consider the model structure described in the code below using \CRANpkg{mgcv} syntax. In this model $y$ is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a random intercept and slope for the covariate at a *region* level, and an ICAR varying intercept and slope at a lower *sub-region* level.

```{r, eval=FALSE, echo=TRUE}
mgcv::gam(
  y ~ covariate +                   # fixed intercept and effect for covariate  
    s(region, bs = "re") +                  # random intercept at level region
    s(region, covariate, bs = "re") +          # random slopes at level region
    s(sub-region, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k) +                    # ICAR varying intercept at level sub-region
    s(sub-region, by = covariate, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k),           # ICAR varying slope for covariate at level sub-region
  data = data, 
  method = "REML")
```

When labelling the new columns which are augmented to the original dataframe from such a model, `st_augment()` follows the formula syntax of the \CRANpkg{lme4}  [@lme4] package, where the *pipe symbol* (`|`) indicates "*grouped by*". Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:staugtab-html)', '\\@ref(tab:staugtab-latex)'))` shows how the augmented columns in this scenario would be labelled. Each column name begins with either `random.effect.` or `mrf.smooth.` as appropriate. An additional column is also added for the standard error of each estimate. These columns are named as above but with `se.` prepended (e.g. `se.random.effect.region`).

```{r}
st_aug_tab_html <- tribble(
  ~`mgcv syntax`, ~`column name`,
  "s(region, bs = 're')", "random.effect.region",
  "s(region, covariate, bs = 're')", "random.effect.covariate|region",
  "s(sub-region, bs = 'mrf', xt = list(nb = data\\$nb)", "mrf.smooth.sub-region",
  "s(sub-region, by = covariate, bs = 'mrf', xt = list(nb = data\\$nb)", "mrf.smooth.covariate|sub-region"
)
```

```{r}
st_aug_tab_latex <- tribble(
  ~`mgcv syntax`, ~`column name`,
  "s(region, bs = 're')", "random.effect.region",
  "s(region, covariate, bs = 're')", "random.effect.covariate|region",
  "s(sub-region, bs = 'mrf', xt = list(nb = data$nb)", "mrf.smooth.sub-region",
  "s(sub-region, by = covariate, bs = 'mrf', xt = list(nb = data$nb)", "mrf.smooth.covariate|sub-region"
)
```

```{r staugtab-html, eval = knitr::is_html_output(), layout = "center"}
knitr::kable(st_aug_tab_html, format = "html", caption = "The naming procedure for augmented columns from different `mgcv` structures.") |> 
  kable_styling(full_width = TRUE)
```

```{r staugtab-latex, eval = knitr::is_latex_output()}
knitr::kable(st_aug_tab_latex, format = "latex", caption = "The naming procedure for augmented columns from different mgcv structures.") |> 
  kableExtra::kable_styling(font_size = 9, full_width = FALSE)
```

## st_quickmap_preds()

These estimates can then be quickly mapped. As it is possible to include more than 1 varying component, the output of this function is a list of plots. They can be viewed individually by indexing, or all at one using the `plotlist=` argument from the `ggarrange()` function which is part of the \CRANpkg{ggpubr}  [@ggpubr] package. We will see examples of this in the following examples. The maps are automatically titled and subtitled according to the type of effect. For example, the map showing `random.effect.region` will have "*region*" as its title and "*random.effect*" as its subtitle.

# Example 1: Indonesia

Modelling earthquakes in Indonesia seems like a good example to
demonstrate this package. Firstly, Indonesia is composed of many
islands. Secondly, earthquake activity is known to be associated with
the presence of faults which exist below sea level and thus do not
respect land boundaries. Therefore it is reasonable to expect similar
behaviour in nearby provinces regardless of whether or not they are
contiguous. We aim to model the intensity of earthquake activity by
province across Indonesia, controlling for proximity to faults.

## Data

The data for this section have been taken from the [USGS earthquake
catalogue API](https://earthquake.usgs.gov/earthquakes/search/). The
datasets and an explanation are available at
<https://github.com/horankev/quake_data> and record all earthquakes in
and close to Indonesia from 1985-2023. Below is a map of Indonesia, with
other neighbouring or bordering countries filled in grey. The many local
faults which lie within 300km of the shore are shown as green lines
in Figure \@ref(fig:fault-buffers).

```{r, fault-buffers, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Indonesia faults. Surrounded by a 10 kilometre buffer. "}

faults_buf <- faults_df |> 
  st_buffer(dist = 10000)

ggplot() + 
  geom_sf(data=nearby_countries_df, fill="gray50", linewidth=0.5, colour="black") +
  geom_sf(data = provinces_df, fill="antiquewhite", colour="black",linewidth=0.5) +
  geom_sf(data = faults_buf, fill="darkgreen", colour="gray50",linewidth=0.9) +
  geom_sf(data=faults_df, colour="yellow", linewidth=0.4) +
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5),
        axis.title = element_blank(),
        axis.ticks = element_blank())

```

To get an interpretable measure of the concentration of faults in any area
they are transformed from linestrings to polygons by setting a buffer of
10km around them, as shown below. Now both our faults and the sizes of provinces are in units of kilometres
squared. This means we can generate a unitless metric of what proportion of any administrative unit
is covered by these buffered faults. This measure across provinces is shown in Figure
\@ref(fig:fault-conc).

```{r, fault-conc, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Indonesia fault concentration. Square kilometre of buffered fault per square kilometre of province area. "}

ggplot() + 
  geom_sf(data=nearby_countries_df, fill="gray50", colour="black", linewidth=0.5) +
  geom_sf(data=provinces_df, aes(fill=fault_concentration), colour="black", linewidth=0.5) +
  geom_sf(data=faults_df, colour="gray10", linetype="dashed", linewidth=0.5) +
  scale_fill_distiller(palette = "YlGn", direction = 1) + 
  labs(fill="fault\nconcentration") + 
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5),
        legend.position = c(0.92,0.8),
        legend.box.background = element_rect(colour = "black", linewidth = 1))

```

Earthquake activity will be seen as the total number of earthquakes with
their epicentre in each province per unit area of that province. We have
restricted counts to earthquakes of magnitude \>5.5 which is the point
at which they are often labelled as damaging.

The occurrences of these earthquakes are shown in Figure \@ref(fig:quake-occur), their total per province in Figure \@ref(fig:quake-totals), and finally, their count per square kilometre can be seen in Figure \@ref(fig:quake-conc).

```{r, quake-occur, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Earthquakes in Indonesia, 1985-2023. Categorised as medium, large or extra-large. "}

ggplot() + 
  geom_sf(data=nearby_countries_df, fill="gray50", linewidth=0.5, colour="black") +
  geom_sf(data=provinces_df, fill="antiquewhite", colour="black", linewidth=0.5) + 
  geom_sf(data=quakes_df |> filter(magfact != "S") |> filter(province != "undersea"), 
          aes(fill=magfact, size=magfact), shape=21, colour="black") +
  scale_fill_manual(values = c("white","khaki","tomato4")) +
  scale_size_manual(values = c(3,5,7)) + 
  labs(fill="category",
       size="category") + 
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5),
        legend.position = c(0.92,0.8),
        legend.key = element_blank(),
        legend.box.background = element_rect(colour = "black", linewidth = 1))

```

```{r, quake-totals, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Earthquake count in Indonesia, 1985-2023, mag > 5.5: count by province. "}

ggplot() +
  geom_sf(data=nearby_countries_df, fill="gray50", colour="black", linewidth=0.5) +
  geom_sf(data=provinces_df, aes(fill=damaging_quakes_total), colour="black", linewidth=0.5) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1) + 
  labs(fill="count") + 
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5),
        legend.position = c(0.92,0.78),
        legend.box.background = element_rect(colour = "black", linewidth = 1))

```

```{r, quake-conc, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Earthquake incidence in Indonesia, 1985-2023, mag > 5.5: count per square kilometre by province. "}

  ggplot() +
  geom_sf(data=nearby_countries_df, fill="gray50", colour="black", linewidth=0.5) +
  geom_sf(data=provinces_df, aes(fill=damaging_quakes_density), colour="black", linewidth=0.5) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1) + 
  labs(fill="incidence") + 
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5),
  axis.title = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.92,0.78),
        legend.box.background = element_rect(colour = "black", linewidth = 1))

```

## Model

As this is count data, we will model it as a poisson distribution with
$\lambda$ as the mean count per province. For $i = 1,...,n$ provinces,
the dependent variable in this model is
\begin{equation}
y_i = \text{earthquake incidence}_i = \frac{\text{earthquake count in province}_i} {\text{province area}_i}, 
(\#eq:eq1)
\end{equation}
while the explanatory variable is
\begin{equation}
x_i = \text{fault intensity}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
(\#eq:eq2)
\end{equation}
Firstly, when excluding the incidence and just modelling counts, where
$y_i = \text{earthquake count in province}_i$, the poisson model is of
the following form: 
\begin{equation}
y_i | \lambda_i \sim \text{Pois}(\lambda_i)
(\#eq:eq3)
\end{equation}
with
\begin{equation}
E(y_i | \lambda_i) = \lambda_i.
(\#eq:eq4)
\end{equation}
We model
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i.
(\#eq:eq5)
\end{equation}
where $\gamma_i$
is a term with a correlation structure reflecting a province’s location
relative to other provinces.

We can describe these relationships by setting up a neighbourhood
structure based on queen contiguity where a pair of provinces are
considered neighbours if they share at least one point of boundary. This can be modelled as a Markov random field to generate an ICAR model with a spatially varying term. Each of these terms will be correlated the others according to the neighbourhood structure we have defined.

The Markov random field here follows a multivariate Gaussian
distribution. $\gamma_i$ is a vector of province effects having a
distribution with mean **0** and precision **P** where

$[\textbf{p}]_{ij} = v_i$ if $i=j$ and $v_i$ is the number of adjacent provinces
to province $i$,

$[\textbf{p}]_{ij} = -1$ if provinces $i$ and $j$ are adjacent, and

$[\textbf{p}]_{ij} = 0$ otherwise.

A further constraint that $\Sigma_j \gamma_j = 0$ is applied so that the
distribution is identifiable.

If we include an offset term (here, area) because we are more interested
in the incidence than in the actual count, then
\begin{equation}
log(\frac{\lambda_i} {\text{area}_i}) = \beta_0 + \beta_1x_i + \gamma_i
(\#eq:eq6)
\end{equation}
which is equivalent to
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i + log(\text{area}_i).
(\#eq:eq7)
\end{equation}
We are still modelling $log(\lambda)$ rather than the incidence, but we
are adding an offset to adjust for differing areas. Modelling $log(\lambda)$ and adding an offset is equivalent to
modelling incidence, and coefficients can be interpreted that way.

When interpreting the estimated coefficients of the model, it can be
useful to look at it in the following form:
\begin{equation}
\lambda_i = e^{\beta_0 + \beta_1 x_i + \gamma_i} \text{area}_i.
(\#eq:eq8)
\end{equation}

## Pre-functions

Such models, however, can not deal with locations which have no
neighbours. In the case of Indonesia, this is quite problematic. It is
composed of many islands. The estimated count of
islands according to @Andrfout2022 is 13,558.
While it is not unusual for a country to have a number of often small
offshore islands, Indonesia is entirely composed of (at least portions
of) an archipelago of islands, so many of these islands or groups of
islands are individual provinces in their own right. We might like them
to be associated with their k nearest neighbours. However, a nearest
neighbours approach to Indonesia as a whole might not be desired.
Therefore, they must either be excluded from the study or somehow
brought within the neighbourhood framework.

In this case, we use `st_bridges()` for setting up the queen contiguity
structure as usual, but with the additional stipulation that unconnected
units (provinces which are islands or collections of islands) are
considered neighbours to their *k* nearest provinces. For this example,
we have set the value of k to 2. The resulting neighbourhood structure
is shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig1html', 'ind-contig1')`).

(ref:my-caption1) Neighbourhood structure for Indonesian provinces created by `st_bridges()` with `k=2`.

```{r, ind-contig1, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption1)"}
# join islands to k=2 nearest neighbours
# various arguments exist for altering colours and sizes
# additional ggplot themes and layers can be added

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "darkblue", linksize = 0.8, 
                 pointcol = "red", pointsize = 2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=nearby_countries_df, 
          fill="gray50", linewidth=0.5, colour="black")

```

```{r, ind-contig1html, fig.width=12, fig.height=6, echo=FALSE, out.width="100%", eval=knitr::is_html_output(), fig.cap = "Neighbourhood structure for Indonesian provinces created by st_bridges() with k=2. "}
# join islands to k=2 nearest neighbours
# various arguments exist for altering colours and sizes
# additional ggplot themes and layers can be added

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "darkblue", linksize = 0.8, 
                 pointcol = "red", pointsize = 2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=nearby_countries_df, 
          fill="gray50", linewidth=0.5, colour="black")

```

This neighbourhood structure now has no unconnected provinces so it is
suitable for use in an ICAR model. However, we will

-   add some additional contiguities using `st_manual_join_nb()`

-   and remove one using `st_manual_cut_nb()`.

To cater for the possibility that a modeller might not be familiar with the names of the various geographic units but still wishes to make manual alterations to their relationships, we can look at a map (Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig2html', 'ind-contig2')`)) where the nodes are shown by
index number instead of as points (using the argument `nodes='numeric'`). This makes it easy to manually cut
and join neighbours as desired. Furthermore, there is an option to
show concave hulls drawn around each unit (using `concavehull = TRUE`). This is also shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig2html', 'ind-contig2')`). These shapes are not used in the assignment of contiguities but it can be useful to see them in a situation such as Indonesia where many individual provinces are actually multipolygons. Without them, it is not clear whether an island is a province in its own right, or which group of islands together form one province.

(ref:my-caption2) Neighbourhood structure for Indonesian provinces. Viewed with `st_quickmap_nb()`, using the arguments `nodes = 'numeric'` and `concavehull = TRUE`.

```{r, ind-contig2, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption2)"}
# with 'concavehull = TRUE' and 'nodes = "numeric"'

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "tomato", linksize = 0.5, 
                 nodes = "numeric", 
                 numericcol = "black", numericsize = 6, 
                 concavehull = TRUE, 
                 hullcol = "darkgreen", hullsize = 0.2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank())

```

```{r, ind-contig2html, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", eval=knitr::is_html_output(), fig.cap = "Neighbourhood structure for Indonesian provinces. Viewed with st\\_quickmap_nb(), using the arguments nodes = 'numeric' and concavehull = TRUE. "}
# with 'concavehull = TRUE' and 'nodes = "numeric"'

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "tomato", linksize = 0.5, 
                 nodes = "numeric", 
                 numericcol = "black", numericsize = 6, 
                 concavehull = TRUE, 
                 hullcol = "darkgreen", hullsize = 0.2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank())

```

Having made some manual adjustments to the neighbourhood structure, the
new structure can be seen in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig3html', 'ind-contig3')`). Edge effects
have also been mitigated by additionally connecting the two extreme
provinces (1 and 23), which would otherwise have only one neighbour, to
their two next closest neighbours.

(ref:my-caption3) Neighbourhood structure for Indonesian provinces, after alterations using `st_manual_join()` and `st_manual_cut()`.

```{r, ind-contig3, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption3)"}
# a series of manual joins and cuts by index number

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_manual_join_nb(1,24) |> 
  st_manual_join_nb(1,30) |> 
  st_manual_join_nb(3,13) |> 
  st_manual_join_nb(13,17) |> 
  st_manual_join_nb(14,25) |> 
  st_manual_join_nb(20,29) |>
  st_manual_join_nb(19,23) |> 
  st_manual_join_nb(16,27) |> 
  st_manual_join_nb(22,23) |> 
  st_manual_join_nb(7,19) |> 
  st_manual_join_nb(7,20) |>
  st_manual_join_nb(19,28) |> 
  st_manual_join_nb(4,18) |> 
  st_manual_join_nb(21,26) |> 
  st_manual_join_nb(22,28) |> 
  st_manual_cut_nb(19,22) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "darkblue", linksize = 0.8, 
                 pointcol = "red", pointsize = 2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=nearby_countries_df, 
          fill="gray50", linewidth=0.5, colour="black") + 
  annotation_scale()

```

```{r, ind-contig3html, fig.width=12, fig.height=6, echo=FALSE, out.width="100%", eval=knitr::is_html_output(), fig.cap = "Neighbourhood structure for Indonesian provinces, after alterations using st\\_manual_join() and st\\_manual_cut(). "}
# a series of manual joins and cuts by index number

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_manual_join_nb(1,24) |> 
  st_manual_join_nb(1,30) |> 
  st_manual_join_nb(3,13) |> 
  st_manual_join_nb(13,17) |> 
  st_manual_join_nb(14,25) |> 
  st_manual_join_nb(20,29) |>
  st_manual_join_nb(19,23) |> 
  st_manual_join_nb(16,27) |> 
  st_manual_join_nb(22,23) |> 
  st_manual_join_nb(7,19) |> 
  st_manual_join_nb(7,20) |>
  st_manual_join_nb(19,28) |> 
  st_manual_join_nb(4,18) |> 
  st_manual_join_nb(21,26) |> 
  st_manual_join_nb(22,28) |> 
  st_manual_cut_nb(19,22) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "darkblue", linksize = 0.8, 
                 pointcol = "red", pointsize = 2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=nearby_countries_df, 
          fill="gray50", linewidth=0.5, colour="black") + 
  annotation_scale()

```

## `mgcv` model

We now create the ICAR model using \CRANpkg{mgcv}. We will be able to simply pass
the contiguity structure we have created as both the data and a column
of it as the neighbourhood structure.

```{r, fig.width=12, fig.height=6}

provinces_df$province <- factor(provinces_df$province)

prep_data <- st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_manual_join_nb(1,24) |> 
  st_manual_join_nb(1,30) |> 
  st_manual_join_nb(3,13) |> 
  st_manual_join_nb(13,17) |> 
  st_manual_join_nb(14,25) |> 
  st_manual_join_nb(20,29) |>
  st_manual_join_nb(19,23) |> 
  st_manual_join_nb(16,27) |> 
  st_manual_join_nb(22,23) |> 
  st_manual_join_nb(7,19) |> 
  st_manual_join_nb(7,20) |>
  st_manual_join_nb(19,28) |> 
  st_manual_join_nb(4,18) |> 
  st_manual_join_nb(21,26) |> 
  st_manual_join_nb(22,28) |> 
  st_manual_cut_nb(19,22)

```

```{r, echo=TRUE}

mod_pois_mrf <- gam(damaging_quakes_total ~ 
                      fault_concentration +
                      s(province, bs='mrf', xt=list(nb=prep_data$nb), k=22) +
                      offset(log(area_province)),
                    data=prep_data, method="REML",family = "poisson")

```

The adjusted R-squared is **0.957** and deviance explained is **88.2%**. The estimated coefficient associated with fault concentration is
-3.2355. As this is positive, it is greater than 1 when exponentiated:
$$
\text{exp}(0.7531) = 0.04
$$
and this is the expected multiplicative association between a unit
change in fault concentration and incidence of earthquakes in a
province.

Controlling for this estimated mean elevated risk of earthquakes due to
concentration of faults, what is the additional risk level of
earthquakes in a province? This can be seen as a measure of the activity
level of faults locally and it is spatially smoothed by the
autoregressive process. It is mapped as the ICAR component of the model
below:

```{r, eval=FALSE}

# these two are equivalent
AER::dispersiontest(mod_pois_mrf)
# AER::dispersiontest(mod_pois_mrf,trafo = 1)

# these two are equivalent
# AER::dispersiontest(mod_pois_mrf,trafo = function(x) x^2)
# AER::dispersiontest(mod_pois_mrf,trafo = 2)

```

## Post-functions

Use `st_augment()` to get estimates from \CRANpkg{mgcv} model in format
similar to that used by the \CRANpkg{broom} package. For instance, we see from the code below that the original dataframe is now augmented with columns for "mrf.smooth.province" and "mrf.smooth.province" which shows the estimate and standard errors for the $\gamma_i$ component. Note that this is how we would expect them to be named, based on the previous discussion surrounding Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:staugtab-html)', '\\@ref(tab:staugtab-latex)'))`.

```{r, echo=TRUE}
mod_pois_mrf |> 
  st_augment(prep_data) |> 
  names()
```

Use `st_augment()` and `st_quickmap_preds` in the following code to get a quick visualisation of
the estimates for $\gamma_i$, as shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-mrf1html', 'ind-mrf1')`). Again, note that the title and subtitle of the image are as previously discussed.

(ref:my-caption4) Estimates of $\gamma_i$ shown as a map using `st_quickmap_preds()`.

```{r, ind-mrf1, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption4)"}
# st_quickmap_preds() outputs a list of ggplots

plot_mrf <- mod_pois_mrf |> 
  st_augment(prep_data) |>
  st_quickmap_preds(scale_low = "darkgreen",
                    scale_mid = "ivory", 
                    scale_high = "darkred", 
                    scale_midpoint = 0)

# in this case, there is only one plot in the list
# so we call it by index

plot_mrf[[1]] +
  coord_sf(datum=NA) +
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=provinces_df, fill=NA, colour="black", linewidth=0.5) + 
  geom_sf(data=nearby_countries_df, fill="gray50", colour="black", 
          linewidth=0.5) + 
  labs(fill="relative\nincidence") +
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(legend.position = c(0.92,0.77),
        legend.box.background = element_rect(colour = "black", linewidth = 1),
        legend.title = element_text())

```

```{r, ind-mrf1html, fig.width=12, fig.height=6, echo=FALSE, out.width="100%", eval=knitr::is_html_output(), fig.cap = "Estimates of $\\gamma_i$ using st_quickmap_preds(). "}
# st_quickmap_preds() outputs a list of ggplots

plot_mrf <- mod_pois_mrf |> 
  st_augment(prep_data) |>
  st_quickmap_preds(scale_low = "darkgreen",
                    scale_mid = "ivory", 
                    scale_high = "darkred", 
                    scale_midpoint = 0)

# in this case, there is only one plot in the list

plot_mrf[[1]] +
  coord_sf(datum=NA) +
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=provinces_df, fill=NA, colour="black", linewidth=0.5) + 
  geom_sf(data=nearby_countries_df, fill="gray50", colour="black", 
          linewidth=0.5) + 
  labs(fill="relative\nincidence") +
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(legend.position = c(0.92,0.77),
        legend.box.background = element_rect(colour = "black", linewidth = 1),
        legend.title = element_text())

```

If we wish to apply the inverse link function (the exponential function in the case of this poisson model) to map these values to a more interpretable scale, this will not be generated by the function `st_quickmap_preds()`. Instead, we must use the augmented dataframe which is produced by `st_augment()` and create the appropriate extra column with the usual \CRANpkg{tidyverse} `mutate()` function. This allows us to produce the map in Figure
\@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-mrf2html', 'ind-mrf2')`). As these coefficients are multiplicatively related
to the earthquake incidence, values below 1 imply an earthquake
incidence which is lower than expected.

The provinces with the 3 most elevated incidences are labelled in red
We can see that, controlling for the added risk of proximity to
faults, the province of Aceh has 7.9 times the expected
incidence, or number of major earthquakes per square kilometre. The two
lowest-scoring provinces, labelled in green, have essentially no incidence of
earthquake epicentres, controlling for what their proximity to faults alone would
suggest.

(ref:my-caption5) Map showing estimates of $\text{exp}(\gamma_i).$ This is produced by adding an additional column to the dataframe produced by `st_augment()`.

```{r, ind-mrf2, fig.width=12, fig.height=6, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption5)"}

qpois_mod_aug <- mod_pois_mrf |> 
  st_augment(prep_data)

largest <- qpois_mod_aug |> 
  arrange(desc(mrf.smooth.province)) |> 
  head(3)
smallest <- qpois_mod_aug |> 
  arrange(mrf.smooth.province) |> 
  head(2)

ggplot() + 
  geom_sf(data=nearby_countries_df, 
          fill="gray50", colour="black", linewidth=0.5) + 
  geom_sf(data=qpois_mod_aug, aes(fill=exp(mrf.smooth.province)), 
          colour="black", linewidth=0.5) + 
  geom_sf_label_repel(data=largest, 
                      aes(label=paste(province,":",
                                      round(exp(mrf.smooth.province),1))),
                      colour="darkred", size=2.5,
                      nudge_y= 300000) +
  geom_sf_label_repel(data=smallest |> mutate(geometry=st_centroid(geometry)), 
               aes(label=paste(province,":",round(exp(mrf.smooth.province),3))),
               nudge_y= 200000, colour="darkgreen", size=2.5) +
  scale_fill_gradient2(low = "darkgreen", 
                       mid = "ivory", 
                       high = "darkred", 
                       midpoint = 1) +
  scale_colour_gradient2(low = "darkgreen", 
                         mid = "ivory", 
                         high = "darkred", 
                         midpoint = 1) +
  guides(colour="none") + 
  labs(fill="exp.\nrelative\nincidence") + 
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5), 
        axis.title = element_blank(),
        axis.ticks = element_blank(), 
        legend.position = c(0.92,0.76),
        legend.box.background = element_rect(colour = "black", linewidth = 1))

```

```{r, ind-mrf2html, fig.width=12, fig.height=6, out.width="100%", eval=knitr::is_html_output(), fig.cap = "Map showing estimates of exp($\\gamma_i$). This is produced by adding an additional column to the dataframe produced by st_augment()."}

qpois_mod_aug <- mod_pois_mrf |> 
  st_augment(prep_data)

largest <- qpois_mod_aug |> 
  arrange(desc(mrf.smooth.province)) |> 
  head(3)
smallest <- qpois_mod_aug |> 
  arrange(mrf.smooth.province) |> 
  head(2)

ggplot() + 
  geom_sf(data=nearby_countries_df, 
          fill="gray50", colour="black", linewidth=0.5) + 
  geom_sf(data=qpois_mod_aug, aes(fill=exp(mrf.smooth.province)), 
          colour="black", linewidth=0.5) + 
  geom_sf_label_repel(data=largest, 
                      aes(label=paste(province,":",
                                      round(exp(mrf.smooth.province),1))),
                      colour="darkred", size=2.5,
                      nudge_y= 300000) +
  geom_sf_label_repel(data=smallest |> mutate(geometry=st_centroid(geometry)), 
               aes(label=paste(province,":",round(exp(mrf.smooth.province),3))),
               nudge_y= 200000, colour="darkgreen", size=2.5) +
  scale_fill_gradient2(low = "darkgreen", 
                       mid = "ivory", 
                       high = "darkred", 
                       midpoint = 1) +
  scale_colour_gradient2(low = "darkgreen", 
                         mid = "ivory", 
                         high = "darkred", 
                         midpoint = 1) +
  guides(colour="none") + 
  labs(fill="exp.\nrelative\nincidence") + 
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", linewidth=1.5), 
        axis.title = element_blank(),
        axis.ticks = element_blank(), 
        legend.position = c(0.92,0.76),
        legend.box.background = element_rect(colour = "black", linewidth = 1))

```

## Workflow

In this example, we have gone through a number of stages carefully, making changes to contiguities that we deemed appropriate as we went. However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps. A rough and ready model (and visual output) can be generated with `sfislands` using nothing more than three or four lines of code, such as the following:

```{r, eval=FALSE, echo=TRUE}
# set up neighbourhood structure

prep_data <- st_bridges(provinces_df, "province")

# define model

mod <- gam(quake_mlxl_total ~ 
                  fault_concentration +
                  s(province, bs='mrf', xt=list(nb=prep_data$nb), k=22) +
                  offset(log(area_province)),
        data=prep_data, method="REML",family = "poisson")

# augment tidy estimates

tidy_ests <- st_augment(mod, prep_data)

# visualise them

st_quickmap_preds(tidy_ests)
```

# Example 2: London

The next example looks only at using the *pre-functions*, but in a situation where the presence of islands is not the issue. Consider the wards and boroughs of London (sourced from the Greater London Authority's [London Datastore](https://data.london.gov.uk/)). In Figure \@ref(fig:lon-contig1) the `st_bridges()`
function is applied to them to construct a queen contiguity
neighbourhood structure. Because there are no disjoint units (or
"islands"), this will be the same as using `st_contiguity()` from
\CRANpkg{sfdep}. The `st_quickmap_nb()` function gives an immediate
visual representation of the structure. This can be supplemented by
adding a layer showing the course of the river Thames.

```{r, lon-contig1, fig.width=12, fig.height=8, echo=TRUE, out.width="100%", fig.cap = "Wards and boroughs of Greater London. Queen contiguity. "}
# same as sfdep:st_contiguity() as there are no islands
# an extra layer for the river Thames

st_bridges(london, "NAME") |> 
  st_quickmap_nb() + 
  geom_sf(data=thames, colour="blue", linewidth=1.5) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))

```

When the study area has a river running through it there can be problems
with constructing neighbourhood structures. Depending on how the
geometries are defined, the presence of a river can either mean that no
units on either side of its banks are considered neighbours or, at the
other extreme, the river is essentially ignored and all units on
opposing banks are considered neighbours.

Depending on the presence of river crossings, two areas which are
physically quite close might be very distinct. Somebody living on the
banks of a river might be more likely to go about their life primarily
on their side of the river, despite the short distance as the crow flies
of facilities on the other side.

`sfislands` makes it easier to take account of this situation. We can look
at just those boroughs which are on either side of the river Thames (Figure
\@ref(fig:lon-contig2))

```{r, lon-contig2, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", fig.cap = "Riverside boroughs of Greater London. Queen contiguity disregarding river. "}
# which boroughs are alongside the river

riverside <- thames |> st_intersects(london) |> unlist() |> unique()

# only map these boroughs

st_bridges(london[riverside,],"NAME") |> 
  st_quickmap_nb(linksize = 0.5) +
  geom_sf(data=thames, colour="blue", linewidth=1.5) + 
  annotation_scale(location="br") +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))
```

In order to take account of actual connectivity, we can add a layer
showing the road and pedestrian bridges or tunnels. These were scraped from a Wikipedia article describing the various crossings of the river Thames. In Figure \@ref(fig:lon-crossings), we
have also drawn a 1 kilometre buffer around them. Until we get to the
eastern portion of the map, all units on opposing sides of the river are
within 1 kilometre of a crossing. This is not the case for the extreme
eastern units so we might not want to consider these units neighbours if
the river flows between them.

```{r, lon-crossings, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Riverside boroughs of Greater London. Road and pedestrian crossing and tunnels surrounded by 1 kilometre buffer shaded green. "}

ggplot() + 
  geom_sf(data=london[riverside,], fill="gray90", colour="gray50", linewidth=0.5) + 
  geom_sf(data=crossings_roadped |> st_buffer(1000), fill="darkgreen", alpha=0.4) + 
  geom_sf(data=thames, colour="blue", linewidth=1.5) + 
  geom_sf(data=crossings_roadped, size=1, colour="yellow") + 
  geom_sf_label_repel(data=crossings_roadped[1:11,], aes(label=crossing), size=2.5, nudge_y=-10000, fill="white", fontface="bold") + 
  geom_sf_label_repel(data=crossings_roadped[12:22,], aes(label=crossing), size=2.5, nudge_y=10000, fill="white", fontface="bold") +
  annotation_scale(location="br") +
  coord_sf(datum=NA)

```

```{r}
# which riverside units do not fall within any of the buffers...

crossing_buffer_summarised <- crossings_roadped |> 
  st_buffer(1000) |> 
  st_transform(st_crs(london)) |> 
  summarise()

touch_buffer <- st_intersects(london[riverside,],
                              crossing_buffer_summarised)
indices <- c()
for (i in seq_along(touch_buffer)) {
  if (length(touch_buffer[[i]]) == 0) {
    indices <- c(indices, i)
  }
}
indices <- which(sapply(touch_buffer, length) == 0)

no_touch_buffer <- london[riverside,][indices,]
```

In order the identify the changes we wish to make, we use the
`nodes = "numeric"` argument in `st_quickmap_nb()`. Now we can identify
each unit by its position in the contiguity structure. Here we have
shaded in pink the units which are not within 1 kilometre of a river
crossing (see Figure \@ref(fig:lon-contig3)).

```{r, lon-contig3, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", fig.cap = "Riverside boroughs of Greater London. Index number for each borough shown at centroid. Boroughs which are not within 1 kilometre of a crossing are shaded pink. "}
# with 'nodes = "numeric"'

st_bridges(london[riverside,],"NAME") |> 
  st_quickmap_nb(nodes = "numeric", numericsize = 4, linksize = 0.5) +
  geom_sf(data=no_touch_buffer, fill="pink", alpha=0.3) + 
  geom_sf(data=crossings_roadped |> st_buffer(1000), 
          fill="darkgreen", alpha=0.3) +
  geom_sf(data=thames, colour="blue", linewidth=1.5) + 
  geom_sf(data=crossings_roadped, size=1, colour="yellow") + 
  annotation_scale(location="br") +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))
```

This allows us to easily cut the ties across the river for these units
manually by using the function `st_manual_cut_nb()`. The results of this editing procedure are shown in Figure \@ref(fig:lon-contig4).
While we are using the index of the units in this example, the function also
accepts names as arguments which may be more convenient in some circumstances. Having made these adjustments,
`st_quickmap_nb()` now shows a connectivity structure (Figure \@ref(fig:lon-contig4)) which reflects the influence of the
presence or absence of river crossings.

This example shows that the pre-functions of `sfislands` have uses for situations which do not involve islands. They can be used to apply domain knowledge to easily design the most appropriate neighbourhood structure.

```{r, lon-contig4, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", fig.cap = "Riverside boroughs of Greater London. Contiguities across the river have been cut for the pink boroughs. "}
# manually cut the links where there is no crossing

st_bridges(london[riverside,],"NAME") |> 
  st_manual_cut_nb(18,17) |> 
  st_manual_cut_nb(19,17) |> 
  st_manual_cut_nb(19,20) |> 
  st_manual_cut_nb(20,21) |> 
  st_manual_cut_nb(21,22) |> 
  st_manual_cut_nb(22,48) |> 
  st_manual_cut_nb(47,48) |> 
  st_manual_cut_nb(45,46) |> 
  st_manual_cut_nb(45,47) |>  
  st_manual_cut_nb(39,65) |> 
  st_manual_cut_nb(1,2) |> 
  st_quickmap_nb(bordercol = "black", bordersize = 0.5, linksize = 0.5) +
  geom_sf(data=no_touch_buffer, fill="pink", alpha=0.3) + 
  geom_sf(data=crossings_roadped |> st_buffer(1000), 
          fill="darkgreen", alpha=0.3) +
  geom_sf(data=thames, colour="blue", linewidth=1.5) +
  annotation_scale(location="br") +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))
```

# Example 3: Liverpool

In this final example, there are no islands or river issues. We saw in the Indonesia example how `sfislands` can extract estimates from the `mgcv` argument `bs="mrf"`. It can also deal with random effects as defined by `bs="re"` in `mgcv` models. Here, we
will construct a hierarchical model in this way which also features an ICAR component. We will
compare two different types of contiguity which could be applied:

* `st_bridges()` from `sfislands`, and 
* `st_knn()` from \CRANpkg{sfdep}.

We will show how their outcomes can be quickly mapped to give an immediate sense of the differences.

```{r}
livadmin <- tribble(
  ~code, ~`area type`, ~description,
  "OA", "output areas", "lowest level of geographical area for census statistics, usually containing 100 - 625 persons",
  "LSOA", "lower layer super output areas", "usually 4 or 5 OAs",
  "MSOA", "middle layer super output areas", "usually 4 or 5 LSOAs"
)

```

```{r livadmin-html, eval = knitr::is_html_output(), layout = "center"}
knitr::kable(livadmin, format = "html", caption = "Three nested levels of administrative divisions in Liverpool.") |> 
  kable_styling(full_width = TRUE)
```

```{r livadmin-latex, eval = knitr::is_latex_output()}
knitr::kable(livadmin, format = "latex", caption = "Three nested levels of administrative divisions in Liverpool.") |> 
  kableExtra::kable_styling(font_size = 8) |> 
  column_spec(3, width = "7cm")
```

```{r, liv-nested, fig.width=10, fig.height=5, out.width="100%", fig.cap = "Nested levels of administrative divisions in Liverpool. "}

ggarrange(
  ggplot() + 
    geom_sf(data=msoa_sf, fill="gray90", colour="black") +  
    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
    geom_sf(data=liverpool |> summarise(), colour="black", fill=NA) + 
    labs(title = "level 1: MSOA polygons") + 
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()),

  ggplot() + 
    geom_sf(data=lsoa_sf, fill="gray90", colour="gray40") +  
    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
    geom_sf(data=liverpool |> summarise(), colour="black", fill=NA) + 
    labs(title = "level 2: LSOA polygons") + 
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()),

  ggplot() + 
    geom_sf(data=liverpool |> summarise(), colour="black", fill="white") + 
    geom_sf(data=liverpool |> st_centroid(), 
            fill="gray90", colour="gray50", size=0.1) +  
    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
    labs(title = "level 3: OA (shown as centroids)") + 
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()),

  ncol=3
)

```

```{r, liv-vars, fig.width=6, fig.height=4, out.width="70%", fig.cap = "Unemployment and limiting long-term illness rates in Liverpool, with extent of Mersey estuary in light blue to the south west. "}

ggarrange(
  ggplot() + 
    geom_sf(data=liverpool, aes(fill=unemployment, colour=unemployment)) + 
    geom_sf(data=liverpool |> summarise(), 
            fill=NA, colour="gray30", linewidth=0.2) +  
    geom_sf(data=mersey, fill="lightblue", colour="gray30", linewidth=0.2) + 
    scale_fill_distiller(palette = "PuBu", direction = 1) + 
    scale_colour_distiller(palette = "PuBu", direction = 1) + 
    labs(fill="unemployment",
         colour="unemployment") + 
    coord_sf(datum=NA) + 
    annotation_scale(location = "bl", width_hint = 0.4) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.0, "in"), pad_y = unit(0.2, "in"),
        style = north_arrow_fancy_orienteering) +
    theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
                                          linewidth=0.6), 
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.text = element_text(size=8),
          legend.title = element_text(size=8)),
  ggplot() + 
    geom_sf(data=liverpool, aes(fill=lt_illness, colour=lt_illness)) + 
    geom_sf(data=liverpool |> summarise(), 
            fill=NA, colour="gray30", linewidth=0.2) +  
    geom_sf(data=mersey, fill="lightblue", colour="gray30", linewidth=0.2) + 
    scale_fill_distiller(palette = "RdYlBu", direction = -1) + 
    scale_colour_distiller(palette = "RdYlBu", direction = -1) + 
    labs(fill="limiting long-term\nillness\n(z-score)",
         colour="limiting long-term\nillness\n(z-score)") + 
    coord_sf(datum=NA) +
    annotation_scale(location = "bl", width_hint = 0.4) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.0, "in"), pad_y = unit(0.2, "in"),
        style = north_arrow_fancy_orienteering) +
    theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
                                          linewidth=0.6), 
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.text = element_text(size=8),
          legend.title = element_text(size=8)),

  ncol=2, legend = "bottom"
)

```

```{r, liv-contigs, fig.width=5, fig.height=5, out.width="60%", echo=TRUE, fig.cap = "First order queen contiguity of output areas (OAs) in Liverpool with magnified inset. "}
# magnify a section

from <- c(xmin = -2.97, xmax = -2.95, ymin = 53.38, ymax = 53.39)
to <- c(-3.01,-2.96,53.32,53.35)

st_bridges(liverpool, "oa_cd") |> 
  st_quickmap_nb(pointsize = 0.1,
                 linksize = 0.1,
                 fillcol = "gray90") +  
  geom_sf(data=mersey, fill="lightblue", colour="gray30", 
          linewidth=0.2, alpha=0.5) + 
  geom_magnify(data=liverpool, 
               from = from, to = to, 
               corners = 0.1, shadow = TRUE, linewidth = 0.6) + 
  theme(panel.background = element_rect(fill = "#F7FFF2",
                                        colour = "black", linewidth=1))

```

```{r}

prepliv <- st_bridges(liverpool, "oa_cd")

# liverpool_model2 <- gam(unemployment ~ lt_illness + 
#                          s(msoa_cd, bs="re") + 
#                          s(msoa_cd, lt_illness, bs="re") +
#                          s(lsoa_cd, bs="re") + 
#                          s(lsoa_cd, lt_illness, bs="re"), # ICAR constituency ICAR varying coefficients
#                        data=prepliv, method="REML")

# saveRDS(liverpool_model2, here("data","liverpool_model2.rds"))

liverpool_hmodel <- readRDS(here("data","liverpool_model2.rds"))

# summary(liverpool_hmodel)

```

```{r}
# liverpool_hsamodel <- gam(unemployment ~ lt_illness +
#                          s(msoa_cd, bs="re") +
#                          s(msoa_cd, lt_illness, bs="re") +
#                          s(lsoa_cd, bs="re") +
#                          s(lsoa_cd, lt_illness, bs="re") +
#                          s(oa_cd, bs='mrf',
#                            xt=list(nb=prepliv$nb),k=20), # ICAR constituency ICAR varying coefficients
#                        data=prepliv, method="REML")

# saveRDS(liverpool_hsamodel, here("data","liverpool_hsamodel.rds")

liverpool_hsamodel <- readRDS(here("data","liverpool_hsamodel.rds"))

# summary(liverpool_hsamodel)

```

```{r}

liverpool_hmodel_aug <- liverpool_hmodel |> 
  st_augment(prepliv) # pipe into function to get estimates

liverpool_hsamodel_aug <- liverpool_hsamodel |> 
  st_augment(prepliv) # pipe into function to get estimates

```

```{r}

liverpool_hmodel_plots <- liverpool_hmodel |> 
  st_augment(prepliv) |> # pipe into function to get estimates
  st_quickmap_preds(borderwidth = NA, 
                    scale_low = "darkblue",
                    scale_mid = "gray75",
                    scale_high = "darkred",
                    scale_midpoint = 0, 
                    framefill = "#F7FFF2",
                    frameline = "black",
                    framesize = 1) # pipe into this for visualisation

```

This example is based on data from the [Office of National Statistics](https://www.nomisweb.co.uk/home/census2001.asp) as presented by @liverpool-notes in their demonstration of techniques for hierarchical spatial regression. They use the different levels of nested administrative boundaries in Liverpool for their demonstration. These are described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:livadmin-html)', '\\@ref(tab:livadmin-latex)'))`, and are mapped separately in Figure \@ref(fig:liv-nested).

Now suppose we wish to model unemployment as a function of limited long-term
illness, taking into account this hierarchical structure. Limiting long-term illness (LLTI) refers to a health problem or disability which limits a person's day-to-day activities [@Barnett2001]. We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at MSOA and LSOA levels. The
distribution of these variables across the OAs of Liverpool are shown in Figure
\@ref(fig:liv-vars).

If $y_{ijk}$ and $x_{ijk}$ are respectively the levels of unemployment and LLTI in each of $i$ OAs within $j$ LSOAs, which in turn are within $k$ MSOAs, then 
\begin{equation}
y_{ijk} = \beta_{0jk} + \beta_{1jk}x_{ijk} + \gamma_{ijk} + \epsilon_{ijk}
(\#eq:eq9)
\end{equation}
\begin{equation}
\beta_{0jk} = \beta_{0k} + u_{0jk}
(\#eq:eq10)
\end{equation}
\begin{equation}
\beta_{1jk} = \beta_{1k} + u_{1jk}
(\#eq:eq11)
\end{equation}
\begin{equation}
\beta_{0k} = \beta_0 + w_{0k}
(\#eq:eq12)
\end{equation}
\begin{equation}
\beta_{1k} = \beta_1 + w_{1k}
(\#eq:eq13)
\end{equation}
\begin{equation}
\epsilon_{ijk} \sim N(0,\sigma_{\epsilon}^2)
(\#eq:eq14)
\end{equation}
\begin{equation}
u_{jk} \sim N(0,\sigma_{u}^2)
(\#eq:eq15)
\end{equation}
\begin{equation}
 w_{k} \sim N(0,\sigma_{w}^2)
(\#eq:eq16)
\end{equation}
where the $\gamma_{ijk}$s comprise a vector of ICAR components for each OA, similar to the provinces in the earlier Indonesia model. The vector has a multivariate normal distribution with mean 0 and precision related to the contiguity structure as before.

There are no islands or discontiguities in this geometry of Liverpool so
using `st_bridges()` will give the same structure as would
`sfdep::st_contiguity()` (see Figure \@ref(fig:liv-contigs)).

The hierarchical model with no mrf component produces estimates as shown
in Figure \@ref(fig:liv-est1).

```{r, liv-est1, fig.width=12, fig.height=3, out.width="100%", fig.cap = "Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas as levels. "}
# put an empty plot on the right to line up with those with an extra lowest level

ggarrange(
  plotlist = liverpool_hmodel_plots,
  nrow = 1, legend = "right"
)

```

```{r}

liverpool_hsamodel_plots <- liverpool_hsamodel |> 
  st_augment(prepliv) |> # pipe into function to get estimates
  st_quickmap_preds(borderwidth = NA, 
                    # legendlimits = "minmax",
                    scale_low = "darkblue",
                    scale_mid = "gray75",
                    scale_high = "darkred",
                    scale_midpoint = 0, 
                    framefill = "#F7FFF2",
                    frameline = "black",
                    framesize = 1) # pipe into this for visualisation

```

If we instead use contiguity as the basis of an additional spatial term,
we get estimates shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'liv-est2html', 'liv-est2')`).

(ref:my-caption6) Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas are the top two levels respectively (MSOA & LSOA). Output areas (OA) as lowest level Markov random field smooth using first-degree queen contiguity from `st_bridges()`.

```{r, liv-est2, fig.width=12, fig.height=6, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption6)"}

ggarrange(
  ggarrange(
    liverpool_hsamodel_plots[[1]],
    liverpool_hsamodel_plots[[2]],
    liverpool_hsamodel_plots[[3]],
    liverpool_hsamodel_plots[[4]],
    nrow = 1
  ),
  liverpool_hsamodel_plots[[5]],
  nrow = 2
)

```

```{r, liv-est2html, fig.width=12, fig.height=6, out.width="100%", eval=knitr::is_html_output(), fig.cap = "Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas are the top two levels respectively (MSOA & LSOA). Output areas (OA) as lowest level Markov random field smooth using first-degree queen contiguity from st\\_bridges(). "}

ggarrange(
  ggarrange(
    liverpool_hsamodel_plots[[1]],
    liverpool_hsamodel_plots[[2]],
    liverpool_hsamodel_plots[[3]],
    liverpool_hsamodel_plots[[4]],
    nrow = 1
  ),
  liverpool_hsamodel_plots[[5]],
  nrow = 2
)

```

```{r}

# 0.7km distance band
prepliv <- liverpool |> 
  mutate(nb = st_centroid(liverpool) |> 
              st_dist_band(upper = 0.7))
names(prepliv$nb) <- prepliv$oa_cd

# liverpool_hsamodel_dist_band <- gam(unemployment ~ lt_illness +
#                          s(msoa_cd, bs="re") +
#                          s(msoa_cd, lt_illness, bs="re") +
#                          s(lsoa_cd, bs="re") +
#                          s(lsoa_cd, lt_illness, bs="re") +
#                          s(oa_cd, bs='mrf',
#                            xt=list(nb=prepliv$nb),k=20), # ICAR constituency ICAR varying coefficients
#                        data=prepliv, method="REML")

# saveRDS(liverpool_hsamodel_dist_band, here("data","liverpool_hsamodel_dist_band.rds"))

liverpool_hsamodel_dist_band <- readRDS(here("data","liverpool_hsamodel_dist_band.rds"))

# summary(liverpool_hsamodel_dist_band)

```

```{r}

liverpool_hsamodel_dist_band_plots <- liverpool_hsamodel_dist_band |> 
  st_augment(prepliv) |> # pipe into function to get estimates
  st_quickmap_preds(borderwidth = NA, 
                    scale_low = "darkblue",
                    scale_mid = "gray75",
                    scale_high = "darkred",
                    scale_midpoint = 0, 
                    framefill = "#F7FFF2",
                    frameline = "black",
                    framesize = 1) # pipe into this for visualisation

```

If instead we consider units to be neighbours if they lie within a
distance band of 700 metres, we get estimates shown in Figure
\@ref(fig:`r ifelse(knitr::is_html_output(), 'liv-est3html', 'liv-est3')`).

A comparison of their AICs (see Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:aic-html)', '\\@ref(tab:aic-latex)'))`) suggests that, in this scenario, queen contiguity is more appropriate than the distance band condition.

(ref:my-caption7) Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas are the top two levels respectively (MSOA & LSOA). Output areas (OA) as lowest level Markov random field smooth using `sfdep::st_dist_band()` with radius of 700 metres.

```{r, liv-est3, fig.width=12, fig.height=6, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption7)"}

ggarrange(
  ggarrange(
    liverpool_hsamodel_dist_band_plots[[1]],
    liverpool_hsamodel_dist_band_plots[[2]],
    liverpool_hsamodel_dist_band_plots[[3]],
    liverpool_hsamodel_dist_band_plots[[4]],
    nrow = 1
  ),
  liverpool_hsamodel_dist_band_plots[[5]],
  nrow = 2
)

```

```{r, liv-est3html, fig.width=12, fig.height=6, out.width="100%", eval=knitr::is_html_output(), fig.cap = "Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas are the top two levels respectively (MSOA & LSOA). Output areas (OA) as lowest level Markov random field smooth using sfdep\\:\\:st\\_dist_band() with radius of 700 metres."}

ggarrange(
  ggarrange(
    liverpool_hsamodel_dist_band_plots[[1]],
    liverpool_hsamodel_dist_band_plots[[2]],
    liverpool_hsamodel_dist_band_plots[[3]],
    liverpool_hsamodel_dist_band_plots[[4]],
    nrow = 1
  ),
  liverpool_hsamodel_dist_band_plots[[5]],
  nrow = 2
)

```

```{r}

aic_tab <- AIC(liverpool_hmodel, liverpool_hsamodel, liverpool_hsamodel_dist_band) |> 
  data.frame() |> 
  mutate(`relative AIC` = round(-AIC / min(AIC),3),
         model = c("hierachical model, no ICAR",
                   "hierachical model, ICAR based on st_bridges()",
                   "hierachical model, ICAR based on sfdep::st_dist_band()")) |> 
  rownames_to_column(var="modcode") |> 
  arrange(AIC) |> 
  select(model, AIC, `relative AIC`)

```

```{r aic-html, eval = knitr::is_html_output(), layout = "center"}
knitr::kable(aic_tab, format = "html", caption = "Comparison of models by AIC.") |> 
  kable_styling(full_width = TRUE)
```

```{r aic-latex, eval = knitr::is_latex_output()}
knitr::kable(aic_tab, format = "latex", caption = "Comparison of models by AIC.") |> 
  kableExtra::kable_styling(font_size = 8)
```

# Summary

These examples have shown the varying scenarios in which `sfislands` can be useful. It aims to contribute to spatial modelling by making an awkward area less awkward. Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included. It also provides helper functions to use these structures in spatial regression models built with `mgcv` which streamline the human effort necessary to examine the estimates.

# Acknowledgements

This publication has emanated from research conducted with the financial support of Science Foundation Ireland under Grant number 18/CRT/6049.

### 2024-03-03 article.tex
initial commit

% !TeX root = RJwrapper.tex
\title{sfislands: An R Package for Accommodating Islands and Disjoint Zones in Areal Spatial Modelling}

\author{by Kevin Horan, Katarina Domijan, and Chris Brunsdon}

\maketitle

\abstract{%
Fitting areal spatial models can be a cumbersome task, particularly when the geographical units are not well-behaved. The presence of islands gives rise to particular issues when creating neighbourhood structures based on contiguity. We might also want to edit the output of a condition-based neighbourhood structure. It might be appropriate to add or remove connections to take account of infrastructure links or other domain-specific issues which are not apparent from a basic set of polygons. This package streamlines these processes with tools to account for the presence of islands and to intuitively make manual alterations to neighbourhood structures with the help of convenient mapping functions. Beyond the setting up of a satisfactory neighbourhood structure, it has further helper functions specific to the \texttt{mgcv} package which reduce the human workload required to extract estimates from models with spatial and hierarchical structures.
}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

A key feature which differentiates spatial statistics is the
non-independence of observations and the expectation that neighbouring
units will be more similar than non-neighbouring ones (Tobler 1970). If this
is not accounted for, the assumptions of many types of models will be
violated. The relationships between all spatial units in a study can be
represented numerically in a spatial weights matrix. In order to build
this, we must first decide on what constitutes being a neighbour. We can
see this as a continuous relationship where degree of neighbourliness is
a function of connectivity, which could be represented as some measure
of distance. Alternatively it could be a binary situation where each
pair of units either are (1) or are not (0) neighbours. This can be
based on a condition such as contiguity of some sort, or a distance
constraint. It is the job of the modeller to formulate a hypothesis
which justifies their choice of neighbourhood structure.

For R users, the \CRANpkg{spdep} package (Bivand et al. 2005) has long been popular for the
creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} (Pebesma 2018) package (Parry and Locke 2024) has presented the same
functionality in addition to extra features in a tidy structure based on
lists in a similar way to \CRANpkg{sf}.

The most appropriate form of neighbourhood structure will depend on the specific context. Briz-Redón et al. (2021) compared different structures in the context of COVID-19 data. They note that Earnest et al. (2007) found that distance-based matrices were more appropriate when examining birth defects in Australia, whereas Duncan, White, and Mengersen (2017) found that a first-order contiguity structure produced a better fit than others in the context of lip cancer incidence in Scotland.

The most commonly used neighbourhood structure is one based on first-order queen contiguity, where units are considered neighbours if they
share at least a vortex of boundary. However, as the name suggests, this
will lead to problems when non-contiguous units such as islands or
exclaves are present. Less obviously, depending on how the geographic
units are described, areas on either sides of rivers may be
inappropriately classified as neighbours or not neighbours. Furthermore,
the presence of infrastructure such as tunnels, bridges or ferry
services might be satisfactory to meet our hypothesis of the required
degree of connectivity to be considered neighbours.

The aim of \texttt{sfislands} is to deal with the situations described above in a
convenient and open manner. It allows us to set up a structure, quickly
map it, and then examine it. The structure can then be edited and the process re-iterated until we have described a
spatial relationship structure with which we are satisfied.

The above can be considered as the \emph{pre-functions} of the package. A
further set of feature of \texttt{sfislands}, which we refer to as \emph{post-functions},
are for use with the \CRANpkg{mgcv} package (Wood 2011). It is straightforward to use
\texttt{sfislands} neighbourhood structures in models with \CRANpkg{mgcv} but it can be quite awkward
to extract the estimates. These \emph{post-functions} conveniently provide the estimates in tidy format of random effects and intrinsic conditional autoregressive (ICAR) components from such models and give the ability to quickly map them for visualisation purposes.

\hypertarget{typical-use-cases}{%
\subsection{Typical use-cases}\label{typical-use-cases}}

In this paper, we will look at three examples to show different use-cases for
\texttt{sfislands}. The first example focuses on earthquakes in
Indonesia. It shows a scenario where all of the functions are used, from
setting up contiguities to modelling and examining the estimates of the
model.

The second example looks at London and how, despite the lack of islands,
the presence of a river means that some of the pre-functions of
\texttt{sfislands} can be useful.

The final example focuses on Liverpool. There are no islands or issues
of discontiguity in this dataset. Instead, we will show how the package
can be used to fit a multilevel model with an ICAR component at the lowest
level. We show how the results can be presented in tidy form and quickly
visualised. We also show how different types of contiguity from the
\CRANpkg{sfdep} package can used within the \CRANpkg{mgcv} modelling
framework and how the estimates derived from them can be conveniently
compared.

\hypertarget{why-use-sfislands}{%
\section{\texorpdfstring{Why use \texttt{sfislands}?}{Why use sfislands?}}\label{why-use-sfislands}}

It greatly speeds up the workflow of fitting areal spatial models.

\begin{itemize}
\item
  \textbf{pre-functions} for setting up contiguities:

  \begin{itemize}
  \item
    it addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data. \texttt{sfislands} will include a contiguity for all units,
  \item
    it gives tools to immediately visualise this structure as a map,
  \item
    these maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax,
  \item
    as the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence,
  \item
    connections made by the package which are not contiguities can be accessed to ensure openness in the process.
  \end{itemize}
\item
  For modelling:

  \begin{itemize}
  \tightlist
  \item
    these neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms}. (Bürkner 2017), \pkg{r-inla} (Bakka et al. 2018) and more.
  \end{itemize}
\item
  \textbf{post-functions} for \texttt{mgcv} models:

  \begin{itemize}
  \item
    it simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models),
  \item
    these effects can be quickly visualised in \CRANpkg{ggplot2} maps.
  \end{itemize}
\end{itemize}

\hypertarget{pre-functions}{%
\section{Pre-functions}\label{pre-functions}}

The first group of functions, shown in Table \ref{tab:prefunc-latex}, deals with the creation
of a contiguity structure in the presence of discontiguities. The
resultant structure can be quickly mapped to check if it is
satisfactory. Connections can be manually added or removed by name or
index number. By an iterative process of changes and examination of a
quick map, a satisfactory structure can be decided upon.

Table \ref{tab:prefunc-latex} shows these functions and their purpose.

\begin{table}

\caption{\label{tab:prefunc-latex}Pre-functions: setting up a neighbourhood structure.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|>{\raggedright\arraybackslash}p{7cm}}
\hline
function & purpose\\
\hline
st\_bridges() & create a neighbourhood contiguity structure, with a k-nearest neighbours condition for islands\\
\hline
st\_quickmap\_nb() & check structure visually on map\\
\hline
st\_check\_islands() & check the contiguities which have been assigned to islands\\
\hline
st\_manual\_join\_nb() & make manual changes by adding connections\\
\hline
st\_manual\_cut\_nb() & make manual changes by removing connections\\
\hline
\end{tabular}
\end{table}

They deal with the creation
of a contiguity structure in the presence of discontiguities. The
resultant structure can be quickly mapped to check if it is
satisfactory. Connections can be manually added or removed by name or
index number. By an iterative process of changes and examination of a
quick map, a satisfactory structure can be decided upon.
Going through each function in more detail:

Use the rectangles shown in Figure \ref{fig:rects1} as a simple demonstration. Rectangles 1-3
are contiguous while 4 and 5 are islands.

\begin{figure}

{\centering \includegraphics{article_files/figure-latex/rects1-1}

}

\caption{Simplified scenario with five rectangles. }\label{fig:rects1}
\end{figure}

\hypertarget{st_bridges}{%
\subsection{st\_bridges()}\label{st_bridges}}

This function requires as arguments an \CRANpkg{sf} dataframe and the name of a column of unique identifiers, preferably names, of each spatial unit. The k-nearest neighbours for connecting islands can then be specified. The output is a \emph{named} neighbourhood structure in either list or matrix form as desired, which can be either a standalone object or included as an additional column in the original \CRANpkg{sf} dataframe. While it is not necessary in all modelling packages for the neighbourhood list or matrix to be \emph{named}, it is good practice to do so and is mandatory when using, for example, \CRANpkg{mgcv}.

In the following examples, we have chosen to ignore islands with the argument \texttt{remove\_islands\ =\ TRUE} and to return a list and matrix structure respectively by specifying this in the \texttt{nb\_structure} argument and choosing \texttt{add\_to\_dataframe\ =\ FALSE}:

\begin{verbatim}
# output a named list

st_bridges(rectangles, 
           "name", 
           remove_islands = TRUE, 
           nb_structure = "list", 
           add_to_dataframe = FALSE) |>
  head()
\end{verbatim}

\begin{verbatim}
#> $Rect1
#> [1] 2 3
#> 
#> $Rect2
#> [1] 1 3
#> 
#> $Rect3
#> [1] 1 2
\end{verbatim}

\begin{verbatim}
# output a named matrix

st_bridges(rectangles, 
           "name", 
           remove_islands = TRUE, 
           nb_structure = "matrix", 
           add_to_dataframe = FALSE) |> 
  head()
\end{verbatim}

\begin{verbatim}
#>       [,1] [,2] [,3]
#> Rect1    0    1    1
#> Rect2    1    0    1
#> Rect3    1    1    0
\end{verbatim}

In these examples, we choose to join islands to their 1 nearest
neighbour, which is the default setting, and to return the output as a column called ``nb'' in the original
\CRANpkg{sf} dataframe (this is the default setting):

\begin{verbatim}
# output a named list as a column "nb" in original dataframe

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1, 
           nb_structure = "list") |> 
  head()
\end{verbatim}

\begin{verbatim}
#> Simple feature collection with 5 features and 2 fields
#> Geometry type: POLYGON
#> Dimension:     XY
#> Bounding box:  xmin: 0 ymin: 0 xmax: 6 ymax: 4
#> CRS:           NA
#>                         geometry  name      nb
#> 1 POLYGON ((0 0, 0 2, 2 2, 2 ... Rect1    2, 3
#> 2 POLYGON ((2 0, 2 2, 4 2, 4 ... Rect2 1, 3, 4
#> 3 POLYGON ((2 2, 2 4, 4 4, 4 ... Rect3 1, 2, 5
#> 4 POLYGON ((5 0, 5 1, 6 1, 6 ... Rect4       2
#> 5 POLYGON ((0.8 3, 0.8 4, 1.8... Rect5       3
\end{verbatim}

\begin{verbatim}
# output a named matrix as a column "nb" in original dataframe

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1, 
           nb_structure = "matrix") |> 
  head()
\end{verbatim}

\begin{verbatim}
#> Simple feature collection with 5 features and 2 fields
#> Geometry type: POLYGON
#> Dimension:     XY
#> Bounding box:  xmin: 0 ymin: 0 xmax: 6 ymax: 4
#> CRS:           NA
#>                         geometry  name nb.1 nb.2 nb.3 nb.4 nb.5
#> 1 POLYGON ((0 0, 0 2, 2 2, 2 ... Rect1    0    1    1    0    0
#> 2 POLYGON ((2 0, 2 2, 4 2, 4 ... Rect2    1    0    1    1    0
#> 3 POLYGON ((2 2, 2 4, 4 4, 4 ... Rect3    1    1    0    0    1
#> 4 POLYGON ((5 0, 5 1, 6 1, 6 ... Rect4    0    1    0    0    0
#> 5 POLYGON ((0.8 3, 0.8 4, 1.8... Rect5    0    0    1    0    0
\end{verbatim}

These structures can serve as the input to models in \CRANpkg{brms},
\pkg{r-inla}, \pkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. Rather than having a separate neighbours
column, it is included as a named list or matrix in the original \CRANpkg{sf}
dataframe, in the spirit of the \CRANpkg{sfdep} package.

\hypertarget{st_quickmap_nb}{%
\subsection{st\_quickmap\_nb()}\label{st_quickmap_nb}}

It is much more intuitive to examine these structures visually than in
matrix or list format. This can be done with the \texttt{st\_quickmap\_nb()}
function as shown in Figure \ref{fig:rects2}.

\begin{verbatim}
# default is 'nodes = "point"'

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_quickmap_nb()
\end{verbatim}

\begin{figure}

{\centering \includegraphics{article_files/figure-latex/rects2-1}

}

\caption{Queen contiguity and islands connected to nearest neighbour. }\label{fig:rects2}
\end{figure}

If we wish to make edits, it might be more useful to represent the nodes
numerically rather than as points (Figure \ref{fig:rects3}).

\begin{verbatim}
# with 'nodes = "numeric"'

st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_quickmap_nb(nodes = "numeric")
\end{verbatim}

\begin{figure}

{\centering \includegraphics{article_files/figure-latex/rects3-1}

}

\caption{Queen contiguity and islands connected to nearest neighbour. Nodes are shown as numeric indices. }\label{fig:rects3}
\end{figure}

\hypertarget{st_check_islands}{%
\subsection{st\_check\_islands()}\label{st_check_islands}}

This function will tell us openly what connections have been made which
are not based on contiguity. It gives both the name and index number of each pair of added connections.

\begin{verbatim}
st_bridges(rectangles, 
           "name", 
           link_islands_k = 1) |> 
  st_check_islands()
\end{verbatim}

\begin{verbatim}
#>   island_names island_num nb_num nb_names
#> 1        Rect4          4      2    Rect2
#> 2        Rect5          5      3    Rect3
\end{verbatim}

\hypertarget{st_manual_join_nb}{%
\subsection{st\_manual\_join\_nb()}\label{st_manual_join_nb}}

If we feel that 4 should also be connected to 3, this can be done
manually (Figure \ref{fig:rects4}).

\begin{verbatim}
st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_quickmap_nb(nodes = "numeric")
\end{verbatim}

\begin{figure}

{\centering \includegraphics{article_files/figure-latex/rects4-1}

}

\caption{With an additional connection between 3 and 4. }\label{fig:rects4}
\end{figure}

\hypertarget{st_manual_cut_nb}{%
\subsection{st\_manual\_cut\_nb()}\label{st_manual_cut_nb}}

And perhaps there is a wide river between rectangles 1 and 2 which
justifies removing the connection. We will edit it this time using
names (Figure \ref{fig:rects5}).

\begin{verbatim}
st_bridges(rectangles, "name", 
           link_islands_k = 1) |> 
  st_manual_join_nb(3,4) |> 
  st_manual_cut_nb("Rect1","Rect2") |> 
  st_quickmap_nb(nodes = "numeric")
\end{verbatim}

\begin{figure}

{\centering \includegraphics{article_files/figure-latex/rects5-1}

}

\caption{With an additional connection between 1 and 2. }\label{fig:rects5}
\end{figure}

The use of these structures is particularly common in CAR-type models.
These are often implemented in a Bayesian framework using
\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. The pre-functions will output the
neighbourhood structure in the desired format for use in any of these
frameworks. A convenient frequentist alternative is to use the
\CRANpkg{mgcv} package. It has the functionality to create such models
using \texttt{bs="mrf"}. It also has the ability to combine these with a
hierarchical structure using \texttt{bs="re"}. While the outputs from the
Bayesian structures mentioned above can be extracted in the same way as
any other component of the model, it can be a bit awkward to get the
estimates from \CRANpkg{mgcv} model. \texttt{sfislands} has two post-functions to conveniently extract and visualise these.

\hypertarget{post-functions}{%
\section{Post-functions}\label{post-functions}}

Table \ref{tab:postfunc-latex} shows the second set of functions in the package and their purpose.

\begin{table}

\caption{\label{tab:postfunc-latex}Post-functions: tidy estimates from mgcv.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|l}
\hline
function & purpose\\
\hline
st\_augment() & augment the original dataframe with model predictions\\
\hline
st\_quickmap\_preds() & generate quick maps of these predictions\\
\hline
\end{tabular}
\end{table}

\hypertarget{st_augment}{%
\subsection{st\_augment()}\label{st_augment}}

This function augments the original dataframe with the estimated means
and standard errors from the \CRANpkg{mgcv} model in a similar manner to
how the \CRANpkg{broom} (Robinson, Hayes, and Couch 2023) package operates. The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the new random effects columns are positioned immediately before it. The estimates which \texttt{st\_augment()} will extract are random effects (which are called in \CRANpkg{mgcv} with \texttt{bs=\textquotesingle{}re\textquotesingle{}}) and ICAR components (\texttt{bs=\textquotesingle{}mrf\textquotesingle{}}).

Consider the model structure described in the code below using \CRANpkg{mgcv} syntax. In this model \(y\) is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a random intercept and slope for the covariate at a \emph{region} level, and an ICAR varying intercept and slope at a lower \emph{sub-region} level.

\begin{verbatim}
mgcv::gam(
  y ~ covariate +                   # fixed intercept and effect for covariate  
    s(region, bs = "re") +                  # random intercept at level region
    s(region, covariate, bs = "re") +          # random slopes at level region
    s(sub-region, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k) +                    # ICAR varying intercept at level sub-region
    s(sub-region, by = covariate, 
      bs = 'mrf',
      xt = list(nb = data$nb), 
      k = k),           # ICAR varying slope for covariate at level sub-region
  data = data, 
  method = "REML")
\end{verbatim}

When labelling the new columns which are augmented to the original dataframe from such a model, \texttt{st\_augment()} follows the formula syntax of the \CRANpkg{lme4} (Bates et al. 2015) package, where the \emph{pipe symbol} (\texttt{\textbar{}}) indicates ``\emph{grouped by}''. Table \ref{tab:staugtab-latex} shows how the augmented columns in this scenario would be labelled. Each column name begins with either \texttt{random.effect.} or \texttt{mrf.smooth.} as appropriate. An additional column is also added for the standard error of each estimate. These columns are named as above but with \texttt{se.} prepended (e.g.~\texttt{se.random.effect.region}).

\begin{table}

\caption{\label{tab:staugtab-latex}The naming procedure for augmented columns from different mgcv structures.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l|l}
\hline
mgcv syntax & column name\\
\hline
s(region, bs = 're') & random.effect.region\\
\hline
s(region, covariate, bs = 're') & random.effect.covariate|region\\
\hline
s(sub-region, bs = 'mrf', xt = list(nb = data\$nb) & mrf.smooth.sub-region\\
\hline
s(sub-region, by = covariate, bs = 'mrf', xt = list(nb = data\$nb) & mrf.smooth.covariate|sub-region\\
\hline
\end{tabular}
\end{table}

\hypertarget{st_quickmap_preds}{%
\subsection{st\_quickmap\_preds()}\label{st_quickmap_preds}}

These estimates can then be quickly mapped. As it is possible to include more than 1 varying component, the output of this function is a list of plots. They can be viewed individually by indexing, or all at one using the \texttt{plotlist=} argument from the \texttt{ggarrange()} function which is part of the \CRANpkg{ggpubr} (Kassambara 2023) package. We will see examples of this in the following examples. The maps are automatically titled and subtitled according to the type of effect. For example, the map showing \texttt{random.effect.region} will have ``\emph{region}'' as its title and ``\emph{random.effect}'' as its subtitle.

\hypertarget{example-1-indonesia}{%
\section{Example 1: Indonesia}\label{example-1-indonesia}}

Modelling earthquakes in Indonesia seems like a good example to
demonstrate this package. Firstly, Indonesia is composed of many
islands. Secondly, earthquake activity is known to be associated with
the presence of faults which exist below sea level and thus do not
respect land boundaries. Therefore it is reasonable to expect similar
behaviour in nearby provinces regardless of whether or not they are
contiguous. We aim to model the intensity of earthquake activity by
province across Indonesia, controlling for proximity to faults.

\hypertarget{data}{%
\subsection{Data}\label{data}}

The data for this section have been taken from the \href{https://earthquake.usgs.gov/earthquakes/search/}{USGS earthquake
catalogue API}. The
datasets and an explanation are available at
\url{https://github.com/horankev/quake_data} and record all earthquakes in
and close to Indonesia from 1985-2023. Below is a map of Indonesia, with
other neighbouring or bordering countries filled in grey. The many local
faults which lie within 300km of the shore are shown as green lines
in Figure \ref{fig:fault-buffers}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/fault-buffers-1}

}

\caption{Indonesia faults. Surrounded by a 10 kilometre buffer. }\label{fig:fault-buffers}
\end{figure}

To get an interpretable measure of the concentration of faults in any area
they are transformed from linestrings to polygons by setting a buffer of
10km around them, as shown below. Now both our faults and the sizes of provinces are in units of kilometres
squared. This means we can generate a unitless metric of what proportion of any administrative unit
is covered by these buffered faults. This measure across provinces is shown in Figure
\ref{fig:fault-conc}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/fault-conc-1}

}

\caption{Indonesia fault concentration. Square kilometre of buffered fault per square kilometre of province area. }\label{fig:fault-conc}
\end{figure}

Earthquake activity will be seen as the total number of earthquakes with
their epicentre in each province per unit area of that province. We have
restricted counts to earthquakes of magnitude \textgreater5.5 which is the point
at which they are often labelled as damaging.

The occurrences of these earthquakes are shown in Figure \ref{fig:quake-occur}, their total per province in Figure \ref{fig:quake-totals}, and finally, their count per square kilometre can be seen in Figure \ref{fig:quake-conc}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/quake-occur-1}

}

\caption{Earthquakes in Indonesia, 1985-2023. Categorised as medium, large or extra-large. }\label{fig:quake-occur}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/quake-totals-1}

}

\caption{Earthquake count in Indonesia, 1985-2023, mag > 5.5: count by province. }\label{fig:quake-totals}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/quake-conc-1}

}

\caption{Earthquake incidence in Indonesia, 1985-2023, mag > 5.5: count per square kilometre by province. }\label{fig:quake-conc}
\end{figure}

\hypertarget{model}{%
\subsection{Model}\label{model}}

As this is count data, we will model it as a poisson distribution with
\(\lambda\) as the mean count per province. For \(i = 1,...,n\) provinces,
the dependent variable in this model is
\begin{equation}
y_i = \text{earthquake incidence}_i = \frac{\text{earthquake count in province}_i} {\text{province area}_i}, 
\label{eq:eq1}
\end{equation}
while the explanatory variable is
\begin{equation}
x_i = \text{fault intensity}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
\label{eq:eq2}
\end{equation}
Firstly, when excluding the incidence and just modelling counts, where
\(y_i = \text{earthquake count in province}_i\), the poisson model is of
the following form:
\begin{equation}
y_i | \lambda_i \sim \text{Pois}(\lambda_i)
\label{eq:eq3}
\end{equation}
with
\begin{equation}
E(y_i | \lambda_i) = \lambda_i.
\label{eq:eq4}
\end{equation}
We model
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i.
\label{eq:eq5}
\end{equation}
where \(\gamma_i\)
is a term with a correlation structure reflecting a province's location
relative to other provinces.

We can describe these relationships by setting up a neighbourhood
structure based on queen contiguity where a pair of provinces are
considered neighbours if they share at least one point of boundary. This can be modelled as a Markov random field to generate an ICAR model with a spatially varying term. Each of these terms will be correlated the others according to the neighbourhood structure we have defined.

The Markov random field here follows a multivariate Gaussian
distribution. \(\gamma_i\) is a vector of province effects having a
distribution with mean \textbf{0} and precision \textbf{P} where

\([\textbf{p}]_{ij} = v_i\) if \(i=j\) and \(v_i\) is the number of adjacent provinces
to province \(i\),

\([\textbf{p}]_{ij} = -1\) if provinces \(i\) and \(j\) are adjacent, and

\([\textbf{p}]_{ij} = 0\) otherwise.

A further constraint that \(\Sigma_j \gamma_j = 0\) is applied so that the
distribution is identifiable.

If we include an offset term (here, area) because we are more interested
in the incidence than in the actual count, then
\begin{equation}
log(\frac{\lambda_i} {\text{area}_i}) = \beta_0 + \beta_1x_i + \gamma_i
\label{eq:eq6}
\end{equation}
which is equivalent to
\begin{equation}
log(\lambda_i) = \beta_0 + \beta_1x_i + \gamma_i + log(\text{area}_i).
\label{eq:eq7}
\end{equation}
We are still modelling \(log(\lambda)\) rather than the incidence, but we
are adding an offset to adjust for differing areas. Modelling \(log(\lambda)\) and adding an offset is equivalent to
modelling incidence, and coefficients can be interpreted that way.

When interpreting the estimated coefficients of the model, it can be
useful to look at it in the following form:
\begin{equation}
\lambda_i = e^{\beta_0 + \beta_1 x_i + \gamma_i} \text{area}_i.
\label{eq:eq8}
\end{equation}

\hypertarget{pre-functions-1}{%
\subsection{Pre-functions}\label{pre-functions-1}}

Such models, however, can not deal with locations which have no
neighbours. In the case of Indonesia, this is quite problematic. It is
composed of many islands. The estimated count of
islands according to Andréfouët, Paul, and Farhan (2022) is 13,558.
While it is not unusual for a country to have a number of often small
offshore islands, Indonesia is entirely composed of (at least portions
of) an archipelago of islands, so many of these islands or groups of
islands are individual provinces in their own right. We might like them
to be associated with their k nearest neighbours. However, a nearest
neighbours approach to Indonesia as a whole might not be desired.
Therefore, they must either be excluded from the study or somehow
brought within the neighbourhood framework.

In this case, we use \texttt{st\_bridges()} for setting up the queen contiguity
structure as usual, but with the additional stipulation that unconnected
units (provinces which are islands or collections of islands) are
considered neighbours to their \emph{k} nearest provinces. For this example,
we have set the value of k to 2. The resulting neighbourhood structure
is shown in Figure \ref{fig:ind-contig1}.

\begin{verbatim}
# join islands to k=2 nearest neighbours
# various arguments exist for altering colours and sizes
# additional ggplot themes and layers can be added

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "darkblue", linksize = 0.8, 
                 pointcol = "red", pointsize = 2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=nearby_countries_df, 
          fill="gray50", linewidth=0.5, colour="black")
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ind-contig1-1}

}

\caption{Neighbourhood structure for Indonesian provinces created by \texttt{st\_bridges()} with \texttt{k=2}.}\label{fig:ind-contig1}
\end{figure}

This neighbourhood structure now has no unconnected provinces so it is
suitable for use in an ICAR model. However, we will

\begin{itemize}
\item
  add some additional contiguities using \texttt{st\_manual\_join\_nb()}
\item
  and remove one using \texttt{st\_manual\_cut\_nb()}.
\end{itemize}

To cater for the possibility that a modeller might not be familiar with the names of the various geographic units but still wishes to make manual alterations to their relationships, we can look at a map (Figure \ref{fig:ind-contig2}) where the nodes are shown by
index number instead of as points (using the argument \texttt{nodes=\textquotesingle{}numeric\textquotesingle{}}). This makes it easy to manually cut
and join neighbours as desired. Furthermore, there is an option to
show concave hulls drawn around each unit (using \texttt{concavehull\ =\ TRUE}). This is also shown in Figure \ref{fig:ind-contig2}. These shapes are not used in the assignment of contiguities but it can be useful to see them in a situation such as Indonesia where many individual provinces are actually multipolygons. Without them, it is not clear whether an island is a province in its own right, or which group of islands together form one province.

\begin{verbatim}
# with 'concavehull = TRUE' and 'nodes = "numeric"'

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "tomato", linksize = 0.5, 
                 nodes = "numeric", 
                 numericcol = "black", numericsize = 6, 
                 concavehull = TRUE, 
                 hullcol = "darkgreen", hullsize = 0.2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank())
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ind-contig2-1}

}

\caption{Neighbourhood structure for Indonesian provinces. Viewed with \texttt{st\_quickmap\_nb()}, using the arguments \texttt{nodes\ =\ \textquotesingle{}numeric\textquotesingle{}} and \texttt{concavehull\ =\ TRUE}.}\label{fig:ind-contig2}
\end{figure}

\begin{verbatim}
# with 'concavehull = TRUE' and 'nodes = "numeric"'

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "tomato", linksize = 0.5, 
                 nodes = "numeric", 
                 numericcol = "black", numericsize = 6, 
                 concavehull = TRUE, 
                 hullcol = "darkgreen", hullsize = 0.2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank())
\end{verbatim}

Having made some manual adjustments to the neighbourhood structure, the
new structure can be seen in Figure \ref{fig:ind-contig3}. Edge effects
have also been mitigated by additionally connecting the two extreme
provinces (1 and 23), which would otherwise have only one neighbour, to
their two next closest neighbours.

\begin{verbatim}
# a series of manual joins and cuts by index number

st_bridges(provinces_df, "province",link_islands_k = 2) |> 
  st_manual_join_nb(1,24) |> 
  st_manual_join_nb(1,30) |> 
  st_manual_join_nb(3,13) |> 
  st_manual_join_nb(13,17) |> 
  st_manual_join_nb(14,25) |> 
  st_manual_join_nb(20,29) |>
  st_manual_join_nb(19,23) |> 
  st_manual_join_nb(16,27) |> 
  st_manual_join_nb(22,23) |> 
  st_manual_join_nb(7,19) |> 
  st_manual_join_nb(7,20) |>
  st_manual_join_nb(19,28) |> 
  st_manual_join_nb(4,18) |> 
  st_manual_join_nb(21,26) |> 
  st_manual_join_nb(22,28) |> 
  st_manual_cut_nb(19,22) |> 
  st_quickmap_nb(fillcol = "antiquewhite1", 
                 bordercol = "black", bordersize = 0.5, 
                 linkcol = "darkblue", linksize = 0.8, 
                 pointcol = "red", pointsize = 2) + 
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=nearby_countries_df, 
          fill="gray50", linewidth=0.5, colour="black") + 
  annotation_scale()
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ind-contig3-1}

}

\caption{Neighbourhood structure for Indonesian provinces, after alterations using \texttt{st\_manual\_join()} and \texttt{st\_manual\_cut()}.}\label{fig:ind-contig3}
\end{figure}

\hypertarget{mgcv-model}{%
\subsection{\texorpdfstring{\texttt{mgcv} model}{mgcv model}}\label{mgcv-model}}

We now create the ICAR model using \CRANpkg{mgcv}. We will be able to simply pass
the contiguity structure we have created as both the data and a column
of it as the neighbourhood structure.

\begin{verbatim}
mod_pois_mrf <- gam(damaging_quakes_total ~ 
                      fault_concentration +
                      s(province, bs='mrf', xt=list(nb=prep_data$nb), k=22) +
                      offset(log(area_province)),
                    data=prep_data, method="REML",family = "poisson")
\end{verbatim}

The adjusted R-squared is \textbf{0.957} and deviance explained is \textbf{88.2\%}. The estimated coefficient associated with fault concentration is
-3.2355. As this is positive, it is greater than 1 when exponentiated:
\[
\text{exp}(0.7531) = 0.04
\]
and this is the expected multiplicative association between a unit
change in fault concentration and incidence of earthquakes in a
province.

Controlling for this estimated mean elevated risk of earthquakes due to
concentration of faults, what is the additional risk level of
earthquakes in a province? This can be seen as a measure of the activity
level of faults locally and it is spatially smoothed by the
autoregressive process. It is mapped as the ICAR component of the model
below:

\hypertarget{post-functions-1}{%
\subsection{Post-functions}\label{post-functions-1}}

Use \texttt{st\_augment()} to get estimates from \CRANpkg{mgcv} model in format
similar to that used by the \CRANpkg{broom} package. For instance, we see from the code below that the original dataframe is now augmented with columns for ``mrf.smooth.province'' and ``mrf.smooth.province'' which shows the estimate and standard errors for the \(\gamma_i\) component. Note that this is how we would expect them to be named, based on the previous discussion surrounding Table \ref{tab:staugtab-latex}.

\begin{verbatim}
mod_pois_mrf |> 
  st_augment(prep_data) |> 
  names()
\end{verbatim}

\begin{verbatim}
#>  [1] "S"                       "M"                      
#>  [3] "L"                       "XL"                     
#>  [5] "quake_total"             "quake_density"          
#>  [7] "damaging_quakes_total"   "damaging_quakes_density"
#>  [9] "area_fault_within"       "area_province"          
#> [11] "fault_concentration"     "nb"                     
#> [13] "mrf.smooth.province"     "se.mrf.smooth.province" 
#> [15] "province"                "province_id"            
#> [17] "geometry"
\end{verbatim}

Use \texttt{st\_augment()} and \texttt{st\_quickmap\_preds} in the following code to get a quick visualisation of
the estimates for \(\gamma_i\), as shown in Figure \ref{fig:ind-mrf1}. Again, note that the title and subtitle of the image are as previously discussed.

\begin{verbatim}
# st_quickmap_preds() outputs a list of ggplots

plot_mrf <- mod_pois_mrf |> 
  st_augment(prep_data) |>
  st_quickmap_preds(scale_low = "darkgreen",
                    scale_mid = "ivory", 
                    scale_high = "darkred", 
                    scale_midpoint = 0)

# in this case, there is only one plot in the list
# so we call it by index

plot_mrf[[1]] +
  coord_sf(datum=NA) +
  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
                                        linewidth=1.5),
        axis.text = element_blank()) +
  geom_sf(data=provinces_df, fill=NA, colour="black", linewidth=0.5) + 
  geom_sf(data=nearby_countries_df, fill="gray50", colour="black", 
          linewidth=0.5) + 
  labs(fill="relative\nincidence") +
  annotation_scale() +
  coord_sf(datum=NA) + 
  theme(legend.position = c(0.92,0.77),
        legend.box.background = element_rect(colour = "black", linewidth = 1),
        legend.title = element_text())
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ind-mrf1-1}

}

\caption{Estimates of \(\gamma_i\) shown as a map using \texttt{st\_quickmap\_preds()}.}\label{fig:ind-mrf1}
\end{figure}

If we wish to apply the inverse link function (the exponential function in the case of this poisson model) to map these values to a more interpretable scale, this will not be generated by the function \texttt{st\_quickmap\_preds()}. Instead, we must use the augmented dataframe which is produced by \texttt{st\_augment()} and create the appropriate extra column with the usual \CRANpkg{tidyverse} \texttt{mutate()} function. This allows us to produce the map in Figure
\ref{fig:ind-mrf2}. As these coefficients are multiplicatively related
to the earthquake incidence, values below 1 imply an earthquake
incidence which is lower than expected.

The provinces with the 3 most elevated incidences are labelled in red
We can see that, controlling for the added risk of proximity to
faults, the province of Aceh has 7.9 times the expected
incidence, or number of major earthquakes per square kilometre. The two
lowest-scoring provinces, labelled in green, have essentially no incidence of
earthquake epicentres, controlling for what their proximity to faults alone would
suggest.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ind-mrf2-1}

}

\caption{Map showing estimates of \(\text{exp}(\gamma_i).\) This is produced by adding an additional column to the dataframe produced by \texttt{st\_augment()}.}\label{fig:ind-mrf2}
\end{figure}

\hypertarget{workflow}{%
\subsection{Workflow}\label{workflow}}

In this example, we have gone through a number of stages carefully, making changes to contiguities that we deemed appropriate as we went. However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps. A rough and ready model (and visual output) can be generated with \texttt{sfislands} using nothing more than three or four lines of code, such as the following:

\begin{verbatim}
# set up neighbourhood structure

prep_data <- st_bridges(provinces_df, "province")

# define model

mod <- gam(quake_mlxl_total ~ 
                  fault_concentration +
                  s(province, bs='mrf', xt=list(nb=prep_data$nb), k=22) +
                  offset(log(area_province)),
        data=prep_data, method="REML",family = "poisson")

# augment tidy estimates

tidy_ests <- st_augment(mod, prep_data)

# visualise them

st_quickmap_preds(tidy_ests)
\end{verbatim}

\hypertarget{example-2-london}{%
\section{Example 2: London}\label{example-2-london}}

The next example looks only at using the \emph{pre-functions}, but in a situation where the presence of islands is not the issue. Consider the wards and boroughs of London (sourced from the Greater London Authority's \href{https://data.london.gov.uk/}{London Datastore}). In Figure \ref{fig:lon-contig1} the \texttt{st\_bridges()}
function is applied to them to construct a queen contiguity
neighbourhood structure. Because there are no disjoint units (or
``islands''), this will be the same as using \texttt{st\_contiguity()} from
\CRANpkg{sfdep}. The \texttt{st\_quickmap\_nb()} function gives an immediate
visual representation of the structure. This can be supplemented by
adding a layer showing the course of the river Thames.

\begin{verbatim}
# same as sfdep:st_contiguity() as there are no islands
# an extra layer for the river Thames

st_bridges(london, "NAME") |> 
  st_quickmap_nb() + 
  geom_sf(data=thames, colour="blue", linewidth=1.5) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/lon-contig1-1}

}

\caption{Wards and boroughs of Greater London. Queen contiguity. }\label{fig:lon-contig1}
\end{figure}

When the study area has a river running through it there can be problems
with constructing neighbourhood structures. Depending on how the
geometries are defined, the presence of a river can either mean that no
units on either side of its banks are considered neighbours or, at the
other extreme, the river is essentially ignored and all units on
opposing banks are considered neighbours.

Depending on the presence of river crossings, two areas which are
physically quite close might be very distinct. Somebody living on the
banks of a river might be more likely to go about their life primarily
on their side of the river, despite the short distance as the crow flies
of facilities on the other side.

\texttt{sfislands} makes it easier to take account of this situation. We can look
at just those boroughs which are on either side of the river Thames (Figure
\ref{fig:lon-contig2})

\begin{verbatim}
# which boroughs are alongside the river

riverside <- thames |> st_intersects(london) |> unlist() |> unique()

# only map these boroughs

st_bridges(london[riverside,],"NAME") |> 
  st_quickmap_nb(linksize = 0.5) +
  geom_sf(data=thames, colour="blue", linewidth=1.5) + 
  annotation_scale(location="br") +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/lon-contig2-1}

}

\caption{Riverside boroughs of Greater London. Queen contiguity disregarding river. }\label{fig:lon-contig2}
\end{figure}

In order to take account of actual connectivity, we can add a layer
showing the road and pedestrian bridges or tunnels. These were scraped from a Wikipedia article describing the various crossings of the river Thames. In Figure \ref{fig:lon-crossings}, we
have also drawn a 1 kilometre buffer around them. Until we get to the
eastern portion of the map, all units on opposing sides of the river are
within 1 kilometre of a crossing. This is not the case for the extreme
eastern units so we might not want to consider these units neighbours if
the river flows between them.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/lon-crossings-1}

}

\caption{Riverside boroughs of Greater London. Road and pedestrian crossing and tunnels surrounded by 1 kilometre buffer shaded green. }\label{fig:lon-crossings}
\end{figure}

In order the identify the changes we wish to make, we use the
\texttt{nodes\ =\ "numeric"} argument in \texttt{st\_quickmap\_nb()}. Now we can identify
each unit by its position in the contiguity structure. Here we have
shaded in pink the units which are not within 1 kilometre of a river
crossing (see Figure \ref{fig:lon-contig3}).

\begin{verbatim}
# with 'nodes = "numeric"'

st_bridges(london[riverside,],"NAME") |> 
  st_quickmap_nb(nodes = "numeric", numericsize = 4, linksize = 0.5) +
  geom_sf(data=no_touch_buffer, fill="pink", alpha=0.3) + 
  geom_sf(data=crossings_roadped |> st_buffer(1000), 
          fill="darkgreen", alpha=0.3) +
  geom_sf(data=thames, colour="blue", linewidth=1.5) + 
  geom_sf(data=crossings_roadped, size=1, colour="yellow") + 
  annotation_scale(location="br") +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/lon-contig3-1}

}

\caption{Riverside boroughs of Greater London. Index number for each borough shown at centroid. Boroughs which are not within 1 kilometre of a crossing are shaded pink. }\label{fig:lon-contig3}
\end{figure}

This allows us to easily cut the ties across the river for these units
manually by using the function \texttt{st\_manual\_cut\_nb()}. The results of this editing procedure are shown in Figure \ref{fig:lon-contig4}.
While we are using the index of the units in this example, the function also
accepts names as arguments which may be more convenient in some circumstances. Having made these adjustments,
\texttt{st\_quickmap\_nb()} now shows a connectivity structure (Figure \ref{fig:lon-contig4}) which reflects the influence of the
presence or absence of river crossings.

This example shows that the pre-functions of \texttt{sfislands} have uses for situations which do not involve islands. They can be used to apply domain knowledge to easily design the most appropriate neighbourhood structure.

\begin{verbatim}
# manually cut the links where there is no crossing

st_bridges(london[riverside,],"NAME") |> 
  st_manual_cut_nb(18,17) |> 
  st_manual_cut_nb(19,17) |> 
  st_manual_cut_nb(19,20) |> 
  st_manual_cut_nb(20,21) |> 
  st_manual_cut_nb(21,22) |> 
  st_manual_cut_nb(22,48) |> 
  st_manual_cut_nb(47,48) |> 
  st_manual_cut_nb(45,46) |> 
  st_manual_cut_nb(45,47) |>  
  st_manual_cut_nb(39,65) |> 
  st_manual_cut_nb(1,2) |> 
  st_quickmap_nb(bordercol = "black", bordersize = 0.5, linksize = 0.5) +
  geom_sf(data=no_touch_buffer, fill="pink", alpha=0.3) + 
  geom_sf(data=crossings_roadped |> st_buffer(1000), 
          fill="darkgreen", alpha=0.3) +
  geom_sf(data=thames, colour="blue", linewidth=1.5) +
  annotation_scale(location="br") +
  coord_sf(datum=NA) + 
  theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
                                        linewidth=1.5))
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/lon-contig4-1}

}

\caption{Riverside boroughs of Greater London. Contiguities across the river have been cut for the pink boroughs. }\label{fig:lon-contig4}
\end{figure}

\hypertarget{example-3-liverpool}{%
\section{Example 3: Liverpool}\label{example-3-liverpool}}

In this final example, there are no islands or river issues. We saw in the Indonesia example how \texttt{sfislands} can extract estimates from the \texttt{mgcv} argument \texttt{bs="mrf"}. It can also deal with random effects as defined by \texttt{bs="re"} in \texttt{mgcv} models. Here, we
will construct a hierarchical model in this way which also features an ICAR component. We will
compare two different types of contiguity which could be applied:

\begin{itemize}
\tightlist
\item
  \texttt{st\_bridges()} from \texttt{sfislands}, and
\item
  \texttt{st\_knn()} from \CRANpkg{sfdep}.
\end{itemize}

We will show how their outcomes can be quickly mapped to give an immediate sense of the differences.

\begin{table}

\caption{\label{tab:livadmin-latex}Three nested levels of administrative divisions in Liverpool.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{7cm}}
\hline
code & area type & description\\
\hline
OA & output areas & lowest level of geographical area for census statistics, usually containing 100 - 625 persons\\
\hline
LSOA & lower layer super output areas & usually 4 or 5 OAs\\
\hline
MSOA & middle layer super output areas & usually 4 or 5 LSOAs\\
\hline
\end{tabular}
\end{table}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/liv-nested-1}

}

\caption{Nested levels of administrative divisions in Liverpool. }\label{fig:liv-nested}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.7\linewidth]{article_files/figure-latex/liv-vars-1}

}

\caption{Unemployment and limiting long-term illness rates in Liverpool, with extent of Mersey estuary in light blue to the south west. }\label{fig:liv-vars}
\end{figure}

\begin{verbatim}
# magnify a section

from <- c(xmin = -2.97, xmax = -2.95, ymin = 53.38, ymax = 53.39)
to <- c(-3.01,-2.96,53.32,53.35)

st_bridges(liverpool, "oa_cd") |> 
  st_quickmap_nb(pointsize = 0.1,
                 linksize = 0.1,
                 fillcol = "gray90") +  
  geom_sf(data=mersey, fill="lightblue", colour="gray30", 
          linewidth=0.2, alpha=0.5) + 
  geom_magnify(data=liverpool, 
               from = from, to = to, 
               corners = 0.1, shadow = TRUE, linewidth = 0.6) + 
  theme(panel.background = element_rect(fill = "#F7FFF2",
                                        colour = "black", linewidth=1))
\end{verbatim}

\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{article_files/figure-latex/liv-contigs-1}

}

\caption{First order queen contiguity of output areas (OAs) in Liverpool with magnified inset. }\label{fig:liv-contigs}
\end{figure}

This example is based on data from the \href{https://www.nomisweb.co.uk/home/census2001.asp}{Office of National Statistics} as presented by Rowe and Arribas-Bel (2024) in their demonstration of techniques for hierarchical spatial regression. They use the different levels of nested administrative boundaries in Liverpool for their demonstration. These are described in Table \ref{tab:livadmin-latex}, and are mapped separately in Figure \ref{fig:liv-nested}.

Now suppose we wish to model unemployment as a function of limited long-term
illness, taking into account this hierarchical structure. Limiting long-term illness (LLTI) refers to a health problem or disability which limits a person's day-to-day activities (Barnett 2001). We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at MSOA and LSOA levels. The
distribution of these variables across the OAs of Liverpool are shown in Figure
\ref{fig:liv-vars}.

If \(y_{ijk}\) and \(x_{ijk}\) are respectively the levels of unemployment and LLTI in each of \(i\) OAs within \(j\) LSOAs, which in turn are within \(k\) MSOAs, then
\begin{equation}
y_{ijk} = \beta_{0jk} + \beta_{1jk}x_{ijk} + \gamma_{ijk} + \epsilon_{ijk}
\label{eq:eq9}
\end{equation}
\begin{equation}
\beta_{0jk} = \beta_{0k} + u_{0jk}
\label{eq:eq10}
\end{equation}
\begin{equation}
\beta_{1jk} = \beta_{1k} + u_{1jk}
\label{eq:eq11}
\end{equation}
\begin{equation}
\beta_{0k} = \beta_0 + w_{0k}
\label{eq:eq12}
\end{equation}
\begin{equation}
\beta_{1k} = \beta_1 + w_{1k}
\label{eq:eq13}
\end{equation}
\begin{equation}
\epsilon_{ijk} \sim N(0,\sigma_{\epsilon}^2)
\label{eq:eq14}
\end{equation}
\begin{equation}
u_{jk} \sim N(0,\sigma_{u}^2)
\label{eq:eq15}
\end{equation}
\begin{equation}
 w_{k} \sim N(0,\sigma_{w}^2)
\label{eq:eq16}
\end{equation}
where the \(\gamma_{ijk}\)s comprise a vector of ICAR components for each OA, similar to the provinces in the earlier Indonesia model. The vector has a multivariate normal distribution with mean 0 and precision related to the contiguity structure as before.

There are no islands or discontiguities in this geometry of Liverpool so
using \texttt{st\_bridges()} will give the same structure as would
\texttt{sfdep::st\_contiguity()} (see Figure \ref{fig:liv-contigs}).

The hierarchical model with no mrf component produces estimates as shown
in Figure \ref{fig:liv-est1}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/liv-est1-1}

}

\caption{Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas as levels. }\label{fig:liv-est1}
\end{figure}

If we instead use contiguity as the basis of an additional spatial term,
we get estimates shown in Figure \ref{fig:liv-est2}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/liv-est2-1}

}

\caption{Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas are the top two levels respectively (MSOA \& LSOA). Output areas (OA) as lowest level Markov random field smooth using first-degree queen contiguity from \texttt{st\_bridges()}.}\label{fig:liv-est2}
\end{figure}

If instead we consider units to be neighbours if they lie within a
distance band of 700 metres, we get estimates shown in Figure
\ref{fig:liv-est3}.

A comparison of their AICs (see Table \ref{tab:aic-latex}) suggests that, in this scenario, queen contiguity is more appropriate than the distance band condition.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/liv-est3-1}

}

\caption{Random intercept and slope estimates from hierarchical model. Middle and lower layer super output areas are the top two levels respectively (MSOA \& LSOA). Output areas (OA) as lowest level Markov random field smooth using \texttt{sfdep::st\_dist\_band()} with radius of 700 metres.}\label{fig:liv-est3}
\end{figure}

\begin{table}

\caption{\label{tab:aic-latex}Comparison of models by AIC.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{l|r|r}
\hline
model & AIC & relative AIC\\
\hline
hierachical model, ICAR based on st\_bridges() & -5126.948 & -1.000\\
\hline
hierachical model, ICAR based on sfdep::st\_dist\_band() & -5107.790 & -0.996\\
\hline
hierachical model, no ICAR & -5082.684 & -0.991\\
\hline
\end{tabular}
\end{table}

\hypertarget{summary}{%
\section{Summary}\label{summary}}

These examples have shown the varying scenarios in which \texttt{sfislands} can be useful. It aims to contribute to spatial modelling by making an awkward area less awkward. Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included. It also provides helper functions to use these structures in spatial regression models built with \texttt{mgcv} which streamline the human effort necessary to examine the estimates.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

This publication has emanated from research conducted with the financial support of Science Foundation Ireland under Grant number 18/CRT/6049.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Andrfout2022}{}}%
Andréfouët, Serge, Mégane Paul, and A. Riza Farhan. 2022. {``Indonesia's 13558 Islands: A New Census from Space and a First Step Towards a One Map for Small Islands Policy.''} \emph{Marine Policy} 135 (January): 104848. \url{https://doi.org/10.1016/j.marpol.2021.104848}.

\leavevmode\vadjust pre{\hypertarget{ref-r-inla}{}}%
Bakka, Haakon, Håvard Rue, Geir-Arne Fuglstad, Andrea Riebler, David Bolin, Elias Krainski, Daniel Simpson, and Finn Lindgren. 2018. {``Spatial Modelling with r-INLA: A Review.''} arXiv. \url{https://doi.org/10.48550/ARXIV.1802.06350}.

\leavevmode\vadjust pre{\hypertarget{ref-Barnett2001}{}}%
Barnett, S. 2001. {``A Multilevel Analysis of the Effects of Rurality and Social Deprivation on Premature Limiting Long Term Illness.''} \emph{Journal of Epidemiology \&Amp; Community Health} 55 (1): 44--51. \url{https://doi.org/10.1136/jech.55.1.44}.

\leavevmode\vadjust pre{\hypertarget{ref-lme4}{}}%
Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015. {``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of Statistical Software} 67 (1): 1--48. \url{https://doi.org/10.18637/jss.v067.i01}.

\leavevmode\vadjust pre{\hypertarget{ref-spdep}{}}%
Bivand, Roger, Andrew Bernat, Marilia Carvalho, Yongwan Chun, Carsten Dormann, Stéphane Dray, Rein Halbersma, et al. 2005. {``The Spdep Package.''} \emph{Comprehensive R Archive Network, Version}, 05--83.

\leavevmode\vadjust pre{\hypertarget{ref-BrizRedn2021}{}}%
Briz-Redón, Álvaro, Adina Iftimi, Juan Francisco Correcher, Jose De Andrés, Manuel Lozano, and Carolina Romero-García. 2021. {``A Comparison of Multiple Neighborhood Matrix Specifications for Spatio-Temporal Model Fitting: A Case Study on COVID-19 Data.''} \emph{Stochastic Environmental Research and Risk Assessment} 36 (1): 271--82. \url{https://doi.org/10.1007/s00477-021-02077-y}.

\leavevmode\vadjust pre{\hypertarget{ref-brms}{}}%
Bürkner, Paul-Christian. 2017. {``{brms}: An {R} Package for {Bayesian} Multilevel Models Using {Stan}.''} \emph{Journal of Statistical Software} 80 (1): 1--28. \url{https://doi.org/10.18637/jss.v080.i01}.

\leavevmode\vadjust pre{\hypertarget{ref-Duncan2017}{}}%
Duncan, Earl W., Nicole M. White, and Kerrie Mengersen. 2017. {``Spatial Smoothing in Bayesian Models: A Comparison of Weights Matrix Specifications and Their Impact on Inference.''} \emph{International Journal of Health Geographics} 16 (1). \url{https://doi.org/10.1186/s12942-017-0120-x}.

\leavevmode\vadjust pre{\hypertarget{ref-Earnest2007}{}}%
Earnest, Arul, Geoff Morgan, Kerrie Mengersen, Louise Ryan, Richard Summerhayes, and John Beard. 2007. {``Evaluating the Effect of Neighbourhood Weight Matrices on Smoothing Properties of Conditional Autoregressive (CAR) Models.''} \emph{International Journal of Health Geographics} 6 (1): 54. \url{https://doi.org/10.1186/1476-072x-6-54}.

\leavevmode\vadjust pre{\hypertarget{ref-ggpubr}{}}%
Kassambara, Alboukadel. 2023. \emph{Ggpubr: 'Ggplot2' Based Publication Ready Plots}. \url{https://rpkgs.datanovia.com/ggpubr/}.

\leavevmode\vadjust pre{\hypertarget{ref-sfdep}{}}%
Parry, Josiah, and Dexter H Locke. 2024. \emph{Sfdep: Spatial Dependence for Simple Features}. \url{https://sfdep.josiahparry.com}.

\leavevmode\vadjust pre{\hypertarget{ref-sf}{}}%
Pebesma, Edzer. 2018. {``{Simple Features for R: Standardized Support for Spatial Vector Data}.''} \emph{{The R Journal}} 10 (1): 439--46. \url{https://doi.org/10.32614/RJ-2018-009}.

\leavevmode\vadjust pre{\hypertarget{ref-broom}{}}%
Robinson, David, Alex Hayes, and Simon Couch. 2023. {``Broom: Convert Statistical Objects into Tidy Tibbles.''} \url{https://CRAN.R-project.org/package=broom}.

\leavevmode\vadjust pre{\hypertarget{ref-liverpool-notes}{}}%
Rowe, Francisco, and Dani Arribas-Bel. 2024. {``{University of Liverpool, Spatial Modeling for Data Scientists, ENVS453, Course repository}.''} \url{https://gdsl-ul.github.io/san/}.

\leavevmode\vadjust pre{\hypertarget{ref-rstan}{}}%
Stan Development Team. 2020. {``{RStan}: The {R} Interface to {Stan}.''} \url{http://mc-stan.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-tobler}{}}%
Tobler, W. R. 1970. {``A Computer Movie Simulating Urban Growth in the Detroit Region.''} \emph{Economic Geography} 46 (sup1): 234--40. \url{https://doi.org/10.2307/143141}.

\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. {``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\leavevmode\vadjust pre{\hypertarget{ref-mgcv}{}}%
Wood, S. N. 2011. {``Fast Stable Restricted Maximum Likelihood and Marginal Likelihood Estimation of Semiparametric Generalized Linear Models''} 73: 3--36. \url{https://CRAN.R-project.org/web/packages/mgcv/index.html}.

\end{CSLReferences}

\bibliography{RJreferences.bib}

\address{%
Kevin Horan\\
Hamilton Institute, Maynooth University\\%
Maynooth\\ Co.~Kildare, Ireland\\
%
\url{https://github.com/horankev}\\%
\textit{ORCiD: \href{https://orcid.org/0009-0003-9378-0084}{0009-0003-9378-0084}}\\%
\href{mailto:kevin.horan.2021@mumail.ie}{\nolinkurl{kevin.horan.2021@mumail.ie}}%
}

\address{%
Katarina Domijan\\
Department of Mathematics and Statistics, Maynooth University\\%
Maynooth\\ Co.~Kildare, Ireland\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-4268-2236}{0000-0002-4268-2236}}\\%
\href{mailto:katarina.domijan@mu.ie}{\nolinkurl{katarina.domijan@mu.ie}}%
}

\address{%
Chris Brunsdon\\
National Centre for Geocomputation, Maynooth University\\%
Maynooth\\ Co.~Kildare, Ireland\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0003-4254-1780}{0000-0003-4254-1780}}\\%
\href{mailto:Christopher.Brunsdon@mu.ie}{\nolinkurl{Christopher.Brunsdon@mu.ie}}%
}

### 2024-03-03 demo_paper.Rmd
initial commit

---
title: "sfislands: An R Package for Customizable Tooltips in Interactive Graphics"
date: "2022-11-09"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  - name: Kevin Horan
    affiliation: 
    - Hamilton Institute, Maynooth University
    address:
    - Maynooth
    - Co. Kildare, Ireland
    url: https://github.com/horankev
    orcid: 0009-0003-9378-0084
    email:  kevin.horan.2021@mumail.ie
  - name: Katarina Domijan
    affiliation:
    - Department of Mathematics and Statistics, Maynooth University
    address:
    - Maynooth
    - Co. Kildare, Ireland
    email: katarina.domijan@mu.ie
    orcid: 0000-0002-4268-2236
  - name: Chris Brunsdon
    affiliation:
    - National Centre for Geocomputation, Maynooth University
    address:
    - Maynooth
    - Co. Kildare, Ireland
    email: Christopher.Brunsdon@mu.ie
    orcid: 0000-0003-4254-1780
type: package
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(ggplot2)
library(palmerpenguins)
library(kableExtra)
```

# Introduction

Interactive data graphics provides plots that allow users to interact them. One of the most basic types of interaction is through tooltips, where users are provided additional information about elements in the plot by moving the cursor over the plot.

This paper will first review some R packages on interactive graphics and their tooltip implementations. A new package \CRANpkg{ToOoOlTiPs} that provides customized tooltips for plot, is introduced. Some example plots will then be given to showcase how these tooltips help users to better read the graphics.

# Background

Some packages on interactive graphics include \CRANpkg{plotly} [@plotly] that interfaces with Javascript for web-based interactive graphics, \CRANpkg{crosstalk} [@crosstalk] that specializes cross-linking elements across individual graphics.  The recent R Journal paper \CRANpkg{tsibbletalk} [@RJ-2021-050] provides a good example of including interactive graphics into an article for the journal. It has both a set of linked plots, and also an animated gif example, illustrating linking between time series plots and feature summaries.

# Customizing tooltip design with \pkg{ToOoOlTiPs}

\pkg{ToOoOlTiPs} is a packages for customizing tooltips in interactive graphics, it features these possibilities.

# A gallery of tooltips examples

The \CRANpkg{palmerpenguins} data [@palmerpenguins]  features three penguin species which has a lovely illustration by Alison Horst in Figure \@ref(fig:penguins-alison).

```{r penguins-alison, out.width = "100%", out.height = "30%", fig.cap = "Artwork by \\@allison\\_horst", fig.alt="A picture of three different penguins with their species: Chinstrap, Gentoo, and Adelie. "}
knitr::include_graphics("figures/penguins.png")
```

Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:penguins-tab-interactive)', '\\@ref(tab:penguins-tab-static)'))` prints at the first few rows of the `penguins` data:

```{r penguins-tab-interactive, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(head(penguins), format = "html", caption = "A basic table")
```

```{r penguins-tab-static, eval = knitr::is_latex_output()}
knitr::kable(head(penguins), format = "latex", caption = "A basic table") %>% 
  kableExtra::kable_styling(font_size = 7)
```

Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:penguins-plotly)', '\\@ref(fig:penguins-ggplot)'))` shows an `r ifelse(knitr::is_html_output(), 'interactive', '')` plot of the penguins data, made using the `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\CRANpkg{plotly}', '\\CRANpkg{ggplot2}'))` package.

```{r penguins-plotly, echo = TRUE, fig.height = 5, fig.cap="A basic interactive plot made with the plotly package on palmer penguin data. Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis. When hovering on a point, a tooltip will show the exact value of the bill depth and length for that point, along with the species name.", include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.alt = "A scatterplot of bill length against bill depth, both measured in millimetre. The three species are shown in different colours and loosely forms three clusters. Adelie has small bill length and large bill depth, Gentoo has small bill depth but large bill length, and Chinstrap has relatively large bill depth and bill length."}
p <- penguins %>% 
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
             color = species)) + 
  geom_point()
ggplotly(p)
```

```{r penguins-ggplot, echo = TRUE, fig.height = 5, fig.cap="A basic non-interactive plot made with the ggplot2 package on palmer penguin data. Three species of penguins are plotted with bill depth on the x-axis and bill length on the y-axis. Visit the online article to access the interactive version made with the plotly package.", include=knitr::is_latex_output(), eval=knitr::is_latex_output()}
penguins %>% 
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, 
             color = species)) + 
  geom_point()
```

# Summary

We have displayed various tooltips that are available in the package \pkg{ToOoOlTiPs}.

### 2024-03-03 article.R

@@ -1,0 +1,1178 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit article.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
+
+# install.packages("devtools")
+devtools::install_github("horankev/sfislands")
+
+# required libraries
+library(tidyverse)
+library(kableExtra)
+library(sf)
+library(sfdep)
+library(sfislands)
+library(mgcv)
+library(here)
+library(ggsflabel)
+library(ggspatial)
+library(ggpubr)
+library(AER) # to test for overdispersion
+library(ggmagnify)
+
+# To scrape data
+library(rvest)
+library(httr)
+library(polite)
+library(janitor)
+
+# for geom_sf_label_repel consistency
+set.seed(123)
+
+# set consistent theme
+theme_set(
+  theme_bw() +
+    theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", linewidth=1.5), 
+          axis.title = element_blank(),
+          axis.ticks = element_blank())
+)
+
+
+
+## -----------------------------------------------------------------------------
+# load data
+
+# datasets are all available at https://github.com/horankev/quake_data 
+# or https://github.com/horankev/london_liverpool_data
+# alongside details on precisely how they were sourced etc
+
+# indonesia
+quakes_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/quakes_df.rds")))
+provinces_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/provinces_df.rds"))) |> 
+  rename(damaging_quakes_total = quake_mlxl_total,
+         damaging_quakes_density = quake_mlxl_density)
+counties_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/counties_df.rds")))
+faults_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/faults_df.rds")))
+nearby_countries_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/nearby_countries_df.rds")))
[…]

### 2024-03-03 article.Rmd
explain saved file use

@@
 
+# these models take a while to run
+# so i am using a saved version of their output
+# the files are large so are not on github
+# uncomment the model and run it to reproduce
+
 prepliv <- st_bridges(liverpool, "oa_cd")
@@
 ```{r}
+
+# these models take a while to run
+# so i am using a saved version of their output
+# the files are large so are not on github
+# uncomment the model and run it to reproduce
+
 # liverpool_hsamodel <- gam(unemployment ~ lt_illness +
@@
 
+# these models take a while to run
+# so i am using a saved version of their output
+# the files are large so are not on github
+# uncomment the model and run it to reproduce
+
 # liverpool_hsamodel_dist_band <- gam(unemployment ~ lt_illness +

### 2024-03-15 article.R
revisions to article

@@ -1,1178 +1,0 @@
-# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
-# Please edit article.Rmd to modify this file
-
-## ----setup, include=FALSE-----------------------------------------------------
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
-
-# install.packages("devtools")
-devtools::install_github("horankev/sfislands")
-
-# required libraries
-library(tidyverse)
-library(kableExtra)
-library(sf)
-library(sfdep)
-library(sfislands)
-library(mgcv)
-library(here)
-library(ggsflabel)
-library(ggspatial)
-library(ggpubr)
-library(AER) # to test for overdispersion
-library(ggmagnify)
-
-# To scrape data
-library(rvest)
-library(httr)
-library(polite)
-library(janitor)
-
-# for geom_sf_label_repel consistency
-set.seed(123)
-
-# set consistent theme
-theme_set(
-  theme_bw() +
-    theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", linewidth=1.5), 
-          axis.title = element_blank(),
-          axis.ticks = element_blank())
-)
-
-
-
-## -----------------------------------------------------------------------------
-# load data
-
-# datasets are all available at https://github.com/horankev/quake_data 
-# or https://github.com/horankev/london_liverpool_data
-# alongside details on precisely how they were sourced etc
-
-# indonesia
-quakes_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/quakes_df.rds")))
-provinces_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/provinces_df.rds"))) |> 
-  rename(damaging_quakes_total = quake_mlxl_total,
-         damaging_quakes_density = quake_mlxl_density)
-counties_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/counties_df.rds")))
-faults_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/faults_df.rds")))
-nearby_countries_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/nearby_countries_df.rds")))
[…]

### 2024-03-15 article.Rmd
revisions to article

@@
 creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
-structures and their compatibility with the \CRANpkg{tidyverse}  [@tidyverse], the \CRANpkg{sfdep}  [@sf] package [@sfdep] has presented the same
+structures and their compatibility with the \CRANpkg{tidyverse}  [@tidyverse], the \CRANpkg{sfdep} package [@sfdep] has presented the same
 functionality in addition to extra features in a tidy structure based on
@@
 services might be satisfactory to meet our hypothesis of the required
-degree of connectivity to be considered neighbours.
+degree of connectivity to be considered neighbours. Again, such information will not be found in a set of polygons.
 
@@
 convenient and open manner. It allows us to set up a structure, quickly
-map it, and then examine it. The structure can then be edited and the process re-iterated until we have described a
+map it, and then examine whether or not it is consistent with our hypothesis of relationships between units. The structure can then be edited and the process re-iterated until we have described a
 spatial relationship structure with which we are satisfied.
@@
 The above can be considered as the *pre-functions* of the package. A
-further set of feature of `sfislands`, which we refer to as *post-functions*,
-are for use with the \CRANpkg{mgcv} package [@mgcv]. It is straightforward to use
-`sfislands` neighbourhood structures in models with \CRANpkg{mgcv} but it can be quite awkward
-to extract the estimates.
-These *post-functions* conveniently provide the estimates in tidy format of random effects and intrinsic conditional autoregressive (ICAR) components from such models and give the ability to quickly map them for visualisation purposes.
+second category of features, which we refer to as *post-functions*,
+are for use with the \CRANpkg{mgcv} package [@mgcv].
+Having fit a model with \CRANpkg{mgcv}, the process of extracting estimates for certain types of effects can be somewhat awkward.
+These *post-functions* augment the original dataframe with these estimates and their standard errors in tidy format.
+They also allow for quick visualisation of the output in map form.
 
@@
 Indonesia. It shows a scenario where all of the functions are used, from
-setting up contiguities to modelling and examining the estimates of the
+setting up contiguities, to modelling and examining the estimates of the
 model.
@@
 
-It greatly speeds up the workflow of fitting areal spatial models.
-
-* **pre-functions** for setting up contiguities:
-
-  + it addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data.
-`sfislands` will include a contiguity for all units,
-
-  + it gives tools to immediately visualise this structure as a map,
-
-  + these maps are created using \CRANpkg{ggplot2}  [@ggplot2], which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax,
-
-  + as the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence,
-
-  + connections made by the package which are not contiguities can be accessed to ensure openness in the process.
-
-* For modelling:
-
-  + these neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms}. [@brms], \pkg{r-inla}  [@r-inla] and more.
-
-* **post-functions** for `mgcv` models:
-
-  + it simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models),
-
-  + these effects can be quickly visualised in \CRANpkg{ggplot2} maps.
+It greatly speeds up each stage of a proposed workflow for fitting areal spatial models.
+
+### Step 1: *Pre-functions* for setting up neighbourhood structure
+
+  1.
+It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data.
+`sfislands` will include a contiguity for all units.
+
+  2. It gives tools to immediately visualise this structure as a map.
+
+  3. These maps are created using \CRANpkg{ggplot2}  [@ggplot2], which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax.
+
+  4. As the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence.
+
+  5. Connections which have been induced by a function from the package but are not contiguities can be accessed to ensure openness in the process.
+
+### Step 2: Modelling
+
+  These neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms}  [@brms], \pkg{r-inla}  [@r-inla] and more.
+
+### Step 3: *Post-functions* for `mgcv` models
+
+  1. It simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models).
+
+  2. These effects can be quickly visualised as \CRANpkg{ggplot2} maps.
 
@@
 The first group of functions, shown in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:prefunc-html)', '\\@ref(tab:prefunc-latex)'))`, deals with the creation
-of a contiguity structure in the presence of discontiguities. The
+of a neighbourhood structure in the presence of discontiguities. The
 resultant structure can be quickly mapped to check if it is
@@
 index number. By an iterative process of changes and examination of a
-quick map, a satisfactory structure can be decided upon.
-
-Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:prefunc-html)', '\\@ref(tab:prefunc-latex)'))` shows these functions and their purpose.
+quickly-generated guide map, a satisfactory structure can be decided upon.
 
@@
 knitr::kable(prefunc, format = "latex", caption = "Pre-functions: setting up a neighbourhood structure.") |> 
+  row_spec(0, bold = TRUE) |> 
   kableExtra::kable_styling(font_size = 9) |> 
@@
 
-They deal with the creation
-of a contiguity structure in the presence of discontiguities. The
-resultant structure can be quickly mapped to check if it is
-satisfactory. Connections can be manually added or removed by name or
-index number. By an iterative process of changes and examination of a
-quick map, a satisfactory structure can be decided upon.
-Going through each function in more detail:
-
-Use the rectangles shown in Figure \@ref(fig:rects1) as a simple demonstration. Rectangles 1-3
-are contiguous while 4 and 5 are islands.
+We will now go through each function in more detail using the set of rectangles shown in Figure \@ref(fig:rects1) for demonstration purposes.
+Rectangles 1, 2 and 3 are contiguous while 4 and 5 can be viewed as "islands".
 
@@
 
-This function requires as arguments an \CRANpkg{sf} dataframe and the name of a column of unique identifiers, preferably names, of each spatial unit.
-The k-nearest neighbours for connecting islands can then be specified.
+This function requires at least two arguments: an \CRANpkg{sf} dataframe and, from that, the name of one column of unique identifiers, ideally names, of each spatial unit.
+It creates a neighbourhood structure where non-island units are joined by first-order queen contiguity, while island units are joined to their k-nearest neighbours.
 The output is a *named* neighbourhood structure in either list or matrix form as desired, which can be either a standalone object or included as an additional column in the original \CRANpkg{sf} dataframe.
@@
 
-In the following examples, we have chosen to ignore islands with the argument `remove_islands = TRUE` and to return a list and matrix structure respectively by specifying this in the `nb_structure` argument and choosing `add_to_dataframe = FALSE`:
+In the first two examples of using `st_bridges()`, we have chosen to ignore islands with the argument `remove_islands = TRUE` and to return a list and matrix structure respectively by specifying this in the `nb_structure` argument and choosing `add_to_dataframe = FALSE`:
 
@@
 
-In these examples, we choose to join islands to their 1 nearest
+Alternatively, in the following examples, we choose to join islands to their 1 nearest
 neighbour, which is the default setting, and to return the output as a column called "nb" in the original
-\CRANpkg{sf} dataframe (this is the default setting):
+\CRANpkg{sf} dataframe (`add_to_dataframe = "TRUE"` is the default setting):
 
@@
 These structures can serve as the input to models in \CRANpkg{brms},
-\pkg{r-inla}, \pkg{rstan}  [@rstan] or \CRANpkg{mgcv}. Rather than having a separate neighbours
-column, it is included as a named list or matrix in the original \CRANpkg{sf}
-dataframe, in the spirit of the \CRANpkg{sfdep} package.
+\pkg{r-inla}, \pkg{rstan}  [@rstan] or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
+object, it is included in the original \CRANpkg{sf} dataframe as a named list or matrix, in the spirit of the \CRANpkg{sfdep} package.
 
@@
 
-This function will tell us openly what connections have been made which
-are not based on contiguity. It gives both the name and index number of each pair of added connections.
+This function will show us transparently what connections have been made which
+are not based on contiguity. It gives both the name and index number of each pair of added connections. In this example, two pairs have been added.
 
 ```{r, echo=TRUE}
+# show summary of non-contiguous connections in a dataframe
+
 st_bridges(rectangles, 
@@
 ```{r, rects4, fig.width=2, fig.height=1, echo=TRUE, fig.cap = "With an additional connection between 3 and 4. "}
+# add an extra connection using numeric index
+
 st_bridges(rectangles, "name", 
@@
 ```{r, rects5, fig.width=2, fig.height=1, echo=TRUE, fig.cap = "With an additional connection between 1 and 2. "}
+# remove an existing connection using unit name, not index
+
 st_bridges(rectangles, "name", 
@@
 
-The use of these structures is particularly common in CAR-type models.
+The use of these neighbourhood structures is particularly associated with CAR-type models.
 These are often implemented in a Bayesian framework using
-\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. The pre-functions will output the
+\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
 neighbourhood structure in the desired format for use in any of these
 frameworks. A convenient frequentist alternative is to use the
-\CRANpkg{mgcv} package. It has the functionality to create such models
+\CRANpkg{mgcv} package which requires a named list of neighbours. It has the functionality to create such models
 using `bs="mrf"`. It also has the ability to combine these with a
@@
 Bayesian structures mentioned above can be extracted in the same way as
-any other component of the model, it can be a bit awkward to get the
+any other component of the model, it can be somewhat awkward to get the
 estimates from \CRANpkg{mgcv} model. `sfislands` has two post-functions to conveniently extract and visualise these.
@@
 knitr::kable(postfunc, format = "latex", caption = "Post-functions: tidy estimates from mgcv.") |> 
+  row_spec(0, bold = TRUE) |> 
   kableExtra::kable_styling(font_size = 9, full_width = FALSE)
@@
 This function augments the original dataframe with the estimated means
-and standard errors from the \CRANpkg{mgcv} model in a similar manner to
-how the \CRANpkg{broom}  [@broom] package operates.
-The `geometry` column, as per convention, remains as the last column of the augmented dataframe, while the new random effects columns are positioned immediately before it.
-The estimates which `st_augment()` will extract are random effects (which are called in \CRANpkg{mgcv} with `bs='re'`) and ICAR components (`bs='mrf'`).
+and standard errors of the spatially varying predictions from a fitted \CRANpkg{mgcv} model in a similar manner to
+how the \CRANpkg{broom} package  [@broom] operates.
+The `geometry` column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it.
+^[In a similar way, `st_augment()` can also be used to append the random effects from `lme4` and `nlme` models to an `sf` dataframe, which can then be easily mapped using `st_quickmap_preds()`.] The spatially varying predictions which `st_augment()` extracts from an \CRANpkg{mgcv} model are
+
+*   random effects (which are called in \CRANpkg{mgcv} with `bs='re'`), and 
+*   ICAR components (`bs='mrf'`).
 
 Consider the model structure described in the code below using \CRANpkg{mgcv} syntax.
-In this model $y$ is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a random intercept and slope for the covariate at a *region* level, and an ICAR varying intercept and slope at a lower *sub-region* level.
+In this model $y$ is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a *region* level, and a set of ICAR varying intercepts and slopes at a lower *sub-region* level.
 
 ```{r, eval=FALSE, echo=TRUE}
+# creating an mgcv model
+
 mgcv::gam(
@@
 
-When labelling the new columns which are augmented to the original dataframe from such a model, `st_augment()` follows the formula syntax of the \CRANpkg{lme4}  [@lme4] package, where the *pipe symbol* (`|`) indicates "*grouped by*".
-Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:staugtab-html)', '\\@ref(tab:staugtab-latex)'))` shows how the augmented columns in this scenario would be labelled.
+When labelling the new prediction columns which are augmented to the original dataframe from such a model, `st_augment()` follows the formula syntax of the \CRANpkg{lme4} package  [@lme4], where the pipe symbol (`|`) indicates "*grouped by*".
+Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:staugtab-html)', '\\@ref(tab:staugtab-latex)'))` shows how the augmented columns in this scenario would be named.
 Each column name begins with either `random.effect.` or `mrf.smooth.` as appropriate.
-An additional column is also added for the standard error of each estimate.
+An additional column is also added for the standard error of each estimate, as calculated by \CRANpkg{mgcv}.
 These columns are named as above but with `se.` prepended (e.g.
@@
 knitr::kable(st_aug_tab_latex, format = "latex", caption = "The naming procedure for augmented columns from different mgcv structures.") |> 
-  kableExtra::kable_styling(font_size = 9, full_width = FALSE)
+  row_spec(0, bold = TRUE) |> 
+  kableExtra::kable_styling(font_size = 9, full_width = FALSE, )
 ```
@@
 These estimates can then be quickly mapped.
-As it is possible to include more than 1 varying component, the output of this function is a list of plots.
-They can be viewed individually by indexing, or all at one using the `plotlist=` argument from the `ggarrange()` function which is part of the \CRANpkg{ggpubr}  [@ggpubr] package.
-We will see examples of this in the following examples.
-The maps are automatically titled and subtitled according to the type of effect.
-For example, the map showing `random.effect.region` will have "*region*" as its title and "*random.effect*" as its subtitle.
+As it is possible to include more than 1 spatially varying component, the output of this function is a list of plots.
+They can be viewed individually by indexing, or all at one using, for example, the `plotlist` argument from the `ggarrange()` function which is part of the \CRANpkg{ggpubr}  [@ggpubr] package.
+We will see this function in practice in the following examples.
+The maps which it generates are automatically titled and subtitled according to the type of effect.
+For example, the map showing predictions for `random.effect.region` will have "*region*" as its title and "*random.effect*" as its subtitle.
 
@@
 
-Modelling earthquakes in Indonesia seems like a good example to
+Modelling earthquakes in Indonesia serves as a good example to
 demonstrate this package. Firstly, Indonesia is composed of many
@@
 behaviour in nearby provinces regardless of whether or not they are
-contiguous. We aim to model the intensity of earthquake activity by
+contiguous. We aim to model the incidence, or count per unit area, of earthquake activity by
 province across Indonesia, controlling for proximity to faults.
@@
 
-The data for this section have been taken from the [USGS earthquake
-catalogue API](https://earthquake.usgs.gov/earthquakes/search/). The
-datasets and an explanation are available at
-<https://github.com/horankev/quake_data> and record all earthquakes in
-and close to Indonesia from 1985-2023. Below is a map of Indonesia, with
-other neighbouring or bordering countries filled in grey. The many local
-faults which lie within 300km of the shore are shown as green lines
-in Figure \@ref(fig:fault-buffers).
+The data for this section have been downloaded from the National Earthquake Information Center, [USGS earthquake
+catalogue](https://earthquake.usgs.gov/earthquakes/search/). The
+datasets with accompanying explanations are available at
+<https://github.com/horankev/quake_data>. They capture all recorded earthquakes in
+and close to Indonesia from the beginning of January 1985 to the end of December 2023.
+Figure \@ref(fig:fault-buffers) shows a map of Indonesia, divided into 33 provinces, with other neighbouring or bordering countries filled in grey.
+The many local
+faults which lie within 300km of the shore are shown in yellow with green outlines.
 
@@
 
-To get an interpretable measure of the concentration of faults in any area
-they are transformed from linestrings to polygons by setting a buffer of
-10km around them, as shown below. Now both our faults and the sizes of provinces are in units of kilometres
+To get an interpretable measure of the concentration of faults in any area,
+these faults are transformed from linestrings to polygons by setting a buffer of
+10km around them, which explains their green outline. Now both our faults and the sizes of provinces are in units of kilometres
 squared. This means we can generate a unitless metric of what proportion of any administrative unit
@@
 
-Earthquake activity will be seen as the total number of earthquakes with
-their epicentre in each province per unit area of that province. We have
-restricted counts to earthquakes of magnitude \>5.5 which is the point
-at which they are often labelled as damaging.
-
-The occurrences of these earthquakes are shown in Figure \@ref(fig:quake-occur), their total per province in Figure \@ref(fig:quake-totals), and finally, their count per square kilometre can be seen in Figure \@ref(fig:quake-conc).
-
-```{r, quake-occur, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Earthquakes in Indonesia, 1985-2023. Categorised as medium, large or extra-large. "}
-
+Earthquake incidence per province has been calculated as the total number of earthquakes with
+an epicentre within that province per unit area. We have
+restricted counts to earthquakes \>5.5 on the moment magnitude scale, which is the point
+at which they are often labelled as potentially damaging.
+
+The occurrences of these earthquakes are shown in Figure \@ref(fig:quake-occur), their total per province in Figure \@ref(fig:quake-totals), and finally, their incidence or count per square kilometre can be seen in Figure \@ref(fig:quake-conc).
+
+```{r, quake-occur, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Earthquakes in Indonesia of magnitude > 5.5, 1985-2023. Categorised by magnitude as medium, large or extra-large. "}
+
+quakes_df_temp <- quakes_df |> 
+  mutate(magfact = case_when(magfact == "M" ~ "M: 5.5 - 6.1",
+                             magfact == "L" ~ "L: 6.1 - 7",
+                             magfact == "XL" ~ "XL: 7+")) |> 
+  mutate(magfact = factor(magfact,
+                          levels = c("M: 5.5 - 6.1",
+                                     "L: 6.1 - 7",
+                                     "XL: 7+")))
 ggplot() + 
@@
   geom_sf(data=provinces_df, fill="antiquewhite", colour="black", linewidth=0.5) + 
-  geom_sf(data=quakes_df |> filter(magfact != "S") |> filter(province != "undersea"), 
+  geom_sf(data=quakes_df_temp |> filter(magfact != "S") |> filter(province != "undersea"), 
           aes(fill=magfact, size=magfact), shape=21, colour="black") +
   scale_fill_manual(values = c("white","khaki","tomato4")) +
-  scale_size_manual(values = c(3,5,7)) + 
-  labs(fill="category",
-       size="category") + 
+  scale_size_manual(values = c(3.5,5.5,7.5)) + 
+  labs(fill="magnitude",
+       size="magnitude") + 
   annotation_scale() +
@@
 \begin{equation}
-y_i = \text{earthquake incidence}_i = \frac{\text{earthquake count in province}_i} {\text{province area}_i}, 
+y_i = \text{earthquake count}_i 
 (\#eq:eq1)
@@
 \begin{equation}
-x_i = \text{fault intensity}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
+x_i = \text{fault concentration}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
 (\#eq:eq2)
@@
 This can be modelled as a Markov random field to generate an ICAR model with a spatially varying term.
-Each of these terms will be correlated the others according to the neighbourhood structure we have defined.
+Each of these terms will be correlated with the others according to the neighbourhood structure we have defined.
 
@@
 
-If we include an offset term (here, area) because we are more interested
-in the incidence than in the actual count, then
+We now include an offset term (here, area) because we are more interested
+in modelling the incidence than in the actual count, such that
 \begin{equation}
@@
 
+Having described the type of model we wish to implement, we now show how `sfislands` can be used to streamline the process.
+
 ## Pre-functions
@@
 of) an archipelago of islands, so many of these islands or groups of
-islands are individual provinces in their own right. We might like them
-to be associated with their k nearest neighbours. However, a nearest
-neighbours approach to Indonesia as a whole might not be desired.
-Therefore, they must either be excluded from the study or somehow
-brought within the neighbourhood framework.
+islands are individual provinces in their own right.
+We might like to hypothesise that just because a province is a disconnected island, this should not mean that it is independent of other nearby provinces in terms of earthquake incidence.
+A standard first-order queen contiguity structure would mean the exclusion of disconnected units entirely from the model.
+An alternative strategy of assigning neighbour status based on a distance metric would overcome this, but the threshold size of distance necessary for such a structure might be inappropriately large for the non-islands provinces.
+We would like to use a compromise between these two strategies.
 
@@
 This neighbourhood structure now has no unconnected provinces so it is
-suitable for use in an ICAR model. However, we will
+suitable for use in an ICAR model.
+However, if we are not entirely happy with this structure because of some domain knowledge about the inter-relationships between certain island provinces, we might wish to
 
@@
 This is also shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig2html', 'ind-contig2')`).
-These shapes are not used in the assignment of contiguities but it can be useful to see them in a situation such as Indonesia where many individual provinces are actually multipolygons.
+These shapes are not used in the assignment of contiguities but it can be useful to see them in a situation such as Indonesia where many individual provinces are actually multipolygons of more than one island.
 Without them, it is not clear whether an island is a province in its own right, or which group of islands together form one province.
@@
 new structure can be seen in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig3html', 'ind-contig3')`). Edge effects
-have also been mitigated by additionally connecting the two extreme
-provinces (1 and 23), which would otherwise have only one neighbour, to
+have also been mitigated by imposing additional connections on the two extreme
+provinces (1 and 23), which would otherwise have only one neighbour, so that they now also include
 their two next closest neighbours.
@@
 
-We now create the ICAR model using \CRANpkg{mgcv}. We will be able to simply pass
-the contiguity structure we have created as both the data and a column
-of it as the neighbourhood structure.
+We now create the ICAR model using, in this case, the \CRANpkg{mgcv} package.
+We will be able to use the output of `st_bridges`, which we have named `prep_data`, as both the data source for the model and the neighbourhood structure (by specifying the column `nb` which contains the neighbourhood list).
 
@@
 
-The adjusted R-squared is **0.957** and deviance explained is **88.2%**. The estimated coefficient associated with fault concentration is
--3.2355. As this is positive, it is greater than 1 when exponentiated:
-$$
-\text{exp}(0.7531) = 0.04
-$$
-and this is the expected multiplicative association between a unit
-change in fault concentration and incidence of earthquakes in a
-province.
-
-Controlling for this estimated mean elevated risk of earthquakes due to
-concentration of faults, what is the additional risk level of
+We can see from the summary below that the adjusted R-squared is **0.957** and deviance explained is **88.2%**.
+
+```{r}
+summary(mod_pois_mrf)
+```
+
+Controlling for the concentration of faults, what is the additional risk level of
 earthquakes in a province? This can be seen as a measure of the activity
 level of faults locally and it is spatially smoothed by the
-autoregressive process. It is mapped as the ICAR component of the model
-below:
+autoregressive process.
+It is represented in the model summary by the component `s(province)`.
+However, the extraction of individual predictions for this component for each province from the `mgcv` model requires a number of steps.
+We know demonstrate how these are streamlined into a single function by the `sfislands` package.
 
@@
 
-Use `st_augment()` to get estimates from \CRANpkg{mgcv} model in format
-similar to that used by the \CRANpkg{broom} package.
-For instance, we see from the code below that the original dataframe is now augmented with columns for "mrf.smooth.province" and "mrf.smooth.province" which shows the estimate and standard errors for the $\gamma_i$ component.
+The function `st_augment()` allow us to add the spatially varying predictions from the model as new columns to the original dataframe in a process
+similar to that of the \CRANpkg{broom} package.
+For instance, we see from the code below that the original dataframe is now augmented with columns called "mrf.smooth.province" and "se.mrf.smooth.province" which show the predictions for the $\gamma_i$ component and their standard errors.
 Note that this is how we would expect them to be named, based on the previous discussion surrounding Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:staugtab-html)', '\\@ref(tab:staugtab-latex)'))`.
+They are positioned immediately before the final `geometry` column of the `sf` dataframe.
 
 ```{r, echo=TRUE}
+# column names of augmented dataframe
+
 mod_pois_mrf |> 
@@
 
-Use `st_augment()` and `st_quickmap_preds` in the following code to get a quick visualisation of
-the estimates for $\gamma_i$, as shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-mrf1html', 'ind-mrf1')`).
+This output can now be piped into the `st_quickmap_preds()` function to get a quick visualisation of
+these estimates for $\gamma_i$ on a map, as shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-mrf1html', 'ind-mrf1')`).
 Again, note that the title and subtitle of the image are as previously discussed.
@@
 The provinces with the 3 most elevated incidences are labelled in red
-We can see that, controlling for the added risk of proximity to
+We can see that, controlling for the effects of proximity to
 faults, the province of Aceh has 7.9 times the expected
@@
 lowest-scoring provinces, labelled in green, have essentially no incidence of
-earthquake epicentres, controlling for what their proximity to faults alone would
+earthquake epicentres within their boundaries, controlling for what their proximity to faults alone would
 suggest.
@@
 
-## Workflow
+## Workflow summary
 
@@
 However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps.
-A rough and ready model (and visual output) can be generated with `sfislands` using nothing more than three or four lines of code, such as the following:
+A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with `sfislands` using nothing more than three or four lines of code, such as the following:
 
 ```{r, eval=FALSE, echo=TRUE}
-# set up neighbourhood structure
+# workflow:
+
+# 1. set up neighbourhood structure
 
@@
 
-# define model
+# 2. define model
 
@@
 
-# augment tidy estimates
+# 3. augment tidy estimates
 
@@
 
-# visualise them
+# 4. visualise them
 
@@
 
-The next example looks only at using the *pre-functions*, but in a situation where the presence of islands is not the issue.
+The next example looks only at using the *pre-functions* of `sfislands`, but in a situation where the presence of actual *islands* is not the problem we seek to deal with.
 Consider the wards and boroughs of London (sourced from the Greater London Authority's [London Datastore](https://data.london.gov.uk/)).
@@
 \CRANpkg{sfdep}. The `st_quickmap_nb()` function gives an immediate
-visual representation of the structure. This can be supplemented by
-adding a layer showing the course of the river Thames.
+visual representation of the structure.
+^[`st_quickmap_nb()` can also be used to visualise any contiguity structure created by `spdep` or `sfdep` as long as that structure is included in an `sf` dataframe as a column named `nb`.] This can be supplemented by
+adding a layer showing the course of the river Thames which is also visible in Figure \@ref(fig:lon-contig1).
 
@@
 
-When the study area has a river running through it there can be problems
-with constructing neighbourhood structures. Depending on how the
-geometries are defined, the presence of a river can either mean that no
-units on either side of its banks are considered neighbours or, at the
-other extreme, the river is essentially ignored and all units on
-opposing banks are considered neighbours.
-
-Depending on the presence of river crossings, two areas which are
-physically quite close might be very distinct. Somebody living on the
-banks of a river might be more likely to go about their life primarily
-on their side of the river, despite the short distance as the crow flies
-of facilities on the other side.
-
-`sfislands` makes it easier to take account of this situation. We can look
-at just those boroughs which are on either side of the river Thames (Figure
-\@ref(fig:lon-contig2))
+When a study area has a river running through it, problems can arise with constructing appropriate neighbourhood structures. Depending on how the
+geometries are defined, the presence of a river can cause problems in two ways.
+In one situation, the river could be expressed as a polygon in its own right meaning that it severs any potential contiguity between units on either side of its banks.
+In this situation, no spatial units will be neighbours with the units directly across the river from them.
+At the other extreme, if the river is not included as a geometry (as is the case here) all units on opposing banks are considered neighbours.
+
+Depending on the presence of river crossings, two areas which are physically quite close but on opposing banks might be very distinct.
+If there is no means of crossing the river within a reasonable distance, somebody living on the banks of a river might be more likely to go about their life primarily on their side of the river, despite the short distance as the crow flies of facilities on the other side.
+This could be relevant in terms of, say, modelling of house prices where we might want to incorporate these issues into a neighbourhood structure.
+
+`sfislands` provides convenient functions for this sort of situation.
+Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames (Figure \@ref(fig:lon-contig2)).
 
@@
 In order to take account of actual connectivity, we can add a layer
-showing the road and pedestrian bridges or tunnels. These were scraped from a Wikipedia article describing the various crossings of the river Thames. In Figure \@ref(fig:lon-crossings), we
-have also drawn a 1 kilometre buffer around them. Until we get to the
-eastern portion of the map, all units on opposing sides of the river are
-within 1 kilometre of a crossing. This is not the case for the extreme
-eastern units so we might not want to consider these units neighbours if
-the river flows between them.
+showing the road and pedestrian bridges or tunnels.
+Details of these were sourced @wikicrossings, which describes the various crossings of the river Thames.
+In Figure \@ref(fig:lon-crossings), we have also drawn a 1 kilometre buffer around each crossing.
+This was chosen as an arbitrary measure of what might be considered a "reasonable" distance within which to consider opposing banks as being connected.
+The vast majority of units on opposing banks have access to a river crossing within this threshold and thus should be considered as neighbours.
+Only the extreme eastern units and one to the south west should not have a connection across the river.
 
@@
 This allows us to easily cut the ties across the river for these units
-manually by using the function `st_manual_cut_nb()`. The results of this editing procedure are shown in Figure \@ref(fig:lon-contig4).
-While we are using the index of the units in this example, the function also
-accepts names as arguments which may be more convenient in some circumstances. Having made these adjustments,
-`st_quickmap_nb()` now shows a connectivity structure (Figure \@ref(fig:lon-contig4)) which reflects the influence of the
-presence or absence of river crossings.
+manually by using the function `st_manual_cut_nb()`. ^[While we are using the index of the units in this example, the function also
+accepts names as arguments which may be more convenient in some circumstances.] Having made these adjustments,
+`st_quickmap_nb()` now shows a connectivity structure (Figure \@ref(fig:lon-contig4)) which reflects how we have hypothesised that influence should extend across the river in the presence or absence of crossings.
 
@@
 
-st_bridges(london[riverside,],"NAME") |> 
+st_bridges(london[riverside,], "NAME") |> 
   st_manual_cut_nb(18,17) |> 
@@
   st_quickmap_nb(bordercol = "black", bordersize = 0.5, linksize = 0.5) +
-  geom_sf(data=no_touch_buffer, fill="pink", alpha=0.3) + 
+  geom_sf(data=no_touch_buffer, fill = "pink", alpha = 0.3) + 
   geom_sf(data=crossings_roadped |> st_buffer(1000), 
-          fill="darkgreen", alpha=0.3) +
-  geom_sf(data=thames, colour="blue", linewidth=1.5) +
-  annotation_scale(location="br") +
+          fill= "darkgreen", alpha = 0.3) +
+  geom_sf(data=thames, colour = "blue", linewidth = 1.5) +
+  annotation_scale(location = "br") +
   coord_sf(datum=NA) + 
   theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
-                                        linewidth=1.5))
+                                        linewidth = 1.5))
 ```
@@
 
-In this final example, there are no islands or river issues.
+In this final example, there are no islands and no river issues.
 We saw in the Indonesia example how `sfislands` can extract estimates from the `mgcv` argument `bs="mrf"`.
 It can also deal with random effects as defined by `bs="re"` in `mgcv` models.
+This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally related components.
 Here, we
-will construct a hierarchical model in this way which also features an ICAR component. We will
-compare two different types of contiguity which could be applied:
-
-* `st_bridges()` from `sfislands`, and 
-* `st_knn()` from \CRANpkg{sfdep}.
-
-We will show how their outcomes can be quickly mapped to give an immediate sense of the differences.
-
-```{r}
-livadmin <- tribble(
-  ~code, ~`area type`, ~description,
-  "OA", "output areas", "lowest level of geographical area for census statistics, usually containing 100 - 625 persons",
-  "LSOA", "lower layer super output areas", "usually 4 or 5 OAs",
-  "MSOA", "middle layer super output areas", "usually 4 or 5 LSOAs"
-)
-
-```
-
-```{r livadmin-html, eval = knitr::is_html_output(), layout = "center"}
-knitr::kable(livadmin, format = "html", caption = "Three nested levels of administrative divisions in Liverpool.") |> 
-  kable_styling(full_width = TRUE)
-```
-
-```{r livadmin-latex, eval = knitr::is_latex_output()}
-knitr::kable(livadmin, format = "latex", caption = "Three nested levels of administrative divisions in Liverpool.") |> 
-  kableExtra::kable_styling(font_size = 8) |> 
-  column_spec(3, width = "7cm")
-```
-
-```{r, liv-nested, fig.width=10, fig.height=5, out.width="100%", fig.cap = "Nested levels of administrative divisions in Liverpool. "}
-
-ggarrange(
-  ggplot() + 
-    geom_sf(data=msoa_sf, fill="gray90", colour="black") +  
-    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
-    geom_sf(data=liverpool |> summarise(), colour="black", fill=NA) + 
-    labs(title = "level 1: MSOA polygons") + 
-  coord_sf(datum=NA) + 
-  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
-                                        linewidth=1.5),
-        axis.text = element_blank()),
-
-  ggplot() + 
-    geom_sf(data=lsoa_sf, fill="gray90", colour="gray40") +  
-    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
-    geom_sf(data=liverpool |> summarise(), colour="black", fill=NA) + 
-    labs(title = "level 2: LSOA polygons") + 
-  coord_sf(datum=NA) + 
-  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
-                                        linewidth=1.5),
-        axis.text = element_blank()),
-
-  ggplot() + 
-    geom_sf(data=liverpool |> summarise(), colour="black", fill="white") + 
-    geom_sf(data=liverpool |> st_centroid(), 
-            fill="gray90", colour="gray50", size=0.1) +  
-    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
-    labs(title = "level 3: OA (shown as centroids)") + 
-  coord_sf(datum=NA) + 
-  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
-                                        linewidth=1.5),
-        axis.text = element_blank()),
-
-  ncol=3
-)
-
-```
-
-```{r, liv-vars, fig.width=6, fig.height=4, out.width="70%", fig.cap = "Unemployment and limiting long-term illness rates in Liverpool, with extent of Mersey estuary in light blue to the south west.
-"}
-
-ggarrange(
-  ggplot() + 
-    geom_sf(data=liverpool, aes(fill=unemployment, colour=unemployment)) + 
-    geom_sf(data=liverpool |> summarise(), 
-            fill=NA, colour="gray30", linewidth=0.2) +  
-    geom_sf(data=mersey, fill="lightblue", colour="gray30", linewidth=0.2) + 
-    scale_fill_distiller(palette = "PuBu", direction = 1) + 
-    scale_colour_distiller(palette = "PuBu", direction = 1) + 
-    labs(fill="unemployment",
-         colour="unemployment") + 
-    coord_sf(datum=NA) + 
-    annotation_scale(location = "bl", width_hint = 0.4) +
-    annotation_north_arrow(location = "bl", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.2, "in"),
-        style = north_arrow_fancy_orienteering) +
-    theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
-                                          linewidth=0.6), 
-          axis.title = element_blank(),
-          axis.ticks = element_blank(),
-          legend.text = element_text(size=8),
-          legend.title = element_text(size=8)),
-  ggplot() + 
-    geom_sf(data=liverpool, aes(fill=lt_illness, colour=lt_illness)) + 
-    geom_sf(data=liverpool |> summarise(), 
-            fill=NA, colour="gray30", linewidth=0.2) +  
-    geom_sf(data=mersey, fill="lightblue", colour="gray30", linewidth=0.2) + 
-    scale_fill_distiller(palette = "RdYlBu", direction = -1) + 
-    scale_colour_distiller(palette = "RdYlBu", direction = -1) + 
-    labs(fill="limiting long-term\nillness\n(z-score)",
-         colour="limiting long-term\nillness\n(z-score)") + 
-    coord_sf(datum=NA) +
-    annotation_scale(location = "bl", width_hint = 0.4) +
-    annotation_north_arrow(location = "bl", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.2, "in"),
-        style = north_arrow_fancy_orienteering) +
-    theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
-                                          linewidth=0.6), 
-          axis.title = element_blank(),
-          axis.ticks = element_blank(),
-          legend.text = element_text(size=8),
-          legend.title = element_text(size=8)),
-
-  ncol=2, legend = "bottom"
-)
-
-```
-
-```{r, liv-contigs, fig.width=5, fig.height=5, out.width="60%", echo=TRUE, fig.cap = "First order queen contiguity of output areas (OAs) in Liverpool with magnified inset. "}
-# magnify a section
-
-from <- c(xmin = -2.97, xmax = -2.95, ymin = 53.38, ymax = 53.39)
-to <- c(-3.01,-2.96,53.32,53.35)
-
-st_bridges(liverpool, "oa_cd") |> 
-  st_quickmap_nb(pointsize = 0.1,
-                 linksize = 0.1,
-                 fillcol = "gray90") +  
-  geom_sf(data=mersey, fill="lightblue", colour="gray30", 
-          linewidth=0.2, alpha=0.5) + 
-  geom_magnify(data=liverpool, 
-               from = from, to = to, 
-               corners = 0.1, shadow = TRUE, linewidth = 0.6) + 
-  theme(panel.background = element_rect(fill = "#F7FFF2",
-                                        colour = "black", linewidth=1))
-
-```
-
-```{r}
-
-# these models take a while to run
-# so i am using a saved version of their output
-# the files are large so are not on github
-# uncomment the model and run it to reproduce
-
-prepliv <- st_bridges(liverpool, "oa_cd")
-
-# liverpool_model2 <- gam(unemployment ~ lt_illness + 
-#                          s(msoa_cd, bs="re") + 
-#                          s(msoa_cd, lt_illness, bs="re") +
-#                          s(lsoa_cd, bs="re") + 
-#                          s(lsoa_cd, lt_illness, bs="re"), # ICAR constituency ICAR varying coefficients
-#                        data=prepliv, method="REML")
-
-# saveRDS(liverpool_model2, here("data","liverpool_model2.rds"))
-
-liverpool_hmodel <- readRDS(here("data","liverpool_model2.rds"))
-
-# summary(liverpool_hmodel)
-
-```
-
-```{r}
-
-# these models take a while to run
-# so i am using a saved version of their output
-# the files are large so are not on github
-# uncomment the model and run it to reproduce
-
-# liverpool_hsamodel <- gam(unemployment ~ lt_illness +
-#                          s(msoa_cd, bs="re") +
-#                          s(msoa_cd, lt_illness, bs="re") +
-#                          s(lsoa_cd, bs="re") +
-#                          s(lsoa_cd, lt_illness, bs="re") +
-#                          s(oa_cd, bs='mrf',
-#                            xt=list(nb=prepliv$nb),k=20), # ICAR constituency ICAR varying coefficients
-#                        data=prepliv, method="REML")
-
-# saveRDS(liverpool_hsamodel, here("data","liverpool_hsamodel.rds")
-
-liverpool_hsamodel <- readRDS(here("data","liverpool_hsamodel.rds"))
-
-# summary(liverpool_hsamodel)
-
-```
-
-```{r}
-
-liverpool_hmodel_aug <- liverpool_hmodel |> 
-  st_augment(prepliv) # pipe into function to get estimates
-
-liverpool_hsamodel_aug <- liverpool_hsamodel |> 
-  st_augment(prepliv) # pipe into function to get estimates
-
-```
-
-```{r}
-
-liverpool_hmodel_plots <- liverpool_hmodel |> 
-  st_augment(prepliv) |> # pipe into function to get estimates
-  st_quickmap_preds(borderwidth = NA, 
-                    scale_low = "darkblue",
-                    scale_mid = "gray75",
-                    scale_high = "darkred",
-                    scale_midpoint = 0, 
-                    framefill = "#F7FFF2",
-                    frameline = "black",
-                    framesize = 1) # pipe into this for visualisation
-
-```
+will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree. We will
+compare two different types of contiguity which could be applied to the ICAR component:
+
+* `st_bridges()` from `sfislands`, which implements first-order queen contiguity in the same way as `sfdep:st_contiguity()` for situations such as this where there are no islands, and 
+* `st_dist_band()` from \CRANpkg{sfdep}, which connects units whose centroids lie within a chosen distance threshold.
+
+We will show how their outcomes can be quickly mapped to give an immediate sense of the differences implied by each model's structure.
 
@@
 
-Now suppose we wish to model unemployment as a function of limited long-term
+Now suppose we wish to model unemployment as a function of limiting long-term
 illness, taking into account this hierarchical structure.
@@
 We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at MSOA and LSOA levels.
-The
-distribution of these variables across the OAs of Liverpool are shown in Figure
+The raw distribution of these variables across the output areas (OAs) of Liverpool are shown in Figure
 \@ref(fig:liv-vars).
 
-If $y_{ijk}$ and $x_{ijk}$ are respectively the levels of unemployment and LLTI in each of $i$ OAs within $j$ LSOAs, which in turn are within $k$ MSOAs, then 
+If $y_{ijk}$ and $x_{ijk}$ are respectively the levels of unemployment and LLTI in each of $i$ OAs within $j$ lower level super output areas (LSOAs), which in turn are within $k$ middle level super output areas (MSOAs), then
 \begin{equation}
@@
 \end{equation}
-where the $\gamma_{ijk}$s comprise a vector of ICAR components for each OA, similar to the provinces in the earlier Indonesia model.
+where the $\gamma_{ijk}$s comprise a vector of ICAR components for each OA, similar in form to the ICAR component for the provinces in the earlier Indonesia model.
 The vector has a multivariate normal distribution with mean 0 and precision related to the contiguity structure as before.
@@
 
-The hierarchical model with no mrf component produces estimates as shown
-in Figure \@ref(fig:liv-est1).
+```{r}
+livadmin <- tribble(
+  ~code, ~`area type`, ~description,
+  "OA", "output areas", "lowest level of geographical area for census statistics, usually containing 100 - 625 persons",
+  "LSOA", "lower layer super output areas", "usually 4 or 5 OAs",
+  "MSOA", "middle layer super output areas", "usually 4 or 5 LSOAs"
+)
+
+```
+
+```{r livadmin-html, eval = knitr::is_html_output(), layout = "center"}
+knitr::kable(livadmin, format = "html", caption = "Three nested levels of administrative divisions in Liverpool.") |> 
+  kable_styling(full_width = TRUE)
+```
+
+```{r livadmin-latex, eval = knitr::is_latex_output()}
+knitr::kable(livadmin, format = "latex", caption = "Three nested levels of administrative divisions in Liverpool.") |> 
+  row_spec(0, bold = TRUE) |> 
+  kableExtra::kable_styling(font_size = 8) |> 
+  column_spec(3, width = "7cm")
+```
+
+```{r, liv-nested, fig.width=10, fig.height=5, out.width="100%", fig.cap = "Nested levels of administrative divisions in Liverpool. "}
+
+ggarrange(
+  ggplot() + 
+    geom_sf(data=msoa_sf, fill="gray90", colour="black") +  
+    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
+    geom_sf(data=liverpool |> summarise(), colour="black", fill=NA) + 
+    labs(title = "level 1: MSOA polygons") + 
+  coord_sf(datum=NA) + 
+  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
+                                        linewidth=1.5),
+        axis.text = element_blank()),
+
+  ggplot() + 
+    geom_sf(data=lsoa_sf, fill="gray90", colour="gray40") +  
+    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
+    geom_sf(data=liverpool |> summarise(), colour="black", fill=NA) + 
+    labs(title = "level 2: LSOA polygons") + 
+  coord_sf(datum=NA) + 
+  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
+                                        linewidth=1.5),
+        axis.text = element_blank()),
+
+  ggplot() + 
+    geom_sf(data=liverpool |> summarise(), colour="black", fill="white") + 
+    geom_sf(data=liverpool |> st_centroid(), 
+            fill="gray90", colour="gray50", size=0.1) +  
+    geom_sf(data=mersey, fill="lightblue", colour="black", linewidth=0.2) + 
+    labs(title = "level 3: OA (shown as centroids)") + 
+  coord_sf(datum=NA) + 
+  theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
+                                        linewidth=1.5),
+        axis.text = element_blank()),
+
+  ncol=3
+)
+
+```
+
+```{r, liv-vars, fig.width=6, fig.height=4, out.width="70%", fig.cap = "Unemployment and limiting long-term illness rates in Liverpool, with extent of Mersey estuary in light blue to the south west.
+"}
+
+ggarrange(
+  ggplot() + 
+    geom_sf(data=liverpool, aes(fill=unemployment, colour=unemployment)) + 
+    geom_sf(data=liverpool |> summarise(), 
+            fill=NA, colour="gray30", linewidth=0.2) +  
+    geom_sf(data=mersey, fill="lightblue", colour="gray30", linewidth=0.2) + 
+    scale_fill_distiller(palette = "PuBu", direction = 1) + 
+    scale_colour_distiller(palette = "PuBu", direction = 1) + 
+    labs(fill="unemployment",
+         colour="unemployment") + 
+    coord_sf(datum=NA) + 
+    annotation_scale(location = "bl", width_hint = 0.4) +
+    annotation_north_arrow(location = "bl", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.2, "in"),
+        style = north_arrow_fancy_orienteering) +
+    theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
+                                          linewidth=0.6), 
+          axis.title = element_blank(),
+          axis.ticks = element_blank(),
+          legend.text = element_text(size=8),
+          legend.title = element_text(size=8)),
+  ggplot() + 
+    geom_sf(data=liverpool, aes(fill=lt_illness, colour=lt_illness)) + 
+    geom_sf(data=liverpool |> summarise(), 
+            fill=NA, colour="gray30", linewidth=0.2) +  
+    geom_sf(data=mersey, fill="lightblue", colour="gray30", linewidth=0.2) + 
+    scale_fill_distiller(palette = "RdYlBu", direction = -1) + 
+    scale_colour_distiller(palette = "RdYlBu", direction = -1) + 
+    labs(fill="limiting long-term\nillness\n(z-score)",
+         colour="limiting long-term\nillness\n(z-score)") + 
+    coord_sf(datum=NA) +
+    annotation_scale(location = "bl", width_hint = 0.4) +
+    annotation_north_arrow(location = "bl", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.2, "in"),
+        style = north_arrow_fancy_orienteering) +
+    theme(panel.background = element_rect(fill = "#F7FFF2", colour = "black", 
+                                          linewidth=0.6), 
+          axis.title = element_blank(),
+          axis.ticks = element_blank(),
+          legend.text = element_text(size=8),
+          legend.title = element_text(size=8)),
+
+  ncol=2, legend = "bottom"
+)
+
+```
+
+```{r, liv-contigs, fig.width=5, fig.height=5, out.width="60%", echo=TRUE, fig.cap = "First order queen contiguity of output areas (OAs) in Liverpool with magnified inset. "}
+# show st_bridges() connection with a magnified section
+
+from <- c(xmin = -2.97, xmax = -2.95, ymin = 53.38, ymax = 53.39)
+to <- c(-3.01,-2.96,53.32,53.35)
+
+st_bridges(liverpool, "oa_cd") |> 
+  st_quickmap_nb(pointsize = 0.1,
+                 linksize = 0.1,
+                 fillcol = "gray90") +  
+  geom_sf(data=mersey, fill="lightblue", colour="gray30", 
+          linewidth=0.2, alpha=0.5) + 
+  geom_magnify(data=liverpool, 
+               from = from, to = to, 
+               corners = 0.1, shadow = TRUE, linewidth = 0.6) + 
+  theme(panel.background = element_rect(fill = "#F7FFF2",
+                                        colour = "black", linewidth=1))
+
+```
+
+To demonstrate how `sfislands` can be used to quickly compare the output of models, we show the output of three similar models.
+The first of these is a hierarchical model as described above, but with no ICAR component at the lowest level.
+Its spatially varying coefficients are shown in Figure \@ref(fig:liv-est1).
+Secondly, we fit the model described above using `st_bridges()` to generate first-order queen contiguities as the structure for the neighbourhood relationships.
+The coefficients of this model are shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'liv-est2html', 'liv-est2')`).
+
+Finally, we can fit a similar model but this time using a within-distance criterion of 700m for neighbourhood status.
+All units within this distance of each other are considered neighbours.
+This was generated with the `sfdep:st_dist_band()` function.
+These coefficients are shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'liv-est3html', 'liv-est3')`).
+
+This shows how `sfislands` can quickly give an intuitive representation of how a set of similar models are operating.
+As a measure of the relative strengths of these three structures, we could compare their AICs (see Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:aic-html)', '\\@ref(tab:aic-latex)'))`).
+These suggest that, in this scenario, inclusion of the ICAR component is appropriate, and queen contiguity is a more suitable choice than the distance-band condition for defining the neighbourhood structure.
+
+```{r}
+
+# these models take a while to run
+# so i am using a saved version of their output
+# the files are large so are not on github
+# uncomment the model and run it to reproduce
+
+prepliv <- st_bridges(liverpool, "oa_cd")
+
+# liverpool_model2 <- gam(unemployment ~ lt_illness + 
+#                          s(msoa_cd, bs="re") + 
+#                          s(msoa_cd, lt_illness, bs="re") +
+#                          s(lsoa_cd, bs="re") + 
+#                          s(lsoa_cd, lt_illness, bs="re"), # ICAR constituency ICAR varying coefficients
+#                        data=prepliv, method="REML")
+
+# saveRDS(liverpool_model2, here("data","liverpool_model2.rds"))
+
+liverpool_hmodel <- readRDS(here("data","liverpool_model2.rds"))
+
+# summary(liverpool_hmodel)
+
+```
+
+```{r}
+
+# these models take a while to run
+# so i am using a saved version of their output
+# the files are large so are not on github
+# uncomment the model and run it to reproduce
+
+# liverpool_hsamodel <- gam(unemployment ~ lt_illness +
+#                          s(msoa_cd, bs="re") +
+#                          s(msoa_cd, lt_illness, bs="re") +
+#                          s(lsoa_cd, bs="re") +
+#                          s(lsoa_cd, lt_illness, bs="re") +
+#                          s(oa_cd, bs='mrf',
+#                            xt=list(nb=prepliv$nb),k=20), # ICAR constituency ICAR varying coefficients
+#                        data=prepliv, method="REML")
+
+# saveRDS(liverpool_hsamodel, here("data","liverpool_hsamodel.rds")
+
+liverpool_hsamodel <- readRDS(here("data","liverpool_hsamodel.rds"))
+
+# summary(liverpool_hsamodel)
+
+```
+
+```{r}
+
+liverpool_hmodel_aug <- liverpool_hmodel |> 
+  st_augment(prepliv) # pipe into function to get estimates
+
+liverpool_hsamodel_aug <- liverpool_hsamodel |> 
+  st_augment(prepliv) # pipe into function to get estimates
+
+```
+
+```{r}
+
+liverpool_hmodel_plots <- liverpool_hmodel |> 
+  st_augment(prepliv) |> # pipe into function to get estimates
+  st_quickmap_preds(borderwidth = NA, 
+                    scale_low = "darkblue",
+                    scale_mid = "gray75",
+                    scale_high = "darkred",
+                    scale_midpoint = 0, 
+                    framefill = "#F7FFF2",
+                    frameline = "black",
+                    framesize = 1) # pipe into this for visualisation
+
+```
 
@@
 
-If we instead use contiguity as the basis of an additional spatial term,
-we get estimates shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'liv-est2html', 'liv-est2')`).
-
 (ref:my-caption6) Random intercept and slope estimates from hierarchical model.
@@
 
-If instead we consider units to be neighbours if they lie within a
-distance band of 700 metres, we get estimates shown in Figure
-\@ref(fig:`r ifelse(knitr::is_html_output(), 'liv-est3html', 'liv-est3')`).
-
-A comparison of their AICs (see Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:aic-html)', '\\@ref(tab:aic-latex)'))`) suggests that, in this scenario, queen contiguity is more appropriate than the distance band condition.
-
 (ref:my-caption7) Random intercept and slope estimates from hierarchical model.
@@
 knitr::kable(aic_tab, format = "latex", caption = "Comparison of models by AIC.") |> 
+  row_spec(0, bold = TRUE) |> 
   kableExtra::kable_styling(font_size = 8)

### 2024-03-15 article.tex
revisions to article

@@
 creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
-structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} (Pebesma 2018) package (Parry and Locke 2024) has presented the same
+structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} package (Parry and Locke 2024) has presented the same
 functionality in addition to extra features in a tidy structure based on
@@
 The above can be considered as the \emph{pre-functions} of the package. A
-further set of feature of \texttt{sfislands}, which we refer to as \emph{post-functions},
-are for use with the \CRANpkg{mgcv} package (Wood 2011). It is straightforward to use
-\texttt{sfislands} neighbourhood structures in models with \CRANpkg{mgcv} but it can be quite awkward
-to extract the estimates.
-These \emph{post-functions} conveniently provide the estimates in tidy format of random effects and intrinsic conditional autoregressive (ICAR) components from such models and give the ability to quickly map them for visualisation purposes.
+second category of features, which we refer to as \emph{post-functions},
+are for use with the \CRANpkg{mgcv} package (Wood 2011).
+Having fit a model with \CRANpkg{mgcv}, the process of extracting estimates for certain types of effects can be somewhat awkward.
+These \emph{post-functions} augment the original dataframe with these estimates and their standard errors in tidy format.
+They also allow for quick visualisation of the output in map form.
 
@@
 
-It greatly speeds up the workflow of fitting areal spatial models.
-
-\begin{itemize}
+It greatly speeds up each stage of a proposed workflow for fitting areal spatial models.
+
+\hypertarget{step-1-pre-functions-for-setting-up-neighbourhood-structure}{%
+\subsubsection{\texorpdfstring{Step 1: \emph{Pre-functions} for setting up neighbourhood structure}{Step 1: Pre-functions for setting up neighbourhood structure}}\label{step-1-pre-functions-for-setting-up-neighbourhood-structure}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \item
-  \textbf{pre-functions} for setting up contiguities:
-
-  \begin{itemize}
+  It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data.
+\texttt{sfislands} will include a contiguity for all units.
 \item
-    it addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data.
-\texttt{sfislands} will include a contiguity for all units,
+  It gives tools to immediately visualise this structure as a map.
 \item
-    it gives tools to immediately visualise this structure as a map,
+  These maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax.
 \item
-    these maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax,
+  As the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence.
 \item
-    as the nodes can be labelled by index, it makes it very easy to add and remove connections as appropriate with confidence,
+  Connections which have been induced by a function from the package but are not contiguities can be accessed to ensure openness in the process.
+\end{enumerate}
+
+\hypertarget{step-2-modelling}{%
+\subsubsection{Step 2: Modelling}\label{step-2-modelling}}
+
+These neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms} (Bürkner 2017), \pkg{r-inla} (Bakka et al. 2018) and more.
+
+\hypertarget{step-3-post-functions-for-mgcv-models}{%
+\subsubsection{\texorpdfstring{Step 3: \emph{Post-functions} for \texttt{mgcv} models}{Step 3: Post-functions for mgcv models}}\label{step-3-post-functions-for-mgcv-models}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \item
-    connections made by the package which are not contiguities can be accessed to ensure openness in the process.
-  \end{itemize}
+  It simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models).
 \item
-  For modelling:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    these neighbourhood structures can be used in modelling packages such as \CRANpkg{mgcv}, \CRANpkg{brms}. (Bürkner 2017), \pkg{r-inla} (Bakka et al. 2018) and more.
-  \end{itemize}
-\item
-  \textbf{post-functions} for \texttt{mgcv} models:
-
-  \begin{itemize}
-  \item
-    it simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models),
-  \item
-    these effects can be quickly visualised in \CRANpkg{ggplot2} maps.
-  \end{itemize}
-\end{itemize}
+  These effects can be quickly visualised as \CRANpkg{ggplot2} maps.
+\end{enumerate}
 
@@
 The first group of functions, shown in Table \ref{tab:prefunc-latex}, deals with the creation
-of a contiguity structure in the presence of discontiguities. The
+of a neighbourhood structure in the presence of discontiguities. The
 resultant structure can be quickly mapped to check if it is
@@
 index number. By an iterative process of changes and examination of a
-quick map, a satisfactory structure can be decided upon.
-
-Table \ref{tab:prefunc-latex} shows these functions and their purpose.
+quickly-generated guide map, a satisfactory structure can be decided upon.
 
@@
 \hline
-function & purpose\\
+\textbf{function} & \textbf{purpose}\\
 \hline
@@
 
-They deal with the creation
-of a contiguity structure in the presence of discontiguities. The
-resultant structure can be quickly mapped to check if it is
-satisfactory. Connections can be manually added or removed by name or
-index number. By an iterative process of changes and examination of a
-quick map, a satisfactory structure can be decided upon.
-Going through each function in more detail:
-
-Use the rectangles shown in Figure \ref{fig:rects1} as a simple demonstration. Rectangles 1-3
-are contiguous while 4 and 5 are islands.
+We will now go through each function in more detail using the set of rectangles shown in Figure \ref{fig:rects1} for demonstration purposes.
+Rectangles 1, 2 and 3 are contiguous while 4 and 5 can be viewed as ``islands''.
 
@@
 
-This function requires as arguments an \CRANpkg{sf} dataframe and the name of a column of unique identifiers, preferably names, of each spatial unit.
-The k-nearest neighbours for connecting islands can then be specified.
+This function requires at least two arguments: an \CRANpkg{sf} dataframe and, from that, the name of one column of unique identifiers, ideally names, of each spatial unit.
+It creates a neighbourhood structure where non-island units are joined by first-order queen contiguity, while island units are joined to their k-nearest neighbours.
 The output is a \emph{named} neighbourhood structure in either list or matrix form as desired, which can be either a standalone object or included as an additional column in the original \CRANpkg{sf} dataframe.
@@
 
-In the following examples, we have chosen to ignore islands with the argument \texttt{remove\_islands\ =\ TRUE} and to return a list and matrix structure respectively by specifying this in the \texttt{nb\_structure} argument and choosing \texttt{add\_to\_dataframe\ =\ FALSE}:
+In the first two examples of using \texttt{st\_bridges()}, we have chosen to ignore islands with the argument \texttt{remove\_islands\ =\ TRUE} and to return a list and matrix structure respectively by specifying this in the \texttt{nb\_structure} argument and choosing \texttt{add\_to\_dataframe\ =\ FALSE}:
 
@@
 
-In these examples, we choose to join islands to their 1 nearest
+Alternatively, in the following examples, we choose to join islands to their 1 nearest
 neighbour, which is the default setting, and to return the output as a column called ``nb'' in the original
-\CRANpkg{sf} dataframe (this is the default setting):
+\CRANpkg{sf} dataframe (\texttt{add\_to\_dataframe\ =\ "TRUE"} is the default setting):
 
@@
 #> CRS:           NA
-#>                         geometry  name      nb
-#> 1 POLYGON ((0 0, 0 2, 2 2, 2 ... Rect1    2, 3
-#> 2 POLYGON ((2 0, 2 2, 4 2, 4 ... Rect2 1, 3, 4
-#> 3 POLYGON ((2 2, 2 4, 4 4, 4 ... Rect3 1, 2, 5
-#> 4 POLYGON ((5 0, 5 1, 6 1, 6 ... Rect4       2
-#> 5 POLYGON ((0.8 3, 0.8 4, 1.8... Rect5       3
+#>    name      nb                       geometry
+#> 1 Rect1    2, 3 POLYGON ((0 0, 0 2, 2 2, 2 ...
+#> 2 Rect2 1, 3, 4 POLYGON ((2 0, 2 2, 4 2, 4 ...
+#> 3 Rect3 1, 2, 5 POLYGON ((2 2, 2 4, 4 4, 4 ...
+#> 4 Rect4       2 POLYGON ((5 0, 5 1, 6 1, 6 ...
+#> 5 Rect5       3 POLYGON ((0.8 3, 0.8 4, 1.8...
 \end{verbatim}
@@
 #> CRS:           NA
-#>                         geometry  name nb.1 nb.2 nb.3 nb.4 nb.5
-#> 1 POLYGON ((0 0, 0 2, 2 2, 2 ... Rect1    0    1    1    0    0
-#> 2 POLYGON ((2 0, 2 2, 4 2, 4 ... Rect2    1    0    1    1    0
-#> 3 POLYGON ((2 2, 2 4, 4 4, 4 ... Rect3    1    1    0    0    1
-#> 4 POLYGON ((5 0, 5 1, 6 1, 6 ... Rect4    0    1    0    0    0
-#> 5 POLYGON ((0.8 3, 0.8 4, 1.8... Rect5    0    0    1    0    0
+#>    name nb.1 nb.2 nb.3 nb.4 nb.5                       geometry
+#> 1 Rect1    0    1    1    0    0 POLYGON ((0 0, 0 2, 2 2, 2 ...
+#> 2 Rect2    1    0    1    1    0 POLYGON ((2 0, 2 2, 4 2, 4 ...
+#> 3 Rect3    1    1    0    0    1 POLYGON ((2 2, 2 4, 4 4, 4 ...
+#> 4 Rect4    0    1    0    0    0 POLYGON ((5 0, 5 1, 6 1, 6 ...
+#> 5 Rect5    0    0    1    0    0 POLYGON ((0.8 3, 0.8 4, 1.8...
 \end{verbatim}
@@
 These structures can serve as the input to models in \CRANpkg{brms},
-\pkg{r-inla}, \pkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. Rather than having a separate neighbours
-column, it is included as a named list or matrix in the original \CRANpkg{sf}
-dataframe, in the spirit of the \CRANpkg{sfdep} package.
+\pkg{r-inla}, \pkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
+object, it is included in the original \CRANpkg{sf} dataframe as a named list or matrix, in the spirit of the \CRANpkg{sfdep} package.
 
@@
 
-This function will tell us openly what connections have been made which
-are not based on contiguity. It gives both the name and index number of each pair of added connections.
-
-\begin{verbatim}
+This function will show us transparently what connections have been made which
+are not based on contiguity. It gives both the name and index number of each pair of added connections. In this example, two pairs have been added.
+
+\begin{verbatim}
+# show summary of non-contiguous connections in a dataframe
+
 st_bridges(rectangles, 
@@
 \begin{verbatim}
+# add an extra connection using numeric index
+
 st_bridges(rectangles, "name", 
@@
 \begin{verbatim}
+# remove an existing connection using unit name, not index
+
 st_bridges(rectangles, "name", 
@@
 
-The use of these structures is particularly common in CAR-type models.
+The use of these neighbourhood structures is particularly associated with CAR-type models.
 These are often implemented in a Bayesian framework using
-\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. The pre-functions will output the
+\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
 neighbourhood structure in the desired format for use in any of these
 frameworks. A convenient frequentist alternative is to use the
-\CRANpkg{mgcv} package. It has the functionality to create such models
+\CRANpkg{mgcv} package which requires a named list of neighbours. It has the functionality to create such models
 using \texttt{bs="mrf"}. It also has the ability to combine these with a
@@
 Bayesian structures mentioned above can be extracted in the same way as
-any other component of the model, it can be a bit awkward to get the
+any other component of the model, it can be somewhat awkward to get the
 estimates from \CRANpkg{mgcv} model. \texttt{sfislands} has two post-functions to conveniently extract and visualise these.
@@
 This function augments the original dataframe with the estimated means
-and standard errors from the \CRANpkg{mgcv} model in a similar manner to
-how the \CRANpkg{broom} (Robinson, Hayes, and Couch 2023) package operates.
-The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the new random effects columns are positioned immediately before it.
-The estimates which \texttt{st\_augment()} will extract are random effects (which are called in \CRANpkg{mgcv} with \texttt{bs=\textquotesingle{}re\textquotesingle{}}) and ICAR components (\texttt{bs=\textquotesingle{}mrf\textquotesingle{}}).
+and standard errors of the spatially varying predictions from a fitted \CRANpkg{mgcv} model in a similar manner to
+how the \CRANpkg{broom} package (Robinson, Hayes, and Couch 2023) operates.
+The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it.
+\footnote{In a similar way, \texttt{st\_augment()} can also be used to append the random effects from \texttt{lme4} and \texttt{nlme} models to an \texttt{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}.} The spatially varying predictions which \texttt{st\_augment()} extracts from an \CRANpkg{mgcv} model are
+
+\begin{itemize}
+\tightlist
+\item
+  random effects (which are called in \CRANpkg{mgcv} with \texttt{bs=\textquotesingle{}re\textquotesingle{}}), and
+\item
+  ICAR components (\texttt{bs=\textquotesingle{}mrf\textquotesingle{}}).
+\end{itemize}
 
 Consider the model structure described in the code below using \CRANpkg{mgcv} syntax.
-In this model \(y\) is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a random intercept and slope for the covariate at a \emph{region} level, and an ICAR varying intercept and slope at a lower \emph{sub-region} level.
-
-\begin{verbatim}
+In this model \(y\) is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a \emph{region} level, and a set of ICAR varying intercepts and slopes at a lower \emph{sub-region} level.
+
+\begin{verbatim}
+# creating an mgcv model
+
 mgcv::gam(
@@
 
-When labelling the new columns which are augmented to the original dataframe from such a model, \texttt{st\_augment()} follows the formula syntax of the \CRANpkg{lme4} (Bates et al.
-2015) package, where the \emph{pipe symbol} (\texttt{\textbar{}}) indicates ``\emph{grouped by}''.
-Table \ref{tab:staugtab-latex} shows how the augmented columns in this scenario would be labelled.
+When labelling the new prediction columns which are augmented to the original dataframe from such a model, \texttt{st\_augment()} follows the formula syntax of the \CRANpkg{lme4} package (Bates et al.
+2015), where the pipe symbol (\texttt{\textbar{}}) indicates ``\emph{grouped by}''.
+Table \ref{tab:staugtab-latex} shows how the augmented columns in this scenario would be named.
 Each column name begins with either \texttt{random.effect.} or \texttt{mrf.smooth.} as appropriate.
-An additional column is also added for the standard error of each estimate.
+An additional column is also added for the standard error of each estimate, as calculated by \CRANpkg{mgcv}.
 These columns are named as above but with \texttt{se.} prepended (e.g.~\texttt{se.random.effect.region}).
@@
 \hline
-mgcv syntax & column name\\
+\textbf{mgcv syntax} & \textbf{column name}\\
 \hline
@@
 These estimates can then be quickly mapped.
-As it is possible to include more than 1 varying component, the output of this function is a list of plots.
-They can be viewed individually by indexing, or all at one using the \texttt{plotlist=} argument from the \texttt{ggarrange()} function which is part of the \CRANpkg{ggpubr} (Kassambara 2023) package.
-We will see examples of this in the following examples.
-The maps are automatically titled and subtitled according to the type of effect.
-For example, the map showing \texttt{random.effect.region} will have ``\emph{region}'' as its title and ``\emph{random.effect}'' as its subtitle.
+As it is possible to include more than 1 spatially varying component, the output of this function is a list of plots.
+They can be viewed individually by indexing, or all at one using, for example, the \texttt{plotlist} argument from the \texttt{ggarrange()} function which is part of the \CRANpkg{ggpubr} (Kassambara 2023) package.
+We will see this function in practice in the following examples.
+The maps which it generates are automatically titled and subtitled according to the type of effect.
+For example, the map showing predictions for \texttt{random.effect.region} will have ``\emph{region}'' as its title and ``\emph{random.effect}'' as its subtitle.
 
@@
 
-The data for this section have been taken from the \href{https://earthquake.usgs.gov/earthquakes/search/}{USGS earthquake
-catalogue API}. The
-datasets and an explanation are available at
-\url{https://github.com/horankev/quake_data} and record all earthquakes in
-and close to Indonesia from 1985-2023. Below is a map of Indonesia, with
-other neighbouring or bordering countries filled in grey. The many local
-faults which lie within 300km of the shore are shown as green lines
-in Figure \ref{fig:fault-buffers}.
+The data for this section have been downloaded from the National Earthquake Information Center, \href{https://earthquake.usgs.gov/earthquakes/search/}{USGS earthquake
+catalogue}. The
+datasets with accompanying explanations are available at
+\url{https://github.com/horankev/quake_data}. They capture all recorded earthquakes in
+and close to Indonesia from the beginning of January 1985 to the end of December 2023.
+Figure \ref{fig:fault-buffers} shows a map of Indonesia, divided into 33 provinces, with other neighbouring or bordering countries filled in grey.
+The many local
+faults which lie within 300km of the shore are shown in yellow with green outlines.
 
@@
 
-Earthquake activity will be seen as the total number of earthquakes with
-their epicentre in each province per unit area of that province. We have
-restricted counts to earthquakes of magnitude \textgreater5.5 which is the point
-at which they are often labelled as damaging.
-
-The occurrences of these earthquakes are shown in Figure \ref{fig:quake-occur}, their total per province in Figure \ref{fig:quake-totals}, and finally, their count per square kilometre can be seen in Figure \ref{fig:quake-conc}.
+Earthquake incidence per province has been calculated as the total number of earthquakes with
+an epicentre within that province per unit area. We have
+restricted counts to earthquakes \textgreater5.5 on the moment magnitude scale, which is the point
+at which they are often labelled as potentially damaging.
+
+The occurrences of these earthquakes are shown in Figure \ref{fig:quake-occur}, their total per province in Figure \ref{fig:quake-totals}, and finally, their incidence or count per square kilometre can be seen in Figure \ref{fig:quake-conc}.
 
@@
 
-\caption{Earthquakes in Indonesia, 1985-2023. Categorised as medium, large or extra-large. }\label{fig:quake-occur}
+\caption{Earthquakes in Indonesia of magnitude > 5.5, 1985-2023. Categorised by magnitude as medium, large or extra-large. }\label{fig:quake-occur}
 \end{figure}
@@
 \begin{equation}
-y_i = \text{earthquake incidence}_i = \frac{\text{earthquake count in province}_i} {\text{province area}_i}, 
+y_i = \text{earthquake count}_i 
 \label{eq:eq1}
@@
 \begin{equation}
-x_i = \text{fault intensity}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
+x_i = \text{fault concentration}_i = \frac{\text{area of buffered faults in province}_i} {\text{province area}_i}.
 \label{eq:eq2}
@@
 
+Having described the type of model we wish to implement, we now show how \texttt{sfislands} can be used to streamline the process.
+
 \hypertarget{pre-functions-1}{%
@@
 This is also shown in Figure \ref{fig:ind-contig2}.
-These shapes are not used in the assignment of contiguities but it can be useful to see them in a situation such as Indonesia where many individual provinces are actually multipolygons.
+These shapes are not used in the assignment of contiguities but it can be useful to see them in a situation such as Indonesia where many individual provinces are actually multipolygons of more than one island.
 Without them, it is not clear whether an island is a province in its own right, or which group of islands together form one province.
@@
 new structure can be seen in Figure \ref{fig:ind-contig3}. Edge effects
-have also been mitigated by additionally connecting the two extreme
-provinces (1 and 23), which would otherwise have only one neighbour, to
+have also been mitigated by imposing additional connections on the two extreme
+provinces (1 and 23), which would otherwise have only one neighbour, so that they now also include
 their two next closest neighbours.
@@
 
-We now create the ICAR model using \CRANpkg{mgcv}. We will be able to simply pass
-the contiguity structure we have created as both the data and a column
-of it as the neighbourhood structure.
+We now create the ICAR model using, in this case, the \CRANpkg{mgcv} package.
+We will be able to use the output of \texttt{st\_bridges}, which we have named \texttt{prep\_data}, as both the data source for the model and the neighbourhood structure (by specifying the column \texttt{nb} which contains the neighbourhood list).
 
@@
 
-The adjusted R-squared is \textbf{0.957} and deviance explained is \textbf{88.2\%}. The estimated coefficient associated with fault concentration is
--3.2355. As this is positive, it is greater than 1 when exponentiated:
-\[
-\text{exp}(0.7531) = 0.04
-\]
-and this is the expected multiplicative association between a unit
-change in fault concentration and incidence of earthquakes in a
-province.
-
-Controlling for this estimated mean elevated risk of earthquakes due to
-concentration of faults, what is the additional risk level of
+We can see from the summary below that the adjusted R-squared is \textbf{0.957} and deviance explained is \textbf{88.2\%}.
+
+\begin{verbatim}
+#> 
+#> Family: poisson 
+#> Link function: log 
+#> 
+#> Formula:
+#> damaging_quakes_total ~ fault_concentration + s(province, bs = "mrf", 
+#>     xt = list(nb = prep_data$nb), k = 22) + offset(log(area_province))
+#> 
+#> Parametric coefficients:
+#>                     Estimate Std. Error z value Pr(>|z|)    
+#> (Intercept)          -9.1066     0.1428 -63.774   <2e-16 ***
+#> fault_concentration  -3.2355     1.3297  -2.433    0.015 *  
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> Approximate significance of smooth terms:
+#>               edf Ref.df Chi.sq p-value    
+#> s(province) 18.94     21  147.7  <2e-16 ***
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> R-sq.(adj) =  0.957   Deviance explained = 88.2%
+#> -REML = 119.75  Scale est. = 1         n = 33
+\end{verbatim}
+
+Controlling for the concentration of faults, what is the additional risk level of
 earthquakes in a province? This can be seen as a measure of the activity
 level of faults locally and it is spatially smoothed by the
-autoregressive process. It is mapped as the ICAR component of the model
-below:
+autoregressive process.
+It is represented in the model summary by the component \texttt{s(province)}.
+However, the extraction of individual predictions for this component for each province from the \texttt{mgcv} model requires a number of steps.
+We know demonstrate how these are streamlined into a single function by the \texttt{sfislands} package.
 
@@
 
-Use \texttt{st\_augment()} to get estimates from \CRANpkg{mgcv} model in format
-similar to that used by the \CRANpkg{broom} package.
-For instance, we see from the code below that the original dataframe is now augmented with columns for ``mrf.smooth.province'' and ``mrf.smooth.province'' which shows the estimate and standard errors for the \(\gamma_i\) component.
+The function \texttt{st\_augment()} allow us to add the spatially varying predictions from the model as new columns to the original dataframe in a process
+similar to that of the \CRANpkg{broom} package.
+For instance, we see from the code below that the original dataframe is now augmented with columns called ``mrf.smooth.province'' and ``se.mrf.smooth.province'' which show the predictions for the \(\gamma_i\) component and their standard errors.
 Note that this is how we would expect them to be named, based on the previous discussion surrounding Table \ref{tab:staugtab-latex}.
-
-\begin{verbatim}
+They are positioned immediately before the final \texttt{geometry} column of the \texttt{sf} dataframe.
+
+\begin{verbatim}
+# column names of augmented dataframe
+
 mod_pois_mrf |> 
@@
 \begin{verbatim}
-#>  [1] "S"                       "M"                      
-#>  [3] "L"                       "XL"                     
-#>  [5] "quake_total"             "quake_density"          
-#>  [7] "damaging_quakes_total"   "damaging_quakes_density"
-#>  [9] "area_fault_within"       "area_province"          
-#> [11] "fault_concentration"     "nb"                     
-#> [13] "mrf.smooth.province"     "se.mrf.smooth.province" 
-#> [15] "province"                "province_id"            
+#>  [1] "province"                "province_id"            
+#>  [3] "S"                       "M"                      
+#>  [5] "L"                       "XL"                     
+#>  [7] "quake_total"             "quake_density"          
+#>  [9] "damaging_quakes_total"   "damaging_quakes_density"
+#> [11] "area_fault_within"       "area_province"          
+#> [13] "fault_concentration"     "nb"                     
+#> [15] "mrf.smooth.province"     "se.mrf.smooth.province" 
 #> [17] "geometry"
@@
 
-Use \texttt{st\_augment()} and \texttt{st\_quickmap\_preds} in the following code to get a quick visualisation of
-the estimates for \(\gamma_i\), as shown in Figure \ref{fig:ind-mrf1}. Again, note that the title and subtitle of the image are as previously discussed.
+This output can now be piped into the \texttt{st\_quickmap\_preds()} function to get a quick visualisation of
+these estimates for \(\gamma_i\) on a map, as shown in Figure \ref{fig:ind-mrf1}. Again, note that the title and subtitle of the image are as previously discussed.
 
@@
 
-\hypertarget{workflow}{%
-\subsection{Workflow}\label{workflow}}
+\hypertarget{workflow-summary}{%
+\subsection{Workflow summary}\label{workflow-summary}}
 
@@
 However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps.
-A rough and ready model (and visual output) can be generated with \texttt{sfislands} using nothing more than three or four lines of code, such as the following:
-
-\begin{verbatim}
-# set up neighbourhood structure
+A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with \texttt{sfislands} using nothing more than three or four lines of code, such as the following:
+
+\begin{verbatim}
+# workflow:
+
+# 1. set up neighbourhood structure
 
@@
 
-The next example looks only at using the \emph{pre-functions}, but in a situation where the presence of islands is not the issue.
+The next example looks only at using the \emph{pre-functions} of \texttt{sfislands}, but in a situation where the presence of actual \emph{islands} is not the problem we seek to deal with.
 Consider the wards and boroughs of London (sourced from the Greater London Authority's \href{https://data.london.gov.uk/}{London Datastore}).
@@
 \CRANpkg{sfdep}. The \texttt{st\_quickmap\_nb()} function gives an immediate
-visual representation of the structure. This can be supplemented by
-adding a layer showing the course of the river Thames.
+visual representation of the structure.
+\footnote{\texttt{st\_quickmap\_nb()} can also be used to visualise any contiguity structure created by \texttt{spdep} or \texttt{sfdep} as long as that structure is included in an \texttt{sf} dataframe as a column named \texttt{nb}.} This can be supplemented by
+adding a layer showing the course of the river Thames which is also visible in Figure \ref{fig:lon-contig1}.
 
@@
 
-When the study area has a river running through it there can be problems
-with constructing neighbourhood structures. Depending on how the
-geometries are defined, the presence of a river can either mean that no
-units on either side of its banks are considered neighbours or, at the
-other extreme, the river is essentially ignored and all units on
-opposing banks are considered neighbours.
-
-Depending on the presence of river crossings, two areas which are
-physically quite close might be very distinct. Somebody living on the
-banks of a river might be more likely to go about their life primarily
-on their side of the river, despite the short distance as the crow flies
-of facilities on the other side.
-
-\texttt{sfislands} makes it easier to take account of this situation. We can look
-at just those boroughs which are on either side of the river Thames (Figure
-\ref{fig:lon-contig2})
+When a study area has a river running through it, problems can arise with constructing appropriate neighbourhood structures. Depending on how the
+geometries are defined, the presence of a river can cause problems in two ways.
+In one situation, the river could be expressed as a polygon in its own right meaning that it severs any potential contiguity between units on either side of its banks.
+In this situation, no spatial units will be neighbours with the units directly across the river from them.
+At the other extreme, if the river is not included as a geometry (as is the case here) all units on opposing banks are considered neighbours.
+
+Depending on the presence of river crossings, two areas which are physically quite close but on opposing banks might be very distinct.
+If there is no means of crossing the river within a reasonable distance, somebody living on the banks of a river might be more likely to go about their life primarily on their side of the river, despite the short distance as the crow flies of facilities on the other side.
+This could be relevant in terms of, say, modelling of house prices where we might want to incorporate these issues into a neighbourhood structure.
+
+\texttt{sfislands} provides convenient functions for this sort of situation.
+Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames (Figure \ref{fig:lon-contig2}).
 
@@
 In order to take account of actual connectivity, we can add a layer
-showing the road and pedestrian bridges or tunnels. These were scraped from a Wikipedia article describing the various crossings of the river Thames. In Figure \ref{fig:lon-crossings}, we
-have also drawn a 1 kilometre buffer around them. Until we get to the
-eastern portion of the map, all units on opposing sides of the river are
-within 1 kilometre of a crossing. This is not the case for the extreme
-eastern units so we might not want to consider these units neighbours if
-the river flows between them.
+showing the road and pedestrian bridges or tunnels.
+Details of these were sourced Wikipedia (2024), which describes the various crossings of the river Thames.
+In Figure \ref{fig:lon-crossings}, we have also drawn a 1 kilometre buffer around each crossing.
+This was chosen as an arbitrary measure of what might be considered a ``reasonable'' distance within which to consider opposing banks as being connected.
+The vast majority of units on opposing banks have access to a river crossing within this threshold and thus should be considered as neighbours.
+Only the extreme eastern units and one to the south west should not have a connection across the river.
 
@@
 This allows us to easily cut the ties across the river for these units
-manually by using the function \texttt{st\_manual\_cut\_nb()}. The results of this editing procedure are shown in Figure \ref{fig:lon-contig4}.
-While we are using the index of the units in this example, the function also
-accepts names as arguments which may be more convenient in some circumstances. Having made these adjustments,
-\texttt{st\_quickmap\_nb()} now shows a connectivity structure (Figure \ref{fig:lon-contig4}) which reflects the influence of the
-presence or absence of river crossings.
+manually by using the function \texttt{st\_manual\_cut\_nb()}. \footnote{While we are using the index of the units in this example, the function also
+  accepts names as arguments which may be more convenient in some circumstances.} Having made these adjustments,
+\texttt{st\_quickmap\_nb()} now shows a connectivity structure (Figure \ref{fig:lon-contig4}) which reflects how we have hypothesised that influence should extend across the river in the presence or absence of crossings.
 
@@
   st_quickmap_nb(bordercol = "black", bordersize = 0.5, linksize = 0.5) +
-  geom_sf(data=no_touch_buffer, fill="pink", alpha=0.3) + 
+  geom_sf(data=no_touch_buffer, fill = "pink", alpha = 0.3) + 
   geom_sf(data=crossings_roadped |> st_buffer(1000), 
-          fill="darkgreen", alpha=0.3) +
-  geom_sf(data=thames, colour="blue", linewidth=1.5) +
-  annotation_scale(location="br") +
+          fill= "darkgreen", alpha = 0.3) +
+  geom_sf(data=thames, colour = "blue", linewidth = 1.5) +
+  annotation_scale(location = "br") +
   coord_sf(datum=NA) + 
   theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", 
-                                        linewidth=1.5))
+                                        linewidth = 1.5))
 \end{verbatim}
@@
 
-In this final example, there are no islands or river issues.
+In this final example, there are no islands and no river issues.
 We saw in the Indonesia example how \texttt{sfislands} can extract estimates from the \texttt{mgcv} argument \texttt{bs="mrf"}.
 It can also deal with random effects as defined by \texttt{bs="re"} in \texttt{mgcv} models.
+This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally related components.
 Here, we
-will construct a hierarchical model in this way which also features an ICAR component. We will
-compare two different types of contiguity which could be applied:
+will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree. We will
+compare two different types of contiguity which could be applied to the ICAR component:
 
@@
 \item
-  \texttt{st\_bridges()} from \texttt{sfislands}, and
+  \texttt{st\_bridges()} from \texttt{sfislands}, which implements first-order queen contiguity in the same way as \texttt{sfdep:st\_contiguity()} for situations such as this where there are no islands, and
 \item
-  \texttt{st\_knn()} from \CRANpkg{sfdep}.
+  \texttt{st\_dist\_band()} from \CRANpkg{sfdep}, which connects units whose centroids lie within a chosen distance threshold.
 \end{itemize}
 
-We will show how their outcomes can be quickly mapped to give an immediate sense of the differences.
-
-\begin{table}
-
-\caption{\label{tab:livadmin-latex}Three nested levels of administrative divisions in Liverpool.}
-\centering
-\fontsize{8}{10}\selectfont
-\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{7cm}}
-\hline
-code & area type & description\\
-\hline
-OA & output areas & lowest level of geographical area for census statistics, usually containing 100 - 625 persons\\
-\hline
-LSOA & lower layer super output areas & usually 4 or 5 OAs\\
-\hline
-MSOA & middle layer super output areas & usually 4 or 5 LSOAs\\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/liv-nested-1}
-
-}
-
-\caption{Nested levels of administrative divisions in Liverpool. }\label{fig:liv-nested}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{article_files/figure-latex/liv-vars-1}
-
-}
-
-\caption{Unemployment and limiting long-term illness rates in Liverpool, with extent of Mersey estuary in light blue to the south west. }\label{fig:liv-vars}
-\end{figure}
-
-\begin{verbatim}
-# magnify a section
-
-from <- c(xmin = -2.97, xmax = -2.95, ymin = 53.38, ymax = 53.39)
-to <- c(-3.01,-2.96,53.32,53.35)
-
-st_bridges(liverpool, "oa_cd") |> 
-  st_quickmap_nb(pointsize = 0.1,
-                 linksize = 0.1,
-                 fillcol = "gray90") +  
-  geom_sf(data=mersey, fill="lightblue", colour="gray30", 
-          linewidth=0.2, alpha=0.5) + 
-  geom_magnify(data=liverpool, 
-               from = from, to = to, 
-               corners = 0.1, shadow = TRUE, linewidth = 0.6) + 
-  theme(panel.background = element_rect(fill = "#F7FFF2",
-                                        colour = "black", linewidth=1))
-\end{verbatim}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.6\linewidth]{article_files/figure-latex/liv-contigs-1}
-
-}
-
-\caption{First order queen contiguity of output areas (OAs) in Liverpool with magnified inset. }\label{fig:liv-contigs}
-\end{figure}
+We will show how their outcomes can be quickly mapped to give an immediate sense of the differences implied by each model's structure.
 
@@
 We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at MSOA and LSOA levels.
-The
-distribution of these variables across the OAs of Liverpool are shown in Figure
+The raw distribution of these variables across the output areas (OAs) of Liverpool are shown in Figure
 \ref{fig:liv-vars}.
 
-If \(y_{ijk}\) and \(x_{ijk}\) are respectively the levels of unemployment and LLTI in each of \(i\) OAs within \(j\) LSOAs, which in turn are within \(k\) MSOAs, then
+If \(y_{ijk}\) and \(x_{ijk}\) are respectively the levels of unemployment and LLTI in each of \(i\) OAs within \(j\) lower level super output areas (LSOAs), which in turn are within \(k\) middle level super output areas (MSOAs), then
 \begin{equation}
@@
 \end{equation}
-where the \(\gamma_{ijk}\)s comprise a vector of ICAR components for each OA, similar to the provinces in the earlier Indonesia model.
+where the \(\gamma_{ijk}\)s comprise a vector of ICAR components for each OA, similar in form to the ICAR component for the provinces in the earlier Indonesia model.
 The vector has a multivariate normal distribution with mean 0 and precision related to the contiguity structure as before.
@@
 
-The hierarchical model with no mrf component produces estimates as shown
-in Figure \ref{fig:liv-est1}.
+\begin{table}
+
+\caption{\label{tab:livadmin-latex}Three nested levels of administrative divisions in Liverpool.}
+\centering
+\fontsize{8}{10}\selectfont
+\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{7cm}}
+\hline
+\textbf{code} & \textbf{area type} & \textbf{description}\\
+\hline
+OA & output areas & lowest level of geographical area for census statistics, usually containing 100 - 625 persons\\
+\hline
+LSOA & lower layer super output areas & usually 4 or 5 OAs\\
+\hline
+MSOA & middle layer super output areas & usually 4 or 5 LSOAs\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/liv-nested-1}
+
+}
+
+\caption{Nested levels of administrative divisions in Liverpool. }\label{fig:liv-nested}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth]{article_files/figure-latex/liv-vars-1}
+
+}
+
+\caption{Unemployment and limiting long-term illness rates in Liverpool, with extent of Mersey estuary in light blue to the south west. }\label{fig:liv-vars}
+\end{figure}
+
+\begin{verbatim}
+# show st_bridges() connection with a magnified section
+
+from <- c(xmin = -2.97, xmax = -2.95, ymin = 53.38, ymax = 53.39)
+to <- c(-3.01,-2.96,53.32,53.35)
+
+st_bridges(liverpool, "oa_cd") |> 
+  st_quickmap_nb(pointsize = 0.1,
+                 linksize = 0.1,
+                 fillcol = "gray90") +  
+  geom_sf(data=mersey, fill="lightblue", colour="gray30", 
+          linewidth=0.2, alpha=0.5) + 
+  geom_magnify(data=liverpool, 
+               from = from, to = to, 
+               corners = 0.1, shadow = TRUE, linewidth = 0.6) + 
+  theme(panel.background = element_rect(fill = "#F7FFF2",
+                                        colour = "black", linewidth=1))
+\end{verbatim}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.6\linewidth]{article_files/figure-latex/liv-contigs-1}
+
+}
+
+\caption{First order queen contiguity of output areas (OAs) in Liverpool with magnified inset. }\label{fig:liv-contigs}
+\end{figure}
+
+To demonstrate how \texttt{sfislands} can be used to quickly compare the output of models, we show the output of three similar models.
+The first of these is a hierarchical model as described above, but with no ICAR component at the lowest level.
+Its spatially varying coefficients are shown in Figure \ref{fig:liv-est1}.
+Secondly, we fit the model described above using \texttt{st\_bridges()} to generate first-order queen contiguities as the structure for the neighbourhood relationships.
+The coefficients of this model are shown in Figure \ref{fig:liv-est2}.
+
+Finally, we can fit a similar model but this time using a within-distance criterion of 700m for neighbourhood status.
+All units within this distance of each other are considered neighbours.
+This was generated with the \texttt{sfdep:st\_dist\_band()} function.
+These coefficients are shown in Figure \ref{fig:liv-est3}.
+
+This shows how \texttt{sfislands} can quickly give an intuitive representation of how a set of similar models are operating.
+As a measure of the relative strengths of these three structures, we could compare their AICs (see Table \ref{tab:aic-latex}).
+These suggest that, in this scenario, inclusion of the ICAR component is appropriate, and queen contiguity is a more suitable choice than the distance-band condition for defining the neighbourhood structure.
 
@@
 \end{figure}
-
-If we instead use contiguity as the basis of an additional spatial term,
-we get estimates shown in Figure \ref{fig:liv-est2}.
 
@@
 \end{figure}
-
-If instead we consider units to be neighbours if they lie within a
-distance band of 700 metres, we get estimates shown in Figure
-\ref{fig:liv-est3}.
-
-A comparison of their AICs (see Table \ref{tab:aic-latex}) suggests that, in this scenario, queen contiguity is more appropriate than the distance band condition.
 
@@
 \hline
-model & AIC & relative AIC\\
+\textbf{model} & \textbf{AIC} & \textbf{relative AIC}\\
 \hline
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sf}{}}%
-Pebesma, Edzer. 2018. {``{Simple Features for R: Standardized Support for Spatial Vector Data}.''} \emph{{The R Journal}} 10 (1): 439--46. \url{https://doi.org/10.32614/RJ-2018-009}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-broom}{}}%
@@
 \url{https://doi.org/10.21105/joss.01686}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wikicrossings}{}}%
+Wikipedia.
+2024.
+{``{List of crossings of the River Thames} --- {W}ikipedia{,} the Free Encyclopedia.''} \url{http://en.wikipedia.org/w/index.php?title=List\%20of\%20crossings\%20of\%20the\%20River\%20Thames\&oldid=1184426738}.
 

### 2024-03-15 article.R

@@ -1,0 +1,1223 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit article.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
+
+# install.packages("devtools")
+devtools::install_github("horankev/sfislands")
+
+# required libraries
+library(tidyverse)
+library(kableExtra)
+library(sf)
+library(sfdep)
+library(sfislands)
+library(mgcv)
+library(here)
+library(ggsflabel)
+library(ggspatial)
+library(ggpubr)
+library(AER) # to test for overdispersion
+library(ggmagnify)
+
+# To scrape data
+library(rvest)
+library(httr)
+library(polite)
+library(janitor)
+
+# for geom_sf_label_repel consistency
+set.seed(123)
+
+# set consistent theme
+theme_set(
+  theme_bw() +
+    theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", linewidth=1.5), 
+          axis.title = element_blank(),
+          axis.ticks = element_blank())
+)
+
+
+
+## -----------------------------------------------------------------------------
+# load data
+
+# datasets are all available at https://github.com/horankev/quake_data 
+# or https://github.com/horankev/london_liverpool_data
+# alongside details on precisely how they were sourced etc
+
+# indonesia
+quakes_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/quakes_df.rds")))
+provinces_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/provinces_df.rds"))) |> 
+  rename(damaging_quakes_total = quake_mlxl_total,
+         damaging_quakes_density = quake_mlxl_density)
+counties_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/counties_df.rds")))
+faults_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/faults_df.rds")))
+nearby_countries_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/nearby_countries_df.rds")))
+indonesia_outline <- provinces_df |> summarise()
[…]

### 2024-03-15 article.tex
updated article

@@
 This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally related components.
-Here, we
-will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree. We will
-compare two different types of contiguity which could be applied to the ICAR component:
+Here, we will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree.
+We will compare two different types of contiguity which could be applied to the ICAR component:
 
@@
 
-Now suppose we wish to model unemployment as a function of limiting long-term
-illness, taking into account this hierarchical structure.
+Now suppose we wish to model unemployment as a function of limiting long-term illness, taking into account this hierarchical structure.
 Limiting long-term illness (LLTI) refers to a health problem or disability which limits a person's day-to-day activities (Barnett 2001).
@@
 
-There are no islands or discontiguities in this geometry of Liverpool so
-using \texttt{st\_bridges()} will give the same structure as would
-\texttt{sfdep::st\_contiguity()} (see Figure \ref{fig:liv-contigs}).
+There are no islands or discontiguities in this geometry of Liverpool so using \texttt{st\_bridges()} will give the same structure as would \texttt{sfdep::st\_contiguity()} (see Figure \ref{fig:liv-contigs}).
 

### 2024-03-21 article.Rmd
updated article
improved draft

@@
 abstract: >
-  Fitting areal spatial models can be a cumbersome task, particularly when the geographical units are not well-behaved.
-The presence of islands gives rise to particular issues when creating neighbourhood structures based on contiguity.
-We might also want to edit the output of a condition-based neighbourhood structure.
-It might be appropriate to add or remove connections to take account of infrastructure links or other domain-specific issues which are not apparent from a basic set of polygons.
-This package streamlines these processes with tools to account for the presence of islands and to intuitively make manual alterations to neighbourhood structures with the help of convenient mapping functions.
-Beyond the setting up of a satisfactory neighbourhood structure, it has further helper functions specific to the `mgcv` package which reduce the human workload required to extract estimates from models with spatial and hierarchical structures.
+  Fitting areal models which use a spatial weights matrix to represent relationships between geographical units can be a cumbersome task, particularly when these units are not well-behaved.
+The two chief aims of `sfislands` are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models.
+The package uses visual aids in the form of easily-generated maps to help this process.
+This paper demonstrates how `sfislands` could be useful to researchers.
+It begins by describing the package's functions in the context of a proposed workflow.
+It then presents three worked examples showing a selection of potential use-cases.
+These range from earthquakes in Indonesia, to river crossings in London, and hierarchical models of output areas in Liverpool.
+We aim to show how the `sfislands` package streamlines much of the human workflow involved in creating and examining such models.
 draft: true
@@
 For R users, the \CRANpkg{spdep} package [@spdep] has long been popular for the
-creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
+creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}  [@sf]
 structures and their compatibility with the \CRANpkg{tidyverse}  [@tidyverse], the \CRANpkg{sfdep} package [@sfdep] has presented the same
@@
 services might be satisfactory to meet our hypothesis of the required
-degree of connectivity to be considered neighbours. Again, such information will not be found in a set of polygons.
+degree of connectivity to be considered neighbours.
+Again, such information will may not be apparent from a basic set of polygons.
+In order to create what a researcher considers to be an appropriate neighbourhood structure, incorporating all of the domain knowledge that they might have about the system, it should be simple and intuitive to add and remove connections between spatial units.
+This might mean adding links to account for man-made infrastructure, or cutting links to incorporate natural barriers such as rivers or mountains.
 
@@
 convenient and open manner. It allows us to set up a structure, quickly
-map it, and then examine whether or not it is consistent with our hypothesis of relationships between units. The structure can then be edited and the process re-iterated until we have described a
+map it, and then examine whether or not we are happy with how it represents our hypothesis of relationships between units.
+The structure can then be edited and the process re-iterated until we have described a
 spatial relationship structure with which we are satisfied.
@@
 second category of features, which we refer to as *post-functions*,
-are for use with the \CRANpkg{mgcv} package [@mgcv].
-Having fit a model with \CRANpkg{mgcv}, the process of extracting estimates for certain types of effects can be somewhat awkward.
+are for use after the creation of a model.
+Having fit a model with \CRANpkg{mgcv}  [@mgcv] in particular, the process of extracting estimates for certain types of effects can be somewhat awkward.
 These *post-functions* augment the original dataframe with these estimates and their standard errors in tidy format.
@@
 Indonesia. It shows a scenario where all of the functions are used, from
-setting up contiguities, to modelling and examining the estimates of the
+setting up contiguities, to modelling and examining the predictions of the
 model.
@@
 The final example focuses on Liverpool. There are no islands or issues
-of discontiguity in this dataset. Instead, we will show how the package
-can be used to fit a multilevel model with an ICAR component at the lowest
-level. We show how the results can be presented in tidy form and quickly
-visualised. We also show how different types of contiguity from the
-\CRANpkg{sfdep} package can used within the \CRANpkg{mgcv} modelling
-framework and how the estimates derived from them can be conveniently
-compared.
+of discontiguity in this dataset.
+Instead, this example shows how the package can be used to quickly fit a set of reasonably complex multilevel models, with or without a spatially autocorrelated component at the lowest level, and how the results can be presented in tidy format and quickly visualised.
+This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the `sfislands` post-functions make it very straightforward to quickly compare the different predictions generated from different structures.
 
@@
 
-It greatly speeds up each stage of a proposed workflow for fitting areal spatial models.
+Below, we outline some of the benefits of the package in the context of a proposed workflow for fitting areal spatial models.
 
@@
 
-### Step 3: *Post-functions* for `mgcv` models
-
-  1. It simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models).
+### Step 3: *Post-functions* for models
+
+  1.
+It simplifies the process of extracting estimates from models, such as those with random effects and Markov random field structures created using \CRANpkg{mgcv}.
+Compatibility with more packages can be added at a future date.
 
@@
 
+One solution when confronted with islands in a dataset is to simply exclude them from the analysis.
 In the first two examples of using `st_bridges()`, we have chosen to ignore islands with the argument `remove_islands = TRUE` and to return a list and matrix structure respectively by specifying this in the `nb_structure` argument and choosing `add_to_dataframe = FALSE`:
@@
 
-The use of these neighbourhood structures is particularly associated with CAR-type models.
+Having decided upon an appropriate neighbourhood structure, the next step is to use this in the context of a model.
+The use of such structures is particularly associated with CAR-type (conditional autoregressive) models.
 These are often implemented in a Bayesian framework using
@@
 The `geometry` column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it.
-^[In a similar way, `st_augment()` can also be used to append the random effects from `lme4` and `nlme` models to an `sf` dataframe, which can then be easily mapped using `st_quickmap_preds()`.] The spatially varying predictions which `st_augment()` extracts from an \CRANpkg{mgcv} model are
+^[In a similar way, `st_augment()` can also be used to append the random effects from \CRANpkg{lme4}   [@lme4] and \CRANpkg{nlme}   [@nlme] models to an `sf` dataframe, which can then be easily mapped using `st_quickmap_preds()`.
+Compatability with models created using different packages can be introduced in the future.] The spatially varying predictions which `st_augment()` extracts from an \CRANpkg{mgcv} model are
 
@@
 Each column name begins with either `random.effect.` or `mrf.smooth.` as appropriate.
-An additional column is also added for the standard error of each estimate, as calculated by \CRANpkg{mgcv}.
+An additional column is also added for the standard error of each prediction, as calculated by \CRANpkg{mgcv}.
 These columns are named as above but with `se.` prepended (e.g.
@@
 As it is possible to include more than 1 spatially varying component, the output of this function is a list of plots.
-They can be viewed individually by indexing, or all at one using, for example, the `plotlist` argument from the `ggarrange()` function which is part of the \CRANpkg{ggpubr}  [@ggpubr] package.
+They can be viewed individually by indexing, or all at once using, for example, the `plotlist` argument from the `ggarrange()` function which is part of the \CRANpkg{ggpubr}  [@ggpubr] package.
 We will see this function in practice in the following examples.
@@
 
-As this is count data, we will model it as a poisson distribution with
+As this is count data, we will model it as a Poisson distribution with
 $\lambda$ as the mean count per province. For $i = 1,...,n$ provinces,
@@
 Firstly, when excluding the incidence and just modelling counts, where
-$y_i = \text{earthquake count in province}_i$, the poisson model is of
+$y_i = \text{earthquake count in province}_i$, the Poisson model is of
 the following form: 
@@
 considered neighbours if they share at least one point of boundary.
-This can be modelled as a Markov random field to generate an ICAR model with a spatially varying term.
+This can be modelled as a Markov random field to generate an intrinsic CAR (ICAR) model with a spatially varying term.
 Each of these terms will be correlated with the others according to the neighbourhood structure we have defined.
@@
 
-Such models, however, can not deal with locations which have no
+Such models, however, can not incorporate locations which have no
 neighbours. In the case of Indonesia, this is quite problematic. It is
@@
 
-```{r, ind-contig2, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption2)"}
+```{r, ind-contig2, fig.width=12, fig.height=6, echo=FALSE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption2)"}
 # with 'concavehull = TRUE' and 'nodes = "numeric"'
@@
 
-Having made some manual adjustments to the neighbourhood structure, the
+Having made some manual adjustments to the neighbourhood structure, outlined in the code below, the
 new structure can be seen in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig3html', 'ind-contig3')`). Edge effects
@@
                       fault_concentration +
-                      s(province, bs='mrf', xt=list(nb=prep_data$nb), k=22) +
+                      s(province, bs='mrf', xt=list(nb=prep_data$nb), k=24) +
                       offset(log(area_province)),
@@
 
-We can see from the summary below that the adjusted R-squared is **0.957** and deviance explained is **88.2%**.
+We can see from the summary below that the adjusted R-squared is **0.983** and deviance explained is **93.3%**.
+The coefficient for `fault_concentration` confirms an expected positive mean global association between earthquake and fault incidence.
 
@@
 
-Controlling for the concentration of faults, what is the additional risk level of
-earthquakes in a province? This can be seen as a measure of the activity
-level of faults locally and it is spatially smoothed by the
-autoregressive process.
+Regardless of this, having controlled for the concentration of faults, what is the additional risk level of earthquakes in a province?
+This can be seen as a measure of the activity level of faults locally and it is spatially smoothed by the autoregressive process.
 It is represented in the model summary by the component `s(province)`.
 However, the extraction of individual predictions for this component for each province from the `mgcv` model requires a number of steps.
-We know demonstrate how these are streamlined into a single function by the `sfislands` package.
+We now demonstrate how these are streamlined into a single function by the `sfislands` package.
 
@@
 
-If we wish to apply the inverse link function (the exponential function in the case of this poisson model) to map these values to a more interpretable scale, this will not be generated by the function `st_quickmap_preds()`.
+If we wish to apply the inverse link function (the exponential function in the case of this Poisson model) to map these values to a more interpretable scale, this will not be generated by the function `st_quickmap_preds()`.
 Instead, we must use the augmented dataframe which is produced by `st_augment()` and create the appropriate extra column with the usual \CRANpkg{tidyverse} `mutate()` function.
@@
 We can see that, controlling for the effects of proximity to
-faults, the province of Aceh has 7.9 times the expected
+faults, the province of Nusa Tenggara Barat has 8.7 times the expected
 incidence, or number of major earthquakes per square kilometre. The two
@@
 visual representation of the structure.
-^[`st_quickmap_nb()` can also be used to visualise any contiguity structure created by `spdep` or `sfdep` as long as that structure is included in an `sf` dataframe as a column named `nb`.] This can be supplemented by
-adding a layer showing the course of the river Thames which is also visible in Figure \@ref(fig:lon-contig1).
+^[`st_quickmap_nb()` can also be used to visualise any contiguity structure created by `spdep` or `sfdep` as long as that structure is included in an `sf` dataframe as a column named `nb`.] Because this map is created using \CRANpkg{ggplot2}, it can be easily supplemented by adding a layer showing the course of the river Thames which is also visible in Figure \@ref(fig:lon-contig1).
 
@@
 
-In order to take account of actual connectivity, we can add a layer
-showing the road and pedestrian bridges or tunnels.
-Details of these were sourced @wikicrossings, which describes the various crossings of the river Thames.
+In order to take account of actual connectivity, we can add a layer showing the road and pedestrian bridges or tunnels.
+Details of these were sourced from the @wikicrossings article titled "*List of crossings of the River Thames*".
 In Figure \@ref(fig:lon-crossings), we have also drawn a 1 kilometre buffer around each crossing.
@@
 The vast majority of units on opposing banks have access to a river crossing within this threshold and thus should be considered as neighbours.
-Only the extreme eastern units and one to the south west should not have a connection across the river.
+Only the extreme eastern units and one to the south west should not have a connection across the river according to this criterion.
 
@@
 
-Now suppose we wish to model unemployment as a function of limiting long-term
-illness, taking into account this hierarchical structure.
+Now suppose we wish to model unemployment as a function of limiting long-term illness, taking into account this hierarchical structure.
 Limiting long-term illness (LLTI) refers to a health problem or disability which limits a person's day-to-day activities [@Barnett2001].
-We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at MSOA and LSOA levels.
+We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at different levels.
 The raw distribution of these variables across the output areas (OAs) of Liverpool are shown in Figure
@@
 
-There are no islands or discontiguities in this geometry of Liverpool so
-using `st_bridges()` will give the same structure as would
-`sfdep::st_contiguity()` (see Figure \@ref(fig:liv-contigs)).
+There are no islands or discontiguities in this geometry of Liverpool so using `st_bridges()` will give the same structure as would `sfdep::st_contiguity()` (see Figure \@ref(fig:liv-contigs)).
 
@@
 Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included.
-It also provides helper functions to use these structures in spatial regression models built with `mgcv` which streamline the human effort necessary to examine the estimates.
+It also provides helper functions to use these structures in spatial regression models built with \CRANpkg{mgcv} which streamline the human effort necessary to examine the estimates.
+In future, compatibility with other modelling packages can be added to broaden the package's capabilities.
 

### 2024-03-21 article.tex
improved draft

@@
 \abstract{%
-Fitting areal spatial models can be a cumbersome task, particularly when the geographical units are not well-behaved.
-The presence of islands gives rise to particular issues when creating neighbourhood structures based on contiguity.
-We might also want to edit the output of a condition-based neighbourhood structure.
-It might be appropriate to add or remove connections to take account of infrastructure links or other domain-specific issues which are not apparent from a basic set of polygons.
-This package streamlines these processes with tools to account for the presence of islands and to intuitively make manual alterations to neighbourhood structures with the help of convenient mapping functions.
-Beyond the setting up of a satisfactory neighbourhood structure, it has further helper functions specific to the \texttt{mgcv} package which reduce the human workload required to extract estimates from models with spatial and hierarchical structures.
+Fitting areal models which use a spatial weights matrix to represent relationships between geographical units can be a cumbersome task, particularly when these units are not well-behaved.
+The two chief aims of \texttt{sfislands} are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models.
+The package uses visual aids in the form of easily-generated maps to help this process.
+This paper demonstrates how \texttt{sfislands} could be useful to researchers.
+It begins by describing the package's functions in the context of a proposed workflow.
+It then presents three worked examples showing a selection of potential use-cases.
+These range from earthquakes in Indonesia, to river crossings in London, and hierarchical models of output areas in Liverpool.
+We aim to show how the \texttt{sfislands} package streamlines much of the human workflow involved in creating and examining such models.
 }
@@
 For R users, the \CRANpkg{spdep} package (Bivand et al. 2005) has long been popular for the
-creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}
+creation of these matrices. More recently, with the increasing use of \CRANpkg{sf} (Pebesma 2018)
 structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} package (Parry and Locke 2024) has presented the same
@@
 second category of features, which we refer to as \emph{post-functions},
-are for use with the \CRANpkg{mgcv} package (Wood 2011).
-Having fit a model with \CRANpkg{mgcv}, the process of extracting estimates for certain types of effects can be somewhat awkward.
+are for use after the creation of a model.
+Having fit a model with \CRANpkg{mgcv} (Wood 2011) in particular, the process of extracting estimates for certain types of effects can be somewhat awkward.
 These \emph{post-functions} augment the original dataframe with these estimates and their standard errors in tidy format.
@@
 The final example focuses on Liverpool. There are no islands or issues
-of discontiguity in this dataset. Instead, we will show how the package
-can be used to fit a multilevel model with an ICAR component at the lowest
-level. We show how the results can be presented in tidy form and quickly
-visualised. We also show how different types of contiguity from the
-\CRANpkg{sfdep} package can used within the \CRANpkg{mgcv} modelling
-framework and how the estimates derived from them can be conveniently
-compared.
+of discontiguity in this dataset.
+Instead, this example shows how the package can be used to quickly fit a set of reasonably complex multilevel models, with or without a spatially autocorrelated component at the lowest level, and how the results can be presented in tidy format and quickly visualised.
+This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the \texttt{sfislands} post-functions make it very straightforward to quickly compare the different predictions generated from different structures.
 
@@
 
-\hypertarget{step-3-post-functions-for-mgcv-models}{%
-\subsubsection{\texorpdfstring{Step 3: \emph{Post-functions} for \texttt{mgcv} models}{Step 3: Post-functions for mgcv models}}\label{step-3-post-functions-for-mgcv-models}}
+\hypertarget{step-3-post-functions-for-models}{%
+\subsubsection{\texorpdfstring{Step 3: \emph{Post-functions} for models}{Step 3: Post-functions for models}}\label{step-3-post-functions-for-models}}
 
@@
 \item
-  It simplifies the process of extracting estimates from models with random effects and Markov random field structures (ICAR models).
+  It simplifies the process of extracting estimates from models, such as those with random effects and Markov random field structures created using \CRANpkg{mgcv}.
+Compatibility with more packages can be added at a future date.
 \item
@@
 
+One solution when confronted with islands in a dataset is to simply exclude them from the analysis.
 In the first two examples of using \texttt{st\_bridges()}, we have chosen to ignore islands with the argument \texttt{remove\_islands\ =\ TRUE} and to return a list and matrix structure respectively by specifying this in the \texttt{nb\_structure} argument and choosing \texttt{add\_to\_dataframe\ =\ FALSE}:
@@
 The \texttt{geometry} column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it.
-\footnote{In a similar way, \texttt{st\_augment()} can also be used to append the random effects from \texttt{lme4} and \texttt{nlme} models to an \texttt{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}.} The spatially varying predictions which \texttt{st\_augment()} extracts from an \CRANpkg{mgcv} model are
+\footnote{In a similar way, \texttt{st\_augment()} can also be used to append the random effects from \CRANpkg{lme4} (Bates et al.
+2015) and \CRANpkg{nlme} (Pinheiro, Bates, and R Core Team 2023) models to an \texttt{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}.
+Compatability with models created using different packages can be introduced in the future.} The spatially varying predictions which \texttt{st\_augment()} extracts from an \CRANpkg{mgcv} model are
 
@@
 Each column name begins with either \texttt{random.effect.} or \texttt{mrf.smooth.} as appropriate.
-An additional column is also added for the standard error of each estimate, as calculated by \CRANpkg{mgcv}.
+An additional column is also added for the standard error of each prediction, as calculated by \CRANpkg{mgcv}.
 These columns are named as above but with \texttt{se.} prepended (e.g.~\texttt{se.random.effect.region}).
@@
 As it is possible to include more than 1 spatially varying component, the output of this function is a list of plots.
-They can be viewed individually by indexing, or all at one using, for example, the \texttt{plotlist} argument from the \texttt{ggarrange()} function which is part of the \CRANpkg{ggpubr} (Kassambara 2023) package.
+They can be viewed individually by indexing, or all at once using, for example, the \texttt{plotlist} argument from the \texttt{ggarrange()} function which is part of the \CRANpkg{ggpubr} (Kassambara 2023) package.
 We will see this function in practice in the following examples.
@@
 
-As this is count data, we will model it as a poisson distribution with
+As this is count data, we will model it as a Poisson distribution with
 \(\lambda\) as the mean count per province. For \(i = 1,...,n\) provinces,
@@
 Firstly, when excluding the incidence and just modelling counts, where
-\(y_i = \text{earthquake count in province}_i\), the poisson model is of
+\(y_i = \text{earthquake count in province}_i\), the Poisson model is of
 the following form:
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ind-contig2-1}
+
+}
+
+\caption{Neighbourhood structure for Indonesian provinces.
+Viewed with \texttt{st\_quickmap\_nb()}, using the arguments \texttt{nodes\ =\ \textquotesingle{}numeric\textquotesingle{}} and \texttt{concavehull\ =\ TRUE}.}\label{fig:ind-contig2}
+\end{figure}
+
 \begin{verbatim}
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ind-contig2-1}
-
-}
-
-\caption{Neighbourhood structure for Indonesian provinces.
-Viewed with \texttt{st\_quickmap\_nb()}, using the arguments \texttt{nodes\ =\ \textquotesingle{}numeric\textquotesingle{}} and \texttt{concavehull\ =\ TRUE}.}\label{fig:ind-contig2}
-\end{figure}
-
-\begin{verbatim}
-# with 'concavehull = TRUE' and 'nodes = "numeric"'
-
-st_bridges(provinces_df, "province",link_islands_k = 2) |> 
-  st_quickmap_nb(fillcol = "antiquewhite1", 
-                 bordercol = "black", bordersize = 0.5, 
-                 linkcol = "tomato", linksize = 0.5, 
-                 nodes = "numeric", 
-                 numericcol = "black", numericsize = 6, 
-                 concavehull = TRUE, 
-                 hullcol = "darkgreen", hullsize = 0.2) + 
-  theme(panel.background = element_rect(fill = "#ECF6F7", colour = "black", 
-                                        linewidth=1.5),
-        axis.text = element_blank())
-\end{verbatim}
-
-Having made some manual adjustments to the neighbourhood structure, the
+Having made some manual adjustments to the neighbourhood structure, outlined in the code below, the
 new structure can be seen in Figure \ref{fig:ind-contig3}. Edge effects
@@
 
-We can see from the summary below that the adjusted R-squared is \textbf{0.957} and deviance explained is \textbf{88.2\%}.
+We can see from the summary below that the adjusted R-squared is \textbf{0.983} and deviance explained is \textbf{93.3\%}.
+The coefficient for \texttt{fault\_concentration} confirms an expected positive mean global association between earthquake and fault incidence.
 
@@
 #> damaging_quakes_total ~ fault_concentration + s(province, bs = "mrf", 
-#>     xt = list(nb = prep_data$nb), k = 22) + offset(log(area_province))
+#>     xt = list(nb = prep_data$nb), k = 24) + offset(log(area_province))
 #> 
@@
 #>                     Estimate Std. Error z value Pr(>|z|)    
-#> (Intercept)          -9.1066     0.1428 -63.774   <2e-16 ***
-#> fault_concentration  -3.2355     1.3297  -2.433    0.015 *  
+#> (Intercept)          -9.5648     0.1744 -54.845  < 2e-16 ***
+#> fault_concentration   5.9971     1.9245   3.116  0.00183 ** 
 #> ---
@@
 #>               edf Ref.df Chi.sq p-value    
-#> s(province) 18.94     21  147.7  <2e-16 ***
+#> s(province) 19.19     23  166.6  <2e-16 ***
 #> ---
@@
 #> 
-#> R-sq.(adj) =  0.957   Deviance explained = 88.2%
-#> -REML = 119.75  Scale est. = 1         n = 33
-\end{verbatim}
-
-Controlling for the concentration of faults, what is the additional risk level of
-earthquakes in a province? This can be seen as a measure of the activity
-level of faults locally and it is spatially smoothed by the
-autoregressive process.
+#> R-sq.(adj) =  0.983   Deviance explained = 93.3%
+#> -REML = 104.81  Scale est. = 1         n = 33
+\end{verbatim}
+
+Regardless of this, having controlled for the concentration of faults, what is the additional risk level of earthquakes in a province?
+This can be seen as a measure of the activity level of faults locally and it is spatially smoothed by the autoregressive process.
 It is represented in the model summary by the component \texttt{s(province)}.
 However, the extraction of individual predictions for this component for each province from the \texttt{mgcv} model requires a number of steps.
-We know demonstrate how these are streamlined into a single function by the \texttt{sfislands} package.
+We now demonstrate how these are streamlined into a single function by the \texttt{sfislands} package.
 
@@
 
-If we wish to apply the inverse link function (the exponential function in the case of this poisson model) to map these values to a more interpretable scale, this will not be generated by the function \texttt{st\_quickmap\_preds()}.
+If we wish to apply the inverse link function (the exponential function in the case of this Poisson model) to map these values to a more interpretable scale, this will not be generated by the function \texttt{st\_quickmap\_preds()}.
 Instead, we must use the augmented dataframe which is produced by \texttt{st\_augment()} and create the appropriate extra column with the usual \CRANpkg{tidyverse} \texttt{mutate()} function.
@@
 visual representation of the structure.
-\footnote{\texttt{st\_quickmap\_nb()} can also be used to visualise any contiguity structure created by \texttt{spdep} or \texttt{sfdep} as long as that structure is included in an \texttt{sf} dataframe as a column named \texttt{nb}.} This can be supplemented by
-adding a layer showing the course of the river Thames which is also visible in Figure \ref{fig:lon-contig1}.
+\footnote{\texttt{st\_quickmap\_nb()} can also be used to visualise any contiguity structure created by \texttt{spdep} or \texttt{sfdep} as long as that structure is included in an \texttt{sf} dataframe as a column named \texttt{nb}.} Because this map is created using \CRANpkg{ggplot2}, it can be easily supplemented by adding a layer showing the course of the river Thames which is also visible in Figure \ref{fig:lon-contig1}.
 
@@
 
-In order to take account of actual connectivity, we can add a layer
-showing the road and pedestrian bridges or tunnels.
-Details of these were sourced Wikipedia (2024), which describes the various crossings of the river Thames.
+In order to take account of actual connectivity, we can add a layer showing the road and pedestrian bridges or tunnels.
+Details of these were sourced from the Wikipedia (2024) article titled ``\emph{List of crossings of the River Thames}''.
 In Figure \ref{fig:lon-crossings}, we have also drawn a 1 kilometre buffer around each crossing.
@@
 Limiting long-term illness (LLTI) refers to a health problem or disability which limits a person's day-to-day activities (Barnett 2001).
-We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at MSOA and LSOA levels.
+We wish to incorporate a fixed intercept and coefficient for LLTI, with additional random intercepts and slopes at different levels.
 The raw distribution of these variables across the output areas (OAs) of Liverpool are shown in Figure
@@
 Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included.
-It also provides helper functions to use these structures in spatial regression models built with \texttt{mgcv} which streamline the human effort necessary to examine the estimates.
+It also provides helper functions to use these structures in spatial regression models built with \CRANpkg{mgcv} which streamline the human effort necessary to examine the estimates.
+In future, compatibility with other modelling packages can be added to broaden the package's capabilities.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-sf}{}}%
+Pebesma, Edzer. 2018. {``{Simple Features for R: Standardized Support for Spatial Vector Data}.''} \emph{{The R Journal}} 10 (1): 439--46. \url{https://doi.org/10.32614/RJ-2018-009}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
+Pinheiro, José, Douglas Bates, and R Core Team. 2023. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}. \url{https://CRAN.R-project.org/package=nlme}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-broom}{}}%

### 2024-03-21 article.R

@@ -423,7 +423,7 @@ st_bridges(provinces_df, "province",link_islands_k = 2) |>
 #> 
 
 
-## ---- ind-contig2, fig.width=12, fig.height=6, echo=TRUE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption2)"----
+## ---- ind-contig2, fig.width=12, fig.height=6, echo=FALSE, out.width="100%", eval=knitr::is_latex_output(), fig.cap = "(ref:my-caption2)"----
 # with 'concavehull = TRUE' and 'nodes = "numeric"'
 
 st_bridges(provinces_df, "province",link_islands_k = 2) |> 
@@ -551,7 +551,7 @@ prep_data <- st_bridges(provinces_df, "province",link_islands_k = 2) |>
 
 mod_pois_mrf <- gam(damaging_quakes_total ~ 
                       fault_concentration +
-                      s(province, bs='mrf', xt=list(nb=prep_data$nb), k=22) +
+                      s(province, bs='mrf', xt=list(nb=prep_data$nb), k=24) +
                       offset(log(area_province)),
                     data=prep_data, method="REML",family = "poisson")
 

### 2024-03-28 article.Rmd
edited article

@@
 title: "sfislands: An R Package for Accommodating Islands and Disjoint Zones in Areal Spatial Modelling"
-date: "2022-11-09"
+date: "2024-03-28"
 abstract: >
   Fitting areal models which use a spatial weights matrix to represent relationships between geographical units can be a cumbersome task, particularly when these units are not well-behaved.
-The two chief aims of `sfislands` are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models.
+The two chief aims of \CRANpkg{sfislands} are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models.
 The package uses visual aids in the form of easily-generated maps to help this process.
-This paper demonstrates how `sfislands` could be useful to researchers.
+This paper demonstrates how \CRANpkg{sfislands} could be useful to researchers.
 It begins by describing the package's functions in the context of a proposed workflow.
@@
 These range from earthquakes in Indonesia, to river crossings in London, and hierarchical models of output areas in Liverpool.
-We aim to show how the `sfislands` package streamlines much of the human workflow involved in creating and examining such models.
+We aim to show how the \CRANpkg{sfislands} package streamlines much of the human workflow involved in creating and examining such models.
 draft: true
@@
 # install.packages("devtools")
-devtools::install_github("horankev/sfislands")
+# devtools::install_github("horankev/sfislands")
 
@@
 represented numerically in a spatial weights matrix. In order to build
-this, we must first decide on what constitutes being a neighbour. We can
+this, we must first decide on what constitutes being a neighbour. We might
 see this as a continuous relationship where degree of neighbourliness is
@@
 For R users, the \CRANpkg{spdep} package [@spdep] has long been popular for the
-creation of these matrices. More recently, with the increasing use of \CRANpkg{sf}  [@sf]
-structures and their compatibility with the \CRANpkg{tidyverse}  [@tidyverse], the \CRANpkg{sfdep} package [@sfdep] has presented the same
+creation of these matrices.
+More recently, with the increasing use of \CRANpkg{sf} structures  [@sf] and their compatibility with the \CRANpkg{tidyverse}  [@tidyverse], the \CRANpkg{sfdep} package [@sfdep] has presented the same
 functionality in addition to extra features in a tidy structure based on
@@
 degree of connectivity to be considered neighbours.
-Again, such information will may not be apparent from a basic set of polygons.
+Again, such information may not be apparent from a basic set of polygons.
 In order to create what a researcher considers to be an appropriate neighbourhood structure, incorporating all of the domain knowledge that they might have about the system, it should be simple and intuitive to add and remove connections between spatial units.
@@
 
-The aim of `sfislands` is to deal with the situations described above in a
+The aim of \CRANpkg{sfislands}  [@sfislands] is to deal with the situations described above in a
 convenient and open manner. It allows us to set up a structure, quickly
@@
 In this paper, we will look at three examples to show different use-cases for
-`sfislands`. The first example focuses on earthquakes in
+\CRANpkg{sfislands}. The first example focuses on earthquakes in
 Indonesia. It shows a scenario where all of the functions are used, from
@@
 
-The second example looks at London and how, despite the lack of islands,
+The second example looks at London and how, despite an absence of islands,
 the presence of a river means that some of the pre-functions of
-`sfislands` can be useful.
+\CRANpkg{sfislands} can be useful.
 
@@
 Instead, this example shows how the package can be used to quickly fit a set of reasonably complex multilevel models, with or without a spatially autocorrelated component at the lowest level, and how the results can be presented in tidy format and quickly visualised.
-This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the `sfislands` post-functions make it very straightforward to quickly compare the different predictions generated from different structures.
-
-# Why use `sfislands`?
+This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the \CRANpkg{sfislands} post-functions make it very straightforward to quickly compare the different predictions generated from different structures.
+
+# Why use \CRANpkg{sfislands}?
 
@@
   1.
-It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data.
-`sfislands` will include a contiguity for all units.
+It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure contains empty records.
+\CRANpkg{sfislands} will include a contiguity for all units.
 
@@
 
-  3. These maps are created using \CRANpkg{ggplot2}  [@ggplot2], which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax.
+  3. These maps are created using \CRANpkg{ggplot2}  [@ggplot2], which allows users to apply additional styling and themes using \CRANpkg{ggplot2} syntax.
 
@@
 
-  5. Connections which have been induced by a function from the package but are not contiguities can be accessed to ensure openness in the process.
+  5. Connections which have been induced by a function from the package but which are not based on geographical contiguity can be accessed to ensure openness in the process.
 
@@
 These structures can serve as the input to models in \CRANpkg{brms},
-\pkg{r-inla}, \pkg{rstan}  [@rstan] or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
+\pkg{r-inla}, \CRANpkg{rstan}  [@rstan] or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
 object, it is included in the original \CRANpkg{sf} dataframe as a named list or matrix, in the spirit of the \CRANpkg{sfdep} package.
@@
 Having decided upon an appropriate neighbourhood structure, the next step is to use this in the context of a model.
-The use of such structures is particularly associated with CAR-type (conditional autoregressive) models.
+The use of such structures is particularly associated with CAR (conditional autoregressive) or ICAR-type (intrinsic conditional autoregressive) models [@besag].
 These are often implemented in a Bayesian framework using
-\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
+\CRANpkg{brms}, \pkg{r-inla} or \CRANpkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
 neighbourhood structure in the desired format for use in any of these
@@
 any other component of the model, it can be somewhat awkward to get the
-estimates from \CRANpkg{mgcv} model. `sfislands` has two post-functions to conveniently extract and visualise these.
+estimates from \CRANpkg{mgcv} models. \CRANpkg{sfislands} has two post-functions to conveniently extract and visualise these.
 
@@
 The `geometry` column, as per convention, remains as the last column of the augmented dataframe, while the predictions are positioned immediately before it.
-^[In a similar way, `st_augment()` can also be used to append the random effects from \CRANpkg{lme4}   [@lme4] and \CRANpkg{nlme}   [@nlme] models to an `sf` dataframe, which can then be easily mapped using `st_quickmap_preds()`.
+^[In a similar way, `st_augment()` can also be used to append the random effects from \CRANpkg{lme4}   [@lme4] and \CRANpkg{nlme}   [@nlme] models to an \CRANpkg{sf} dataframe, which can then be easily mapped using `st_quickmap_preds()`.
 Compatability with models created using different packages can be introduced in the future.] The spatially varying predictions which `st_augment()` extracts from an \CRANpkg{mgcv} model are
@@
 Consider the model structure described in the code below using \CRANpkg{mgcv} syntax.
-In this model $y$ is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a *region* level, and a set of ICAR varying intercepts and slopes at a lower *sub-region* level.
+In this model $y$ is the dependent variable which is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a *region* level, and a set of ICAR varying intercepts and slopes at a lower *sub-region* level.
 
@@
 
-# Example 1: Indonesia
+# Indonesia (example 1)
 
@@
 quakes_df_temp <- quakes_df |> 
-  mutate(magfact = case_when(magfact == "M" ~ "M: 5.5 - 6.1",
-                             magfact == "L" ~ "L: 6.1 - 7",
+  mutate(magfact = case_when(magfact == "M" ~ "M: 5.5 - 6",
+                             magfact == "L" ~ "L: 6.1 - 6.9",
                              magfact == "XL" ~ "XL: 7+")) |> 
   mutate(magfact = factor(magfact,
-                          levels = c("M: 5.5 - 6.1",
-                                     "L: 6.1 - 7",
+                          levels = c("M: 5.5 - 6",
+                                     "L: 6.1 - 6.9",
                                      "XL: 7+")))
@@
 considered neighbours if they share at least one point of boundary.
-This can be modelled as a Markov random field to generate an intrinsic CAR (ICAR) model with a spatially varying term.
+This can be modelled as a Markov random field to generate an ICAR model with a spatially varying term.
 Each of these terms will be correlated with the others according to the neighbourhood structure we have defined.
@@
 
-Having described the type of model we wish to implement, we now show how `sfislands` can be used to streamline the process.
+Having described the type of model we wish to implement, we now show how \CRANpkg{sfislands} can be used to streamline the process.
 
@@
 is shown in Figure \@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-contig1html', 'ind-contig1')`).
+Note how is can be styled with a combination of internal arguments (size, colour, fill etc.) and additional \CRANpkg{ggplot2} layers.
 
@@
 
-Regardless of this, having controlled for the concentration of faults, what is the additional risk level of earthquakes in a province?
+Returning to the initial question, what is the additional risk level of earthquakes in a province, having controlled for the concentration of faults?
 This can be seen as a measure of the activity level of faults locally and it is spatially smoothed by the autoregressive process.
@@
 However, the extraction of individual predictions for this component for each province from the `mgcv` model requires a number of steps.
-We now demonstrate how these are streamlined into a single function by the `sfislands` package.
+We now demonstrate how these are streamlined into a single function by the \CRANpkg{sfislands} package.
 
@@
 similar to that of the \CRANpkg{broom} package.
-For instance, we see from the code below that the original dataframe is now augmented with columns called "mrf.smooth.province" and "se.mrf.smooth.province" which show the predictions for the $\gamma_i$ component and their standard errors.
+For instance, we see from the output of the following code chunk that the original dataframe is now augmented with columns called `mrf.smooth.province` and `se.mrf.smooth.province` which show the predictions for the $\gamma_i$ component and their standard errors.
 Note that this is how we would expect them to be named, based on the previous discussion surrounding Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:staugtab-html)', '\\@ref(tab:staugtab-latex)'))`.
-They are positioned immediately before the final `geometry` column of the `sf` dataframe.
+They are positioned immediately before the final `geometry` column of the `sf` dataframe, and after the neighbours list column, `nb`.
 
@@
 # so we call it by index
+# it is then supplemented with additional ggplot functions
 
@@
 This allows us to produce the map in Figure
-\@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-mrf2html', 'ind-mrf2')`). As these coefficients are multiplicatively related
-to the earthquake incidence, values below 1 imply an earthquake
-incidence which is lower than expected.
-
-The provinces with the 3 most elevated incidences are labelled in red
-We can see that, controlling for the effects of proximity to
-faults, the province of Nusa Tenggara Barat has 8.7 times the expected
-incidence, or number of major earthquakes per square kilometre. The two
-lowest-scoring provinces, labelled in green, have essentially no incidence of
-earthquake epicentres within their boundaries, controlling for what their proximity to faults alone would
-suggest.
+\@ref(fig:`r ifelse(knitr::is_html_output(), 'ind-mrf2html', 'ind-mrf2')`).
+As these coefficients are multiplicatively related to the earthquake incidence, values below 1 imply an earthquake incidence which is lower than expected.
+
+The provinces with the 3 most elevated incidences are labelled in red.
+We can see that, controlling for the effects of proximity to faults, the province of Nusa Tenggara Barat has 8.7 times the expected incidence, or number of major earthquakes per square kilometre.
+The two lowest-scoring provinces, labelled in green, have essentially no incidence of earthquake epicentres within their boundaries, controlling for what their proximity to faults alone would suggest.
 
@@
 However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps.
-A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with `sfislands` using nothing more than three or four lines of code, such as the following:
+A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with \CRANpkg{sfislands} using nothing more than three or four lines of code, such as the following:
 
@@
 
-# Example 2: London
-
-The next example looks only at using the *pre-functions* of `sfislands`, but in a situation where the presence of actual *islands* is not the problem we seek to deal with.
-Consider the wards and boroughs of London (sourced from the Greater London Authority's [London Datastore](https://data.london.gov.uk/)).
-In Figure \@ref(fig:lon-contig1) the `st_bridges()`
+# London (example 2)
+
+The next example looks only at using the *pre-functions* of \CRANpkg{sfislands}, but in a situation where the presence of actual *islands* is not the problem we seek to deal with.
+Consider the wards and boroughs of London (sourced from the Greater London Authority's [London Datastore](https://data.london.gov.uk/)) and available at
+<https://github.com/horankev/london_liverpool_data>. In Figure \@ref(fig:lon-contig1) the `st_bridges()`
 function is applied to them to construct a queen contiguity
@@
 geometries are defined, the presence of a river can cause problems in two ways.
-In one situation, the river could be expressed as a polygon in its own right meaning that it severs any potential contiguity between units on either side of its banks.
+In one situation, the river could be expressed as a polygon in its own right meaning that, using the condition of queen contiguity, it severs any potential contiguity between units on either side of its banks.
 In this situation, no spatial units will be neighbours with the units directly across the river from them.
-At the other extreme, if the river is not included as a geometry (as is the case here) all units on opposing banks are considered neighbours.
+At the other extreme, if the river is not included as a geometry (as is the case here) all units on opposing banks are automatically considered neighbours.
 
@@
 If there is no means of crossing the river within a reasonable distance, somebody living on the banks of a river might be more likely to go about their life primarily on their side of the river, despite the short distance as the crow flies of facilities on the other side.
-This could be relevant in terms of, say, modelling of house prices where we might want to incorporate these issues into a neighbourhood structure.
-
-`sfislands` provides convenient functions for this sort of situation.
-Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames (Figure \@ref(fig:lon-contig2)).
+This could be relevant in terms of, say, modelling of house prices where we might want to incorporate issues such as local amenities into a neighbourhood structure.
+
+\CRANpkg{sfislands} provides convenient functions for this sort of situation.
+Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames.
+Figure \@ref(fig:lon-contig2) shows the resultant contiguities when the river is ignored.
 
@@
 accepts names as arguments which may be more convenient in some circumstances.] Having made these adjustments,
-`st_quickmap_nb()` now shows a connectivity structure (Figure \@ref(fig:lon-contig4)) which reflects how we have hypothesised that influence should extend across the river in the presence or absence of crossings.
-
-This example shows that the pre-functions of `sfislands` have uses for situations which do not involve islands.
+`st_quickmap_nb()` now shows a connectivity structure (Figure \@ref(fig:lon-contig4)) which reflects our hypothesis of how influence should extend across the river in the presence or absence of crossings.
+
+This example shows that the pre-functions of \CRANpkg{sfislands} have uses for situations which do not involve islands.
 They can be used to apply domain knowledge to easily design the most appropriate neighbourhood structure.
@@
 
-# Example 3: Liverpool
+# Liverpool (example 3)
 
 In this final example, there are no islands and no river issues.
-We saw in the Indonesia example how `sfislands` can extract estimates from the `mgcv` argument `bs="mrf"`.
+We saw in the Indonesia example how \CRANpkg{sfislands} can extract estimates from the `mgcv` argument `bs="mrf"`.
 It can also deal with random effects as defined by `bs="re"` in `mgcv` models.
-This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally related components.
+This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally correlated components.
 Here, we will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree.
@@
 
-* `st_bridges()` from `sfislands`, which implements first-order queen contiguity in the same way as `sfdep:st_contiguity()` for situations such as this where there are no islands, and 
+* `st_bridges()` from \CRANpkg{sfislands}, which implements first-order queen contiguity in the same way as `sfdep:st_contiguity()` for situations such as this where there are no islands, and 
 * `st_dist_band()` from \CRANpkg{sfdep}, which connects units whose centroids lie within a chosen distance threshold.
@@
 
-There are no islands or discontiguities in this geometry of Liverpool so using `st_bridges()` will give the same structure as would `sfdep::st_contiguity()` (see Figure \@ref(fig:liv-contigs)).
+There are no islands or discontiguities in this geometry of Liverpool so using `st_bridges()` will give the same structure as would `sfdep::st_contiguity()` (see Figure \@ref(fig:liv-contigs), with a magnified section for visibility purposes, created using \pkg{ggmagnify}  [@ggmagnify]).
 
@@
 
-To demonstrate how `sfislands` can be used to quickly compare the output of models, we show the output of three similar models.
+To demonstrate how \CRANpkg{sfislands} can be used to quickly compare the output of models, we show the output of three similar models.
 The first of these is a hierarchical model as described above, but with no ICAR component at the lowest level.
@@
 
-This shows how `sfislands` can quickly give an intuitive representation of how a set of similar models are operating.
+This shows how \CRANpkg{sfislands} can quickly give an intuitive representation of how a set of similar models are operating.
 As a measure of the relative strengths of these three structures, we could compare their AICs (see Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:aic-html)', '\\@ref(tab:aic-latex)'))`).
@@
   mutate(`relative AIC` = round(-AIC / min(AIC),3),
-         model = c("hierachical model, no ICAR",
-                   "hierachical model, ICAR based on st_bridges()",
-                   "hierachical model, ICAR based on sfdep::st_dist_band()")) |> 
+         model = c("hierarchical model, no ICAR",
+                   "hierarchical model, ICAR based on st_bridges()",
+                   "hierarchical model, ICAR based on sfdep::st_dist_band()")) |> 
   rownames_to_column(var="modcode") |> 
@@
 
-These examples have shown the varying scenarios in which `sfislands` can be useful.
+These examples have shown the varying scenarios in which \CRANpkg{sfislands} can be useful.
 It aims to contribute to spatial modelling by making an awkward area less awkward.
 Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included.
-It also provides helper functions to use these structures in spatial regression models built with \CRANpkg{mgcv} which streamline the human effort necessary to examine the estimates.
+Even when no islands are present, it provides a simple procedure for tailoring a neighbourhood structure to match a given hypothesis.
+It also provides helper functions to use these structures in spatial regression models, notably those built with \CRANpkg{mgcv}, which streamline the human effort necessary to examine the estimates.
 In future, compatibility with other modelling packages can be added to broaden the package's capabilities.

### 2024-03-28 article.R

@@ -5,7 +5,7 @@
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
 
 # install.packages("devtools")
-devtools::install_github("horankev/sfislands")
+# devtools::install_github("horankev/sfislands")
 
 # required libraries
 library(tidyverse)
@@ -330,12 +330,12 @@ ggplot() +
 ## ---- quake-occur, fig.width=12, fig.height=6, out.width="100%", fig.cap = "Earthquakes in Indonesia of magnitude > 5.5, 1985-2023. Categorised by magnitude as medium, large or extra-large. "----
 
 quakes_df_temp <- quakes_df |> 
-  mutate(magfact = case_when(magfact == "M" ~ "M: 5.5 - 6.1",
-                             magfact == "L" ~ "L: 6.1 - 7",
+  mutate(magfact = case_when(magfact == "M" ~ "M: 5.5 - 6",
+                             magfact == "L" ~ "L: 6.1 - 6.9",
                              magfact == "XL" ~ "XL: 7+")) |> 
   mutate(magfact = factor(magfact,
-                          levels = c("M: 5.5 - 6.1",
-                                     "L: 6.1 - 7",
+                          levels = c("M: 5.5 - 6",
+                                     "L: 6.1 - 6.9",
                                      "XL: 7+")))
 ggplot() + 
   geom_sf(data=nearby_countries_df, fill="gray50", linewidth=0.5, colour="black") +
@@ -593,6 +593,7 @@ plot_mrf <- mod_pois_mrf |>
 
 # in this case, there is only one plot in the list
 # so we call it by index
+# it is then supplemented with additional ggplot functions
 
 plot_mrf[[1]] +
   coord_sf(datum=NA) +
@@ -1202,9 +1203,9 @@ ggarrange(
 aic_tab <- AIC(liverpool_hmodel, liverpool_hsamodel, liverpool_hsamodel_dist_band) |> 
   data.frame() |> 
   mutate(`relative AIC` = round(-AIC / min(AIC),3),
-         model = c("hierachical model, no ICAR",
-                   "hierachical model, ICAR based on st_bridges()",
-                   "hierachical model, ICAR based on sfdep::st_dist_band()")) |> 
+         model = c("hierarchical model, no ICAR",
+                   "hierarchical model, ICAR based on st_bridges()",
[…]

### 2024-03-28 article.R
reading models from github

@@ -1,1224 +1,0 @@
-# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
-# Please edit article.Rmd to modify this file
-
-## ----setup, include=FALSE-----------------------------------------------------
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
-
-# install.packages("devtools")
-# devtools::install_github("horankev/sfislands")
-
-# required libraries
-library(tidyverse)
-library(kableExtra)
-library(sf)
-library(sfdep)
-library(sfislands)
-library(mgcv)
-library(here)
-library(ggsflabel)
-library(ggspatial)
-library(ggpubr)
-library(AER) # to test for overdispersion
-library(ggmagnify)
-
-# To scrape data
-library(rvest)
-library(httr)
-library(polite)
-library(janitor)
-
-# for geom_sf_label_repel consistency
-set.seed(123)
-
-# set consistent theme
-theme_set(
-  theme_bw() +
-    theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", linewidth=1.5), 
-          axis.title = element_blank(),
-          axis.ticks = element_blank())
-)
-
-
-
-## -----------------------------------------------------------------------------
-# load data
-
-# datasets are all available at https://github.com/horankev/quake_data 
-# or https://github.com/horankev/london_liverpool_data
-# alongside details on precisely how they were sourced etc
-
-# indonesia
-quakes_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/quakes_df.rds")))
-provinces_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/provinces_df.rds"))) |> 
-  rename(damaging_quakes_total = quake_mlxl_total,
-         damaging_quakes_density = quake_mlxl_density)
-counties_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/counties_df.rds")))
-faults_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/faults_df.rds")))
-nearby_countries_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/nearby_countries_df.rds")))
[…]

### 2024-03-28 article.Rmd
reading models from github

@@
 
-liverpool_hmodel <- readRDS(here("data","liverpool_model2.rds"))
+# read locally
+# liverpool_hmodel <- readRDS(here("data","liverpool_model2.rds"))
+
+# read from github
+liverpool_hmodel <- readRDS(gzcon(url("https://github.com/horankev/r_journal_sfislands/raw/main/data/liverpool_model2.rds")))
 
@@
 
-liverpool_hsamodel <- readRDS(here("data","liverpool_hsamodel.rds"))
+# read locally
+# liverpool_hsamodel <- readRDS(here("data","liverpool_hsamodel.rds"))
+
+# read from github
+liverpool_hsamodel <- readRDS(gzcon(url("https://github.com/horankev/r_journal_sfislands/raw/main/data/liverpool_hsamodel.rds")))
 
@@
 
-liverpool_hsamodel_dist_band <- readRDS(here("data","liverpool_hsamodel_dist_band.rds"))
+# read locally
+# liverpool_hsamodel_dist_band <- readRDS(here("data","liverpool_hsamodel_dist_band.rds"))
+
+# read from github
+liverpool_hsamodel_dist_band <- readRDS(gzcon(url("https://github.com/horankev/r_journal_sfislands/raw/main/data/liverpool_hsamodel_dist_band.rds")))
 

### 2024-03-28 article.tex
reading models from github

@@
 Fitting areal models which use a spatial weights matrix to represent relationships between geographical units can be a cumbersome task, particularly when these units are not well-behaved.
-The two chief aims of \texttt{sfislands} are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models.
+The two chief aims of \CRANpkg{sfislands} are to simplify the process of creating an appropriate neighbourhood matrix, and to quickly visualise the predictions of subsequent models.
 The package uses visual aids in the form of easily-generated maps to help this process.
-This paper demonstrates how \texttt{sfislands} could be useful to researchers.
+This paper demonstrates how \CRANpkg{sfislands} could be useful to researchers.
 It begins by describing the package's functions in the context of a proposed workflow.
@@
 These range from earthquakes in Indonesia, to river crossings in London, and hierarchical models of output areas in Liverpool.
-We aim to show how the \texttt{sfislands} package streamlines much of the human workflow involved in creating and examining such models.
+We aim to show how the \CRANpkg{sfislands} package streamlines much of the human workflow involved in creating and examining such models.
 }
@@
 
-For R users, the \CRANpkg{spdep} package (Bivand et al. 2005) has long been popular for the
-creation of these matrices. More recently, with the increasing use of \CRANpkg{sf} (Pebesma 2018)
-structures and their compatibility with the \CRANpkg{tidyverse} (Wickham et al. 2019), the \CRANpkg{sfdep} package (Parry and Locke 2024) has presented the same
+For R users, the \CRANpkg{spdep} package (Bivand, Hauke, and Kossowski 2013) has long been popular for the
+creation of these matrices.
+More recently, with the increasing use of \CRANpkg{sf} structures (Pebesma 2018) and their compatibility with the \CRANpkg{tidyverse} (Wickham et al.
+2019), the \CRANpkg{sfdep} package (Parry 2023) has presented the same
 functionality in addition to extra features in a tidy structure based on
@@
 
-The aim of \texttt{sfislands} is to deal with the situations described above in a
+The aim of \CRANpkg{sfislands} (Horan, Domijan, and Brunsdon 2024) is to deal with the situations described above in a
 convenient and open manner. It allows us to set up a structure, quickly
@@
 In this paper, we will look at three examples to show different use-cases for
-\texttt{sfislands}. The first example focuses on earthquakes in
+\CRANpkg{sfislands}. The first example focuses on earthquakes in
 Indonesia. It shows a scenario where all of the functions are used, from
@@
 
-The second example looks at London and how, despite the lack of islands,
+The second example looks at London and how, despite an absence of islands,
 the presence of a river means that some of the pre-functions of
-\texttt{sfislands} can be useful.
+\CRANpkg{sfislands} can be useful.
 
@@
 Instead, this example shows how the package can be used to quickly fit a set of reasonably complex multilevel models, with or without a spatially autocorrelated component at the lowest level, and how the results can be presented in tidy format and quickly visualised.
-This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the \texttt{sfislands} post-functions make it very straightforward to quickly compare the different predictions generated from different structures.
-
-\hypertarget{why-use-sfislands}{%
-\section{\texorpdfstring{Why use \texttt{sfislands}?}{Why use sfislands?}}\label{why-use-sfislands}}
+This example also gives an opportunity to show how, when different types of contiguity from, say, the \CRANpkg{spdep} package are used in an otherwise equivalent model, the \CRANpkg{sfislands} post-functions make it very straightforward to quickly compare the different predictions generated from different structures.
+
+\hypertarget{why-use}{%
+\section{\texorpdfstring{Why use \CRANpkg{sfislands}?}{Why use ?}}\label{why-use}}
 
@@
 \item
-  It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure is not compatible with their data.
-\texttt{sfislands} will include a contiguity for all units.
+  It addresses an issue commonly seen in online help forums where an inexperienced user wishes to get started with a model but fails at the first hurdle because their neighbourhood structure contains empty records.
+\CRANpkg{sfislands} will include a contiguity for all units.
 \item
@@
 \item
-  These maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply the desired styling and themes using \CRANpkg{ggplot2} syntax.
+  These maps are created using \CRANpkg{ggplot2} (Wickham 2016), which allows users to apply additional styling and themes using \CRANpkg{ggplot2} syntax.
 \item
@@
 \item
-  Connections which have been induced by a function from the package but are not contiguities can be accessed to ensure openness in the process.
+  Connections which have been induced by a function from the package but which are not based on geographical contiguity can be accessed to ensure openness in the process.
 \end{enumerate}
@@
 These structures can serve as the input to models in \CRANpkg{brms},
-\pkg{r-inla}, \pkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}. \CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list. Rather than having a separate neighbours
+\pkg{r-inla}, \CRANpkg{rstan} (Stan Development Team 2020) or \CRANpkg{mgcv}.
+\CRANpkg{brms} requires a matrix structure while \CRANpkg{mgcv} models use a list.
+Rather than having a separate neighbours
 object, it is included in the original \CRANpkg{sf} dataframe as a named list or matrix, in the spirit of the \CRANpkg{sfdep} package.
@@
 Having decided upon an appropriate neighbourhood structure, the next step is to use this in the context of a model.
-The use of such structures is particularly associated with CAR-type (conditional autoregressive) models.
+The use of such structures is particularly associated with CAR (conditional autoregressive) or ICAR-type (intrinsic conditional autoregressive) models (Besag 1974).
 These are often implemented in a Bayesian framework using
-\CRANpkg{brms}, \pkg{r-inla} or \pkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
+\CRANpkg{brms}, \pkg{r-inla} or \CRANpkg{rstan}. For example, the \CRANpkg{brms} ICAR structure requires the neighbourhood relationships to be in matrix form. The pre-functions will output the
 neighbourhood structure in the desired format for use in any of these
@@
 any other component of the model, it can be somewhat awkward to get the
-estimates from \CRANpkg{mgcv} model. \texttt{sfislands} has two post-functions to conveniently extract and visualise these.
+estimates from \CRANpkg{mgcv} models. \CRANpkg{sfislands} has two post-functions to conveniently extract and visualise these.
 
@@
 \footnote{In a similar way, \texttt{st\_augment()} can also be used to append the random effects from \CRANpkg{lme4} (Bates et al.
-2015) and \CRANpkg{nlme} (Pinheiro, Bates, and R Core Team 2023) models to an \texttt{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}.
+2015) and \CRANpkg{nlme} (Pinheiro, Bates, and R Core Team 2023) models to an \CRANpkg{sf} dataframe, which can then be easily mapped using \texttt{st\_quickmap\_preds()}.
 Compatability with models created using different packages can be introduced in the future.} The spatially varying predictions which \texttt{st\_augment()} extracts from an \CRANpkg{mgcv} model are
@@
 Consider the model structure described in the code below using \CRANpkg{mgcv} syntax.
-In this model \(y\) is the dependent variable while it is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a \emph{region} level, and a set of ICAR varying intercepts and slopes at a lower \emph{sub-region} level.
+In this model \(y\) is the dependent variable which is being estimated with a fixed intercept, a fixed slope for some covariate, a set of random intercepts and slopes for the covariate at a \emph{region} level, and a set of ICAR varying intercepts and slopes at a lower \emph{sub-region} level.
 
@@
 
-\hypertarget{example-1-indonesia}{%
-\section{Example 1: Indonesia}\label{example-1-indonesia}}
+\hypertarget{indonesia-example-1}{%
+\section{Indonesia (example 1)}\label{indonesia-example-1}}
 
@@
 
-Having described the type of model we wish to implement, we now show how \texttt{sfislands} can be used to streamline the process.
+Having described the type of model we wish to implement, we now show how \CRANpkg{sfislands} can be used to streamline the process.
 
@@
 we have set the value of k to 2. The resulting neighbourhood structure
-is shown in Figure \ref{fig:ind-contig1}.
+is shown in Figure \ref{fig:ind-contig1}. Note how is can be styled with a combination of internal arguments (size, colour, fill etc.) and additional \CRANpkg{ggplot2} layers.
 
@@
 However, the extraction of individual predictions for this component for each province from the \texttt{mgcv} model requires a number of steps.
-We now demonstrate how these are streamlined into a single function by the \texttt{sfislands} package.
+We now demonstrate how these are streamlined into a single function by the \CRANpkg{sfislands} package.
 
@@
 similar to that of the \CRANpkg{broom} package.
-For instance, we see from the code below that the original dataframe is now augmented with columns called ``mrf.smooth.province'' and ``se.mrf.smooth.province'' which show the predictions for the \(\gamma_i\) component and their standard errors.
+For instance, we see from the output of the following code chunk that the original dataframe is now augmented with columns called \texttt{mrf.smooth.province} and \texttt{se.mrf.smooth.province} which show the predictions for the \(\gamma_i\) component and their standard errors.
 Note that this is how we would expect them to be named, based on the previous discussion surrounding Table \ref{tab:staugtab-latex}.
-They are positioned immediately before the final \texttt{geometry} column of the \texttt{sf} dataframe.
+They are positioned immediately before the final \texttt{geometry} column of the \texttt{sf} dataframe, and after the neighbours list column, \texttt{nb}.
 
@@
 This allows us to produce the map in Figure
-\ref{fig:ind-mrf2}. As these coefficients are multiplicatively related
-to the earthquake incidence, values below 1 imply an earthquake
-incidence which is lower than expected.
-
-The provinces with the 3 most elevated incidences are labelled in red
-We can see that, controlling for the effects of proximity to
-faults, the province of Nusa Tenggara Barat has 8.7 times the expected
-incidence, or number of major earthquakes per square kilometre. The two
-lowest-scoring provinces, labelled in green, have essentially no incidence of
-earthquake epicentres within their boundaries, controlling for what their proximity to faults alone would
-suggest.
+\ref{fig:ind-mrf2}. As these coefficients are multiplicatively related to the earthquake incidence, values below 1 imply an earthquake incidence which is lower than expected.
+
+The provinces with the 3 most elevated incidences are labelled in red.
+We can see that, controlling for the effects of proximity to faults, the province of Nusa Tenggara Barat has 8.7 times the expected incidence, or number of major earthquakes per square kilometre.
+The two lowest-scoring provinces, labelled in green, have essentially no incidence of earthquake epicentres within their boundaries, controlling for what their proximity to faults alone would suggest.
 
@@
 However, in practice, at least in a first iteration, it might not be necessary to go through all of these steps.
-A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with \texttt{sfislands} using nothing more than three or four lines of code, such as the following:
+A rough and ready model, complete with spatially varying coefficients and visual output, can be generated with \CRANpkg{sfislands} using nothing more than three or four lines of code, such as the following:
 
@@
 
-\hypertarget{example-2-london}{%
-\section{Example 2: London}\label{example-2-london}}
-
-The next example looks only at using the \emph{pre-functions} of \texttt{sfislands}, but in a situation where the presence of actual \emph{islands} is not the problem we seek to deal with.
-Consider the wards and boroughs of London (sourced from the Greater London Authority's \href{https://data.london.gov.uk/}{London Datastore}).
-In Figure \ref{fig:lon-contig1} the \texttt{st\_bridges()}
+\hypertarget{london-example-2}{%
+\section{London (example 2)}\label{london-example-2}}
+
+The next example looks only at using the \emph{pre-functions} of \CRANpkg{sfislands}, but in a situation where the presence of actual \emph{islands} is not the problem we seek to deal with.
+Consider the wards and boroughs of London (sourced from the Greater London Authority's \href{https://data.london.gov.uk/}{London Datastore}) and available at
+\url{https://github.com/horankev/london_liverpool_data}. In Figure \ref{fig:lon-contig1} the \texttt{st\_bridges()}
 function is applied to them to construct a queen contiguity
@@
 If there is no means of crossing the river within a reasonable distance, somebody living on the banks of a river might be more likely to go about their life primarily on their side of the river, despite the short distance as the crow flies of facilities on the other side.
-This could be relevant in terms of, say, modelling of house prices where we might want to incorporate these issues into a neighbourhood structure.
-
-\texttt{sfislands} provides convenient functions for this sort of situation.
-Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames (Figure \ref{fig:lon-contig2}).
+This could be relevant in terms of, say, modelling of house prices where we might want to incorporate issues such as local amenities into a neighbourhood structure.
+
+\CRANpkg{sfislands} provides convenient functions for this sort of situation.
+Let us start by restricting our wards and boroughs of interest to just those which are on either side of the river Thames.
+Figure \ref{fig:lon-contig2} shows the resultant contiguities when the river is ignored.
 
@@
   accepts names as arguments which may be more convenient in some circumstances.} Having made these adjustments,
-\texttt{st\_quickmap\_nb()} now shows a connectivity structure (Figure \ref{fig:lon-contig4}) which reflects how we have hypothesised that influence should extend across the river in the presence or absence of crossings.
-
-This example shows that the pre-functions of \texttt{sfislands} have uses for situations which do not involve islands.
+\texttt{st\_quickmap\_nb()} now shows a connectivity structure (Figure \ref{fig:lon-contig4}) which reflects our hypothesis of how influence should extend across the river in the presence or absence of crossings.
+
+This example shows that the pre-functions of \CRANpkg{sfislands} have uses for situations which do not involve islands.
 They can be used to apply domain knowledge to easily design the most appropriate neighbourhood structure.
@@
 
-\hypertarget{example-3-liverpool}{%
-\section{Example 3: Liverpool}\label{example-3-liverpool}}
+\hypertarget{liverpool-example-3}{%
+\section{Liverpool (example 3)}\label{liverpool-example-3}}
 
 In this final example, there are no islands and no river issues.
-We saw in the Indonesia example how \texttt{sfislands} can extract estimates from the \texttt{mgcv} argument \texttt{bs="mrf"}.
+We saw in the Indonesia example how \CRANpkg{sfislands} can extract estimates from the \texttt{mgcv} argument \texttt{bs="mrf"}.
 It can also deal with random effects as defined by \texttt{bs="re"} in \texttt{mgcv} models.
-This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally related components.
+This means it is possible to create quite complex models featuring both a traditional multi-level structure and conditionally correlated components.
 Here, we will construct a hierarchical model in this way which also features an ICAR component at the lowest level of the tree.
@@
 \item
-  \texttt{st\_bridges()} from \texttt{sfislands}, which implements first-order queen contiguity in the same way as \texttt{sfdep:st\_contiguity()} for situations such as this where there are no islands, and
+  \texttt{st\_bridges()} from \CRANpkg{sfislands}, which implements first-order queen contiguity in the same way as \texttt{sfdep:st\_contiguity()} for situations such as this where there are no islands, and
 \item
@@
 
-There are no islands or discontiguities in this geometry of Liverpool so using \texttt{st\_bridges()} will give the same structure as would \texttt{sfdep::st\_contiguity()} (see Figure \ref{fig:liv-contigs}).
+There are no islands or discontiguities in this geometry of Liverpool so using \texttt{st\_bridges()} will give the same structure as would \texttt{sfdep::st\_contiguity()} (see Figure \ref{fig:liv-contigs}, with a magnified section for visibility purposes, created using \pkg{ggmagnify} (Hugh-Jones 2024)).
 
@@
 
-To demonstrate how \texttt{sfislands} can be used to quickly compare the output of models, we show the output of three similar models.
+To demonstrate how \CRANpkg{sfislands} can be used to quickly compare the output of models, we show the output of three similar models.
 The first of these is a hierarchical model as described above, but with no ICAR component at the lowest level.
@@
 
-This shows how \texttt{sfislands} can quickly give an intuitive representation of how a set of similar models are operating.
+This shows how \CRANpkg{sfislands} can quickly give an intuitive representation of how a set of similar models are operating.
 As a measure of the relative strengths of these three structures, we could compare their AICs (see Table \ref{tab:aic-latex}).
@@
 \hline
-hierachical model, ICAR based on st\_bridges() & -5126.948 & -1.000\\
-\hline
-hierachical model, ICAR based on sfdep::st\_dist\_band() & -5107.790 & -0.996\\
-\hline
-hierachical model, no ICAR & -5082.684 & -0.991\\
+hierarchical model, ICAR based on st\_bridges() & -5126.948 & -1.000\\
+\hline
+hierarchical model, ICAR based on sfdep::st\_dist\_band() & -5107.790 & -0.996\\
+\hline
+hierarchical model, no ICAR & -5082.684 & -0.991\\
 \hline
@@
 
-These examples have shown the varying scenarios in which \texttt{sfislands} can be useful.
+These examples have shown the varying scenarios in which \CRANpkg{sfislands} can be useful.
 It aims to contribute to spatial modelling by making an awkward area less awkward.
 Rather than having a default attitude of ignoring islands when building neighbourhood structures based on contiguity, it hopes to encourage at least an examination of whether or not it is appropriate for them to be included.
-It also provides helper functions to use these structures in spatial regression models built with \CRANpkg{mgcv} which streamline the human effort necessary to examine the estimates.
+Even when no islands are present, it provides a simple procedure for tailoring a neighbourhood structure to match a given hypothesis.
+It also provides helper functions to use these structures in spatial regression models, notably those built with \CRANpkg{mgcv}, which streamline the human effort necessary to examine the estimates.
 In future, compatibility with other modelling packages can be added to broaden the package's capabilities.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-besag}{}}%
+Besag, Julian.
+1974.
+{``Spatial Interaction and the Statistical Analysis of Lattice Systems.''} \emph{Journal of the Royal Statistical Society: Series B (Methodological)} 36 (2): 192--225.
+\url{https://doi.org/10.1111/j.2517-6161.1974.tb00999.x}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-spdep}{}}%
-Bivand, Roger, Andrew Bernat, Marilia Carvalho, Yongwan Chun, Carsten Dormann, Stéphane Dray, Rein Halbersma, et al.
-2005.
-{``The Spdep Package.''} \emph{Comprehensive R Archive Network, Version}, 05--83.
+Bivand, Roger, Jan Hauke, and Tomasz Kossowski.
+2013.
+{``Computing the Jacobian in Gaussian Spatial Autoregressive Models: An Illustrated Comparison of Available Methods.''} \emph{Geographical Analysis} 45 (2): 150--79.
+\url{http://www.jstatsoft.org/v63/i18/}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-sfislands}{}}%
+Horan, Kevin, Katarina Domijan, and Chris Brunsdon. 2024. \emph{{sfislands}: Streamlines the Process of Fitting Areal Spatial Models}. \url{https://horankev.github.io/sfislands/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ggmagnify}{}}%
+Hugh-Jones, David. 2024. \emph{Ggmagnify: Create a Magnified Inset of Part of a "Ggplot" Object}. \url{https://github.com/hughjonesd/ggmagnify}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-ggpubr}{}}%
-Kassambara, Alboukadel. 2023. \emph{Ggpubr: 'Ggplot2' Based Publication Ready Plots}. \url{https://rpkgs.datanovia.com/ggpubr/}.
+Kassambara, Alboukadel. 2023. \emph{{ggpubr}: {``{ggplot2}''} Based Publication Ready Plots}. \url{https://CRAN.R-project.org/package=ggpubr}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-sfdep}{}}%
-Parry, Josiah, and Dexter H Locke. 2024. \emph{Sfdep: Spatial Dependence for Simple Features}. \url{https://sfdep.josiahparry.com}.
+Parry, Josiah. 2023. \emph{{sfdep}: Spatial Dependence for Simple Features}. \url{https://CRAN.R-project.org/package=sfdep}.
 
@@
 2024.
-{``{List of crossings of the River Thames} --- {W}ikipedia{,} the Free Encyclopedia.''} \url{http://en.wikipedia.org/w/index.php?title=List\%20of\%20crossings\%20of\%20the\%20River\%20Thames\&oldid=1184426738}.
+{``{List of crossings of the River Thames} --- {W}ikipedia{,} the Free Encyclopedia.''} \url{http://en.wikipedia.org/w/index.php?title=List/\%20of/\%20crossings/\%20of/\%20the/\%20River/\%20Thames\&oldid=1184426738}.
 

### 2024-03-28 article.R

@@ -1,0 +1,1236 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit article.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
+
+# install.packages("devtools")
+# devtools::install_github("horankev/sfislands")
+
+# required libraries
+library(tidyverse)
+library(kableExtra)
+library(sf)
+library(sfdep)
+library(sfislands)
+library(mgcv)
+library(here)
+library(ggsflabel)
+library(ggspatial)
+library(ggpubr)
+library(AER) # to test for overdispersion
+library(ggmagnify)
+
+# To scrape data
+library(rvest)
+library(httr)
+library(polite)
+library(janitor)
+
+# for geom_sf_label_repel consistency
+set.seed(123)
+
+# set consistent theme
+theme_set(
+  theme_bw() +
+    theme(panel.background = element_rect(fill = "#F6F3E9", colour = "black", linewidth=1.5), 
+          axis.title = element_blank(),
+          axis.ticks = element_blank())
+)
+
+
+
+## -----------------------------------------------------------------------------
+# load data
+
+# datasets are all available at https://github.com/horankev/quake_data 
+# or https://github.com/horankev/london_liverpool_data
+# alongside details on precisely how they were sourced etc
+
+# indonesia
+quakes_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/quakes_df.rds")))
+provinces_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/provinces_df.rds"))) |> 
+  rename(damaging_quakes_total = quake_mlxl_total,
+         damaging_quakes_density = quake_mlxl_density)
+counties_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/counties_df.rds")))
+faults_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/faults_df.rds")))
+nearby_countries_df <- readRDS(gzcon(url("https://github.com/horankev/quake_data/raw/main/datasets/nearby_countries_df.rds")))
[…]

### 2024-04-16 README.Rmd
add link to arxiv preprint

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<img src="figures/logo.png" align="right" height="240"/>

<!-- badges: start -->
<!-- badges: end -->

This repository contains the code and data for the paper:

**sfislands: An R Package for Accommodating Islands and Disjoint Zones in Areal Spatial Modelling**.

A pre-print of the paper can be found on **arxiv** [here](https://arxiv.org/abs/2404.09863)
